
# Record formation

## Introduction
We have produced an Agent-Based Model, replicating the process of formation of the archaeological record in a specific site, considering different variables. In order to consider the possible relation between those variables, we have decided to express them as a Directed Acyclic Graph (DAG), following the ECS-DAG methodology proposed by Ferguson et al. (2020). In this case, however, and because our goal is to obtain an accurate initial template for the development of the model, we have not considered setting an exposure variable, but rather to propose different theoretic directional and causal relationships without relying on the conventional exposure-outcome assessment. Therefore, highlights in the graph below do not intend to differentiate among confounders, mediators, etc., but to emphasize the possible weight that each variable has for producing the outcome (produced waste by $m^2$).

At this point, we are considering as our outcome variable (produced waste by $m^2$) only 14C dating samples; this is elements that can be dated with 14C techniques. Therefore, we are considering mostly consumed food, which is directly derived from number of people present in the site, but also related to other factors, as specified in the DAG. However, even in this case, not all variables affect both the outcome and their assigned mediators. This is the case, for example, for the variable *Mobility pattern*. At this point, and although we are aware that mobility can have an impact and how food is consumed within a camp (Colette Berbesque et al., 2016) we have decided, for the sake of simplicity, not to consider the effect of this variable on the final 14C outcome, although its effects are indeed considered regarding the population size of the site.

In the same way, some variables are dependent on others, but this is not accounted for by the model, as it is expected that they are provided as expert information. Such is the case for most categorical variables. For example, *Economy type* affects *Site functionality*, in the sense that it is highly unlikely that a site with hunter-gatherer economy would function as a pen (since domestic animals did were not in use by Prehistoric hunter-gatherers), as much as it is unlikely that a site with a sedentary *Mobility pattern* would have a function of logistic camp. We consider that including these restrictions unnecessarily complicates the code and compromises its readability and performance and, therefore the user is prompted to define these variables, based on her/his own knowledge and expertise. In any case, once these variables have been defined, they do have an effect on the outcome, both as mediators and exposures, as it is expressed in the DAG, and, therefore, they must be carefully considered before initialising the model.


```{r, echo = FALSE}
library(DiagrammeR)

grViz("
digraph record_formation {

graph [layout = dot, ## Set title and layout
       label = 'Record formation process',
       labelloc = 't'
       fontname = Helvetica,
       fontsize = 30]

# node definitions with substituted label text
node [fontname = Helvetica, ## General node definition
      style = filled,
      color = MistyRose]
      
## Per node definition
a [label = '@@1',
   fillcolor = Orangered,
   fixedsize = shape,
   width = 3,
   height = 0.8,
   fontsize = 20]
b [label = '@@2']
c [label = '@@3',
   fillcolor = orange,
   fixedsize = shape,
   width = 4.5,
   height = 1,
   fontsize = 30]
d [label = '@@4',
   fillcolor = Orangered]
e [label = '@@5']
f [label = '@@6']
g [label = '@@7',
   fillcolor = Orangered,
   fixedsize = shape,
   width = 3,
   height = 0.8,
   fontsize = 20]
h [label = '@@8',
   fillcolor = Orangered]
i [label = '@@9']

## General edge definition
edge [color = grey,
       arrowhead = vee,
       arrowtail = vee]

## Specific edges
a -> {b c h}
b -> {c}
d -> {a b i} 
e -> {a b c h} 
f -> {a b e h} 
g -> {a b c e f h} 
h -> {c} 
i -> {c} 
#c -> {b} [dir = both, label = 'per sqm']

}

## Label definition
[1]: 'Population'
[2]: 'Area'
[3]: 'Waste produced/sqrm'
[4]: 'Site type'
[5]: 'Site functionality'
[6]: 'Mobility pattern'
[7]: 'Economy type'
[8]: 'Anthropogenic deposition rates'
[9]: 'Other deposition rates'
")


```

Finally, the variable *Area* deserves its own mention. When a site is in use, this variable (the amount of land occupied by the settlers) is conditioned by many factors, such as the ones portrayed in the DAG. However, this (the total area of the site) can be recovered by the archaeologists in the field. This has been coded as the variable *Maximum area*, and it is the value that divides the total waste by the occupied $m^2$. In addition, this variable can effectively act as a modulator for carrying capacity ($K$) within constrained sites (essentially caves and rockshelters). Therefore, the model considers the variable *Area* for internal purposes (but can be retrieved if necessary). This is, to condition $K$ on the size of the site. For open-air sites, this has no effect, as the group fission points are usually not related to the total occupied area, but to other factors, such as social structure, available resources, etc. Variation in the intensity of internal spatial occupation has not been considered due to its high uncertainty.

We have considered the following random variables:

* **Waste produced per $m^2$ (outcome variable)**: The actual archaeological record present in each $m^2$. At this stage we are considering only 14C dates and, because it may be in the interest of the research question to know if these dates have been retrieved on domestic vs wild samples, this has been constructed as 2-column data frame, where the first column is the variable that we are interested in, that is the total number of samples produced (a discrete variable with range $[0,\propto K]$. $K$ being the population's carrying capacity), and the second column is the number of how many of those are domestic. The number of domestics is spefified on the probability being domestic, either established by the researcher or based on the economy tye. *See specifications of function* `Wp_sq`
* **Economy type**: This variable defines whether the group inhabiting the site is hunter-gatherer or a farmer group. Thus, it is a dichotomic variable.
* **Population**: The size of the group at each time step. Discrete variable with range $[0,\sim K]$.
* **Site type**: This represents whether the site is a cave, a rockshelter or an open-air site. Therefore, this is a categoric variable with three possible values.
* **Anthropogenic deposition rates**: The anthropogenic deposition rates are crucial for the formation of the archaeological record. This is a continuous variable with a range $[0,\infty]$. *The upper range must be reassessed based on literature*.
* **Mobility pattern**: In this case, we have considered different mobility strategies, which can affect, directly or indirectly, the formation of the record. We have considered three possible strategies: "residential", "logistic" and "sedentary". Thus, a categorical variable with three possible outcomes.
* **Site functionality**: Five different possible site uses have been considered: "habitat", "logistic", "pen", "butcher" and "funeral", giving a categorical variable with five possible outcomes.
* **Maximum Area**: A single integer, the maximum possible size of the site in $m^2$. This is provided as expert knowledge.
* **Area**: The occupied area is entangled with many other variables, particularly depending on population size. It can modulate $K$ for constrained sites. For simplicity, it has been considered as a discrete variable with range $[0,MaxArea]$, measured in $m^2$. In any case, $Area \leq MaxArea$.
* **Other deposition rates**: The same in nature as anthropogenic deposition rates, but different in its causality and effect. It is also considered as a continuous variable with range $[0,\infty]$. *The upper range must be reassessed based on literature*.

## Set functions, initial conditions and caveats

These are the basic functions that we will be using
```{r}
set.seed(12345)

## Function 1. Simulation of death process
#' @param pd: The probability matrix for mortality by age
death <- function(x,pd){
  age <- as.numeric(x[1])
  return(rbinom(1,1,prob = pd[pd$Age==age,2]))
}

## Function 2. Simluation of carrying capacity limitation
## If the population exceeds the carrying capacity, it eliminates oversize
## with 0.8 probability per person exceeding.
#' @param x: Data frame or matrix. Population (number of people)
#' @param k: Integer. Carrying capacity

K_lim <- function(x,k){
  p <- nrow(x)
  if (p>k){
    o <- rbinom(p-k,1,0.8)
    o <- sum(o[o==1])
    o[o==0] <- 1 ## Avoids problem eliminating all the df if remove == 0
    x <- x[-sample(1:nrow(x),o,replace = FALSE),]
  }
  return(x)
}

## Function 3. Waste production per square meter
#' @param x: Vector. W_p (Waste produced)
#' @param E_t: Factor. E_t (Economy type)
#' @param dm_r: Probability value. If E_t = 'hg', dm_r = 0. 
#' If E_t = 'm', dm_r = 0.5. If E_t = 'f', dm_r = 1. Default is NA.
#' Because values for E_t = c('m','f') are unrealisitc, it is strongly advised
#' that the researcher establishes her/his own values based on the expert knowledge 
#' @param P_y: Vector. P_y (Number of people per year)
#' @param S_t: Factor. S_t (Site type)
#' @param Adr: Vector. Adr (Anthropogenic deposition rates)
#' @param M_p: Factor. M_p (Mobility pattern)
#' @param S_f: Factor. S_f (site functionality)
#' @param Ar: Integer. Ar (Area of occupation)
#' @param Odr: Vector. Odr (Other deposition rates)

Wp_sq <- function(x,E_t,dm_r = NA,P_y,Ar){
  
  ## Sets the probability of the dates being domestic
  if (is.na(dm_r)==TRUE){
    if (E_t == 'hg'){
      dm_r <- 0
    } else if (E_t == 'm') {
      dm_r <- 0.5
    } else {
      dm_r <- 1
    }
  } else {
    dm_r <- dm_r
  }
  
  try(if(dm_r > 1 | dm_r < 0) stop("Probabilities must be between 0 and 1"))
  
  ## So far, we consider only radiocarbon dates
  ## Assumed one goat per person/year. Needs reassessment
  w <- ((P_y*4000)/5)
  x <- w/Ar

  ## How many of these dates are domestic
  dom <- rbinom(1,size = x, prob = dm_r)
  x <- data.frame("Number of samples" = round(x), "Number of domestics" = dom)
  return(x)
}
```

We must set some initial values and carrying capacities. Site economy will be implemented later

```{r}
ts <- 1000 # Time span
Gs_init <- 20 # Initial population
Gs_init <- data.frame("Age" = sample(10:30,Gs_init,10:30, replace = TRUE),
                      "Sex" = sample(c("M","F"), Gs_init, prob = c(0.5,0.5), replace = TRUE))

## Structure of random variables
# W_p <- data.frame("Number of samples" = rep(0,ts),
#                  "Number of domestics" = rep(0,ts)) # Waste produced
# E_t <- factor(NA, c('hg','m','f')) # Economy type
# P_y <- rep(NA,ts) ## Number of people per year (Population)
# S_t <- factor(NA, c('c','rs','oa')) ## Site type
# Adr <- rep(NA,ts) ## Anthropogenic deposition rates
# M_p <- factor(NA, c('r','l','s')) ## Mobility pattern
# S_f <- factor(NA, c('h','l','p','b','f')) ## Site functionality
# Max_Ar <- c(NA) ## Maximum size of the site
# Ar <- c(NA) ## Occupied area of the site
# Odr <- rep(NA,ts) ## Other deposition rates

## Definition of random variables
W_p <- data.frame("Number of samples" = rep(0,ts),
                  "Number of domestics" = rep(0,ts)) # Waste produced
E_t <- factor('hg', c('hg','m','f')) # Hunter-gatherers
P_y <- rep(NA,ts) ## Number of people per year (Population)
S_t <- factor('c', c('c','rs','oa')) ## Cave
Adr <- rep(NA,ts) ## Anthropogenic deposition rates
M_p <- factor('l', c('r','l','s')) ## Logistic mobility
S_f <- factor('h', c('h','l','p','b','f')) ## Habitat
Max_Ar <- 100
Ar <- rep(NA,ts)
Odr <- rep(NA,ts) ## Other deposition rates

## K is determined by the site type and economy

## K <- 4-6 For rock-shelters
## K <- 15 small cave
## K <- 40 medium cave
K <- 70 ## large cave
## K <- 100 open-air (meso) depending on resources
## K <- 200 open-air (neo)

## Probability matrix for mortality
prob_d <- data.frame("Age" = c(0:99),
                     "P_d" = c(rep(0.13,2),
                               rep(0.08,8),
                               rep(0.01,25),
                               rep(0.1,15),
                               rep(0.15,10),
                               rep(0.25,5),
                               rep(0.99,35)))

```

Camp size depends on camp type (which can be rockshelter, cave or open-air)
If camp type = rockshelter or cave, a maximum size must be specified

We can preset values if the size of the camp is unknown
  - For rockshelters: small (4 sqrm), medium (15 sqrm), large (30 sqrm)
  - For caves: small (100 sqrm), medium (750 sqrm), large (1500 sqrm). More than this (or even within this). For more than this, subdivisions of the space are required.
  
If camp type = open-air, size increases as population size increases with a density 20 sqrm/person. Size constrains can be set also here.

For waste production, this is computed as total waste production / sqrm

## Birth-death process for a single site
```{r}
Pop <- Gs_init ## Population (Age and Sex)
camp_size <- rep(NA,ts) ## Size of camp. Depends on people and type of camp

for (i in 1:ts){
  #Are born
  ## Female fertile population
  W <- Pop[Pop$Sex=="F",]
  W_fert <- W[W$Age>10 & W$Age<30,]
  
  ## Male fertile population
  M <- Pop[Pop$Sex=="M",]
  M_fert <- M[M$Age>15 & M$Age<40,]

  ## Probability of having descendance per woman
  # Penalisation in case there are too few men
  pen <- round(nrow(M_fert)*2/nrow(W_fert),2) ## Assumes one man can have two women
  pen[pen>1] <- 1
  
  ## 0.3 is drawn from ethnographic literature (each !Kung woman waits 3 years to have another son)
  P_o <- 0.3*pen

  Offspring <- sum(rbinom(nrow(W_fert),1,P_o))
  New_pop <- data.frame("Age" = rep(0,Offspring),
                        "Sex" = sample(c("M","F"),Offspring,prob=c(0.5,0.5),replace = TRUE))
  Pop <- rbind(Pop,New_pop)
  
  ## Die
  vec_d <- apply(Pop,1,death,prob_d)
  Pop <- Pop[vec_d==0,]
  
  ## Apply carrying capacty restrictions
  Pop <- K_lim(Pop, K)
  
  ## One year passes
  Pop$Age <- Pop$Age+1
  P_y[i] <- nrow(Pop)
  
  ## Waste production
  W_p[i,] <- Wp_sq(W_p[i,1], E_t = E_t, P_y = P_y[i], Ar = Max_Ar)
}

plot(P_y, type = "l", col = "red",
     xlab = "Year", ylab = "Population",
     main = "Pop")
```

There are still many things to include, basically related to deposition rates, economy-related, waste generation... and I need to do a better text for this chapter. I'm on it.

