---
output:
  html_document: default
  pdf_document: default
---

# Landscape & Resources {#env}


To simulate the interaction between two groups, we need an environment in which they can interact in. This chapter walks you through how to generate a landscape and an environment for your simulated groups to explore.
An important part of the simulation is to generate a landscape and an environment. Environmental variable are often autocorrelated. For example, The temperature at a given time is never _extremly_  different to the temperature slightly before and slightly after this time. Same is true with the slope, or the elevation of a place, 

## Autocorrelated noise generation


An interesting way to generate such auto-correlated variable is to use [coloured noise](https://en.wikipedia.org/wiki/Colors_of_noise). This use specific distribution to define the relation between frequencies found in your signal with the power of these frequencies. In [pink noise](https://en.wikipedia.org/wiki/Pink_noise) (or $\frac{1}{f}$-noise) for example, the power of each frequencis will be invertly proportional to the frequencies (_ie_ you will have, low frequencies with high power, high frequencies with low ones).  We defined an `environment` function, which in turns uses `TK95` to generate a power law noise.

To generate a single auto-correlated variable (like temperature), who's spectrum decomposition ( power wrt frequency) has a slope of -3 and a standard deviation of 1.5:

```{r generateFakeClimate}
tstep <- 1000
faketemp <- environment(tstep, omega=3, delta=1.5) + 1.5
plot(faketemp, type="l", ylab="temperature change", xlab="time",lwd=3)   
```

You can inspect the code of `environment` and `TK95` functions here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
a = knitr::spin('../../R/climate.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/climate.Rmd', '.')
res = knitr::knit_child(input='climate.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('climate.Rmd')
```
</details>


This generates autocorrelated time series that can be used to define temporal change in climate (rain, temperature, etc....). But we haven't used such changes for the Archaeoriddle challenge. Coloured noise remains a very straightforward way to simulated factor and could be easily added in the simulations.

## Slope and Elevation

An important aspect of the simulation is a way to generate elevation and slope. To generate 2D autocorrelation we used a slightly methods known as [perlin noise](https://en.wikipedia.org/wiki/Perlin_noise) ; using the implementation found on [this stackoverflow.com post](https://stackoverflow.com/questions/15387328/realistic-simulated-elevation-data-in-r-perlin-noise).

<br>

And you can see the code here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
a = knitr::spin('../../R/perlin_noise.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/perlin_noise.Rmd', '.')
res = knitr::knit_child(input='perlin_noise.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('perlin_noise.Rmd')
```
</details>


Various parameters for the Perlin noise will result in either a higher or lower level of definition for our future DEM raster. Our aim is to find a balance between having the environment sufficiently detailed across a relatively wide area, while also ensuring it remains manageable in terms of memory usage.

```{r perlin_noise}
require(terra)
require(sf)

a <- 0.6
k <- 8
m <- perlin_noise(2, 2, 2^k, 2^k)
for ( i in 2:k )
  m <- m + a^i * perlin_noise(2^i, 2^i, 2^k, 2^k)
```

If we change k (that dictates the total resolution of the DEM), the size of the raster grow explonentially and would be quickly hard to manage.


```{r explorePerlin}

allsizes <- sapply(1:7,function(k)
{
    print(k)
    mt <- perlin_noise(2, 2, 2^k, 2^k)
    for ( i in 2:k ) mt <- mt + 0.6^i * perlin_noise(2^i, 2^i, 2^k, 2^k)
    object.size(mt)  
})
plot(2^(1:7),allsizes/1000,type="b",pch=20,col="red",cex=2,ylab="size of raster (kb)", xlab="resolution")

```

To enhance realism, slight adjustments were made to the extremes in order to create seas and mountains that more closely resemble those found in an _real_ DEM.


```{r padjuperlin}
height <- (m[,])*20
#height[height<50]=0
height <- height^3
height <- height+abs(mean(height))
height[height<min(height)*.25] <- min(height)*.25
```


We can then visualise the result as a 2D raster. We setup the total extent of the raster to define exactly where it is on our GIS using a latlon projection.

We also define the mean high as the water level, all the thing below are then considered under water.

```{r 2dVisu}
height.ras <- rast(height, extent=c(-3.5, 0.5, -0.5, 3.5), crs="+proj=latlon")
maxh <- max(height.ras[], na.rm=T)
plot(height.ras, col=col_ramp(20), legend=F, reset=F)
height.wat <- height.ras
height.wat[height.wat>mean(height.wat[])] <- NA
plot(height.wat, col="lightblue", add=T, legend=F)

```

This has been put in a function  plotMap

```{r}
plotMap(height.ras,height.wat)
```


The actual raster used for the original _archaeoriddle_ is available in `data_original/east_narnia4x.tif` and looks like

```{r originalArchaeoriddle}
original.ras <- rast("data_original/east_narnia4x.tif")
original.wat <- original.ras
original.wat[original.wat>mean(original.wat[])] <- NA
plotMap(original.ras,original.wat)
```

If you really want to feel like being in a video game, you can play with rayshader, but getting the right parameters to have a nice rendering is tricky, slow, and it's not unlikely you'll loose a lot of time `¯\_(ツ)_/¯`

```{r threedim, eval=F}
require(rayshader)
height_map <- raster_to_matrix(raster::raster(height)) 

hm_shade <- sphere_shade(height_map, texture = "imhof1", zscale=5) 
hm_ray <- add_shadow(hm_shade, ray_shade(height_map, zscale = 19), 0.8) 
hm_ambiant <- add_shadow(hm_ray, ambient_shade(height_map), 0,5) 
hm_lamb <- add_shadow(hm_ambiant, lamb_shade(height_map), 0) 
plot_3d(
  hm_lamb,height_map, zscale = 10, fov = 0, theta = 20, zoom = .5, phi = 45,
  windowsize = c(1000, 800),water = TRUE, waterdepth = 0,
  wateralpha = 1, watercolor = "lightblue", waterlinecolor = "lightblue",
  waterlinealpha = .7, baseshape="hex")

#HQ render, very slow
#render_highquality( samples=256, clear = TRUE,light=TRUE)
```


## Environmental resources

We will generate zones of higher resources/fitness that will attract settlements and support higher carrying capacity _ie_ population size.
The resources will be located at given hotspots with an area around in which they decay
with a logistic function. We provide a very basic function `logisticdecay` which allow to define multiple source with different decay array around them.


```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/logistic_decay.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/logistic_decay.Rmd', '.')
```

Here's the code for the generation of resources under logistic decay 
<details><summary>Show code</summary>
**Code**
```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='logistic_decay.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('logistic_decay.Rmd')
```
</details>

<br>

For the original _archaeoriddle_, these hotspots and their radius were manually selected given the topography of Rabbit Hole. These manual setups are given below:

```{r setresource}
##resource geolocalisation set manually
goodresources <- vect(
  cbind(
    x=c(-0.2300711, -3.1455282, -0.5086485, -1.9639755,
        -0.4077843, 0.019688, -3.116710),
    y=c(3.6423000, -0.2551019, -0.7440748, 1.1303214,
        1.0248567, 0.2194895, 2.0267718)
  )
)
#spread of resources
areas <- 4 * c(100, 2000, 200, 300, 100, 200, 400)
#speed of ressource decay:
ks <- c(0.0002, 0.0001000, 0.0001600, 0.0001800, 0.00040, .0002, 0.0002)/4


crs(goodresources) <- crs(original.ras)
```

The function `logisticdecay` generates individual rasters for all the identified hotspots, which are then merged to create an aggregate raster referred to as 'resources'.

When considering the selection of hotspots, one may choose to retain the one utilized for the _archaeoriddle_ but to adjust these resources center to your own topography ;  one can use `locator()` to find the coordinates of the hotspot by clicking on the map. An example of that is given in [this  section of the chapter](#manual-selection)

These resource span along all the map but we want to restrict them to place where peoplewill be able to settle
Given this new landscape we can now generate sites in specific locations. We first need to define _where_ the sites _can_ appear (not in water nor too high in the mountains) ; so we first create a raster 'viable' which will give a binary maps where group of human can leave or not.

```{r}
height.groups <- height.ras # taking back the raster created before
height.groups[height.groups<mean(height.groups[])] <- NA #replace values that are below mean level (under water)
height.groups[height.groups<(maxh*.7)] <- 1 # values below 70% of the maximum height are set as 1
height.groups[height.groups>(maxh*.7)] <- 200 # value above at 200 (high mountains)
height.groups[is.na(height.groups)] <- -1 #  NA values (below waters) are set to -1
height.poly <- as.polygons(height.groups) # convert height.groups as polygons that will be of three type: -1 => under water, 1, viable, and 200, high mountaines
viable <- makeValid(height.poly[2,]) # select polygon with '1' and make it a valid polygon as it may have some loop and problems
```

We now have also have a polygon `viable` representing the space where sites can be create.
This space is representing by the green area on the map below.

```{r,viablearea}
plot(viable,col=adjustcolor("chartreuse",.8))
```

We then loop over all center manually defined previously and apply the logicitc decay within the `viable` raster.

```{r groupresources}

allres <- lapply(
  seq_along(goodresources),
  function(i){
    logisticdecay(goodresources[i], mask(height.ras, viable),
                  x=areas[i], k=ks[i]
    )
  }
)

allna <- sapply(allres, function(i)any(is.na(values(i))))
allres <- allres[!allna]
```

And when the add all the ressource raster, which gives us a final `ress` raster as below:

```{r showresrouce}
ress <- allres[[1]]
for(i in 2:length(allres))
    ress <- ress + allres[[i]]
ress <- mask(ress,viable)
plot(ress)
```





## Manually Select Resources {#manual-selection}

Below some useful code to manually select the resource origins on the map (by clicking on the screen) and explore different decays around them

```{r, eval=F}
plotMap(height.ras, height.wat, paste0("year ", 0))
new_site <- vect(as.data.frame(locator()), geom=c("x", "y"))  # select point on the screen
crs(new_site) <- crs(height.ras)
d2 <- logisticdecay(new_site, height.ras, x=20000) # generate a logistic decay around the point

plot(d2)
w <- (0.7*d2 + 0.3*ress) / (0.7*minmax(d2)[2] + 0.3*minmax(ress)[2])
plot(w)
plot(mask(w*logisticdecay(new_site, height.ras, k=0.00002, x=1), territory))

multisamp <- sapply(
  1:1000,
  function(i){
    spatSample(x=w, size=1, method="weights", xy=T)[1:2]
  }
)

```
