---
output:
  html_document: default
  pdf_document: default
  always_allow_html: true
---

# Record loss

## Theoretical model
### General purpose
Our main interest is to show and test the effect that different post-depositional process have on the archaeological record. The most obvious of these relates to the direct loss of the record, while other effect can imply other types of loss of records, but also vertical and horizontal movements which can distort our comprehension of an archaeological site. Although the horizontal disposition of archaeological waste is important to understand different aspects, such as the activities deployed in a site, the function of that site or it can even bring information on demographic processes, it will not be taken into account here. The reason for this is that modelling horizontal post-depositional movements (1) requires specific types of information which are frequently not available for field researchers, (2) adds complexity layers, the payoff of which, given its cost in terms of computation and comprehension of the model, may be insufficient and (3) given that we are focusing on the production of 14C dates, we are more interested in vertical movements, which can obscure the construction of consistent stratigraphies. 

However, we understand that this is a key aspect of the record formation. For example, it is frequent that hydric, gullying or sloping processes concentrate the archaeological waste in specific parts of the site. This would bias our perception of record concentration both horizontally, but also vertically. This *record loss* model assumes and normalises this when $W$ is retrieved from the previous *record formation* model. However, this needs to be taken into account both if the user wants to introduce the values for $W$, and for the analysis with real archaeological data. There is not a single solution to control for this, as it largely depends on the excavated area and whether this is representative of the complete site or not. For example, many surveys spread across a site could actually be representative of that site's record, whereas a single small survey in that same site is likely prone to present some specific bias due to post-depostional conditions. In this case, and in order to propose a correct variable, the user needs to make the proper calculations according to the site under study, but always having in mind that the model will work over the variable $W$ which is *waste produced per* $m^2$.

### Model layers
We have considered the following two main ways in which the archaeological record can be affected by post-depositional processes. The first one affects how record is lost as a continuum, both by time-dependent taphonomic loss and by anthropic action. The second one, however, although it can also be applied on a *per year* basis, rather points to the way how archaeologists perceive their site, stratigraphy and layers, and how those affect each other.

#### Time dependent record loss
The are two main components for this part of the model. As shown in the graph below, the first one is *time-dependent taphonomic loss* ($tdtl$) and the second one is *anthropic action* ($aa$). The first one is the only component that cannot be eliminated from this model. These two variables can be applied together (advised), separately or none at all. 

```{r, echo = FALSE}
library(DiagrammeR)

grViz("
digraph record_formation {

graph [layout = dot, ## Set title and layout
       label = 'Record loss process',
       labelloc = 't'
       fontname = Helvetica,
       fontsize = 30]

# node definitions with substituted label text
node [fontname = Helvetica, ## General node definition
      style = filled,
      color = MistyRose]
      
## Per node definition
a [label = '@@1',
   fillcolor = Orangered]
b [label = '@@2',
   fillcolor = Orangered]
c [label = '@@3',
   fillcolor = orange,
   fixedsize = shape,
   width = 6.5,
   height = 1,
   fontsize = 30]

## General edge definition
edge [color = grey,
       arrowhead = vee,
       arrowtail = vee]

## Specific edges
a -> {c}
b -> {c}
}

## Label definition
[1]: 'Time-dependent taphonomic loss'
[2]: 'Anthropic action'
[3]: 'Remaining archaeological waste'

")
```

These variables are modelled as follows:

* **Time-dependent taphonomic loss** ($tdtf$): This is a continuous variable. It represents the amount of loss record for each time $t$. It follows Surovell's (2007, 2009) works assessing the loss of archaeological record through time. Because this measure is intended as comprehensive, there is now need to condition the time-dependent record loss in different sub-processes which would add unnecessary complexity and uncertainty. Thus, this can be conceived as an overall record loss, and it is modelled through an adaptation of Surovell's proposal 

 $$n_t=\frac{a(b+t)^c}{a(b+t_1)^c} \times W_t$$
 
Where $a$, $b$ and $c$ are the constants proposed by Surovell, based on taphonomic regression, $t$ is the time step and $W$ is the original waste.

* **Anthropic action** ($aa$): However, Surovell's proposal is thought to measure the loss of record from the moment of abandonment of the site, and it does not take into account the possible effect of human continued daily activities (such as trampling, cleaning, etc.) for the case of sites with frequent reoccupation or long occupational time spans. In this regard, we apply the above method for each year during the occupational time span, to which we add the possible effect of $aa$, which is modelled as 

$$W_t = W_t - (W_t \times h_t \times s_t)$$
Where $h$ is the strength of anthropic effect, which can be high ($0.005$), medium ($0.003$) or low ($0.001$) and $s$ accounts for sedimentation rates, considering those the time that it takes for the record to be buried, point from which this variable has no effect. $s$ can be fast, medium or slow, being $10$, $30$ and $80$ years respectively.

#### Post-depositional mixture
This model layer accounts for vertical movement within the archaeological record. In the previous layer, the loss of record has already been accounted for according to Surovell's proposition and an adition of the possible effect of everyday activity in subsequent years of occupation of the site. However, for the remaining record, there is still a wide range of processes that can affect vertical deposition. In turn, vertical post-depositional movement can be one of the main causes confounding archaeological interpretation, as it can result in layers representing chronological palimpsests which, in extreme cases, can be almost impossible to disentangle. 

Essentially, this layer of the model reproduces vertical displacement, resulting in mixed chronologies, according to the user's specification. It can be used on the full chronological distribution of waste $W$. However, for computational efficiency, it is probably advisable to subdivide and group that chronological span in smaller chunks, representing the layers of the site. We have considered the following effects:

```{r, echo = FALSE}
library(DiagrammeR)

grViz("
digraph record_formation {

graph [layout = dot, ## Set title and layout
       label = 'Record formation process',
       labelloc = 't'
       fontname = Helvetica,
       fontsize = 30]

# node definitions with substituted label text
node [fontname = Helvetica, ## General node definition
      style = filled,
      color = MistyRose]
      
## Per node definition
a [label = '@@1',
   fillcolor = orange,
   fixedsize = shape,
   width = 4.5,
   height = 1,
   fontsize = 30]
b [label = '@@2',
   fillcolor = grey]
c [label = '@@3',
   fillcolor = Orangered]
d [label = '@@4',
   fillcolor = grey]
e [label = '@@5',
   fillcolor = grey]
f [label = '@@6',
   fillcolor = Orangered]
g [label = '@@7']
h [label = '@@8']
i [label = '@@9',
   fillcolor = Orangered]
j [label = '@@10',
   fillcolor = Orangered]
l [label = '@@11',
   fillcolor = grey]

## General edge definition
edge [color = grey,
       arrowhead = vee,
       arrowtail = vee]

## Specific edges
b -> {a}
c -> {a}
d -> {a}
e -> {a}
f -> {a}
g -> {b, c}
h -> {b, c, d, e, f, i, j}
i -> {a}
j -> {a}
l -> {a}

}

## Label definition
[1]: 'W'
[2]: 'Aoelic erosion'
[3]: 'Hydric activity'
[4]: 'Pedogeny'
[5]: 'Biogeny'
[6]: 'Posterior Anthropic action'
[7]: 'Site type'
[8]: 'Soil type'
[9]: 'Animal action'
[10]: 'General pedoturbation'
[11]: 'Intervention' #(as in time-series intervention). A disruptive phenomenon, like a rockfall


")
```

Where light pink nodes are indirect effects on mediators, orange nodes are the effects considered and grey nodes are effects that have been acknowledge, but not incorporated for different reasons. The following effects, or variables, have been considered:

* **Site type** (*St*): This is a discrete variable with three possible categories (*cave*, *rockshelter* and *open-air*), and it affects mainly hydric processes as, in terms of vertical movements, these can affect caves particularly.
* **Soil type** (*Sot*): This is a discrete variable with four possible categories (*gravel*, *sand*, *silt* and *clay*). Each category is defined by the thickness of its grain (from the > 2mm of gravels to the almost invisible clay) and this thickness can condition the effect of vertical movements in general.
* **Hydric activity** (*Ha*): Categorical variable with three values, indicating the intensity of the process.
* **General pedoturbation** (*Gp*): In order to simplify the model, this variable includes general pedoturbation processes. It is a categorical variable with three values, indicating the intensity of the process.
* **Posterior anthropic action** (*Paa*): Unlike the previously seen variable $Anthropic action$ or $aa$, this variable does not account for daily occupational activities, such as trampling or cleaning. This variable accounts for aggressive human action against the archaeological sequence, which can affect one or many layers of the site. Some examples for this variables could be Neolithic pits or extraction of manure common in modern and contemporary caves used as pens. Of course, this also includes any type of contemporary mechanical and/or agricultural action, such as the use of diggers or plough. This is a categorical variable with three values, indicating the intensity of the process.
* **Animal action** (*Ana*): Animals frequently excavate dens both in cave and open-air settlements, and these burrows can have an effect on sedimentary vertical organisation. This is a categorical variable with three values, indicating the intensity of the process.

The following variables have been acknowledged. However, in an attempt to keep the model manageable and meaningful, we have decided not to include them for the following specific reasons.
* **Aeolic erosion**: Although aeolic erosion can have a certain effect mainly on horizontal post-depositional effects, its effect on vertical effects can usually be neglected, except for extreme cases.
* **Pedogeny**: Pedogeny is the process of soil formation. There are many factors involved in the process, including chemical and organic components. However, the process itself does not necessarily affect the record vertically. For simplication, the possible effects of this process have been considered in the general variable $Gp$.
* **Biogeny**: Biogeny contributes to pedogenic effect by specifically creating soils. Nevertheless, because of the same reasons above, it has not included as a specific variable, but it has been encapsulated within the general $Gp$.
* **Intervention**: Intervention is understood as extreme events that can change the physiognomy of a site. These can include extreme floods, earthquakes provoking rock collapses and other phenomena of the sort. These phenomena have very different effects on the record each, which makes it virtually impossible to propose a one-way-fits-all general model, applicable across many sites. Therefore, it is advised that the user considers these phenomenon individually. For example, for a rock collapse, which frequently seals the archaeological record, developing two models, prior and posterior to that collapse.

## Implementation






