---
output:
  html_document: default
  pdf_document: default
---
# Landscape, climate & Environment

```{r, echo=FALSE, message=FALSE}
library(printr)
library(knitr)
library(archaeoriddle)

```

## Temperature and climate

An important part of the simulation is to generate a landscape and an environment.
Doing so often rely on
Let's Create a fake environment. We use the `environment` function, which in turns
uses `TK95` to generate a power law noise.

<br>

```{r, echo=FALSE}
?TK95
```
<br>

```{r, echo=FALSE}
?environment
```
<br>

To generate a fake climate with various properties:

```{r generateFakeClimate}
tstep <- 1000
faketemp <- environment(tstep, omega=3, delta=1.5) + 1.5
plot(faketemp, type="l", ylab="temperature change", xlab="time")   
```

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/climate.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/climate.Rmd', 'spin_functions/')
```

You can inspect the code of `environment` and `TK95` functions here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='spin_functions/climate.Rmd', quiet=TRUE)
cat(res, sep = '\n')
```
</details>

## Slope and Elevation

This autocorelate time series can be used to adjust temporal change in climate (rain, temperature, etc....). But we also need a way to generate elevation and slope. We can use similar methods to generate 2D autocorellation. We use here an implementation of what is known as [perlin noise](https://en.wikipedia.org/wiki/Perlin_noise) found on [this stackoverflow.com post](https://stackoverflow.com/questions/15387328/realistic-simulated-elevation-data-in-r-perlin-noise). This is the usage of the function:

<br>

```{r perlinnoise, echo=FALSE}
?perlin_noise
```
<br>

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/perlin_noise.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/perlin_noise.Rmd', 'spin_functions/')
```

And you can see the code here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='spin_functions/perlin_noise.Rmd', quiet=TRUE)
cat(res, sep = '\n')
```
</details>


Different parameters for the perlin noise will give us a higher a lower definition for our future DEM raster. Here we choose a trade off between an environment detailed enough for on a relatively wide area but small enough to be handled easily in memory.

```{r perlin_noise}
require(terra)
require(sf)

a <- 0.6
k <- 8
m <- perlin_noise(2, 2, 2^k, 2^k)
for ( i in 2:k )
  m <- m + a^i * perlin_noise(2^i, 2^i, 2^k, 2^k)
plot(rast(m))
object.size(m)
```

To make it slightly more realistic we tweaked a bit the extreme, to create seas and mountains a bit closer to what a _real_ DEM would look like.


```{r padjuperlin}
height <- (m[,])*20
#height[height<50]=0
height <- height^3
height <- height+abs(mean(height))
height[height<min(height)*.25] <- min(height)*.25
```


Visual the map as a simple 2D raster


```{r 2dVisu}

# height.ras <- rast(height, extent=c(-3.5, 0.5, -0.5, 3.5), crs="+proj=latlon")
height.ras <- rast("data_original/east_narnia4x.tif")
maxh <- max(height.ras[], na.rm=T)
col_ramp <- colorRampPalette(c("#54843f", "grey", "white"))

plot(height.ras^1.9, col=col_ramp(20), legend=F, reset=F)
height.wat <- height.ras
height.wat[height.wat>mean(height.wat[])] <- NA
plot(height.wat, col="lightblue", add=T, legend=F)

```

If you really want to feel like being in a video game, you can play with rayshader, but its slow, tricky and you will probably loose a lot of time `¯\_(ツ)_/¯`

```{r threedim, eval=F}
require(rayshader)
height_map <- raster_to_matrix(raster::raster(height)) 

hm_shade <- sphere_shade(height_map, texture = "imhof1", zscale=5) 
hm_ray <- add_shadow(hm_shade, ray_shade(height_map, zscale = 19), 0.8) 
hm_ambiant <- add_shadow(hm_ray, ambient_shade(height_map), 0,5) 
hm_lamb <- add_shadow(hm_ambiant, lamb_shade(height_map), 0) 
plot_3d(
  hm_lamb,height_map, zscale = 10, fov = 0, theta = 20, zoom = .5, phi = 45,
  windowsize = c(1000, 800),water = TRUE, waterdepth = 0,
  wateralpha = 1, watercolor = "lightblue", waterlinecolor = "lightblue",
  waterlinealpha = .7, baseshape="hex")

#HQ render, very slow
#render_highquality( samples=256, clear = TRUE,light=TRUE)

```


## Settlements position

Given this new landscape we can now generate sites in specific location (not on mountains, not in water) with exponentially distributed size and make a network out of it.  



```{r splitArea}
n <- 20
size <- rexp(n)

height.groups <- height.ras
height.groups[height.groups<mean(height.groups[])] <- NA
height.groups[height.groups<(maxh*.7)] <- 1
height.groups[height.groups>(maxh*.7)] <- 200
height.groups[is.na(height.groups)] <- -1
height.poly <- as.polygons(height.groups)
viable <- makeValid(height.poly[2,])

sites <- spatSample(viable, n)

plot(height.ras^1.9, col=col_ramp(50), legend=F, reset=F)
plot(height.wat, col="lightblue", add=T, legend=F)
plot(sites, cex=2, pch=21, add=T, bg=rainbow(length(sites), alpha=.6))

# get only the non submerged actual land and plot it
above_level <- height.ras > mean(height.ras[])
coastline <- st_as_sf(as.polygons(above_level))[2,]
plot(coastline, col=NA, bgc=adjustcolor("cyan", 0.1), add=T)

```

In order to produce file to be given to people we need a couple of line of code that should be run online when a satisfying environment has been found:

```{r, eval=F}
# st_write(dsn="coastline2.shp",coastline)
# writeRaster(filename="data_original/east_narnia4x.tif",height.ras,overwrite=T)
```

## Network, Site Size And Climate -- DEPRECATED

The following steps where designed but never used.

Use sites to create a network

```{r sitesnetwork, eval=F}

library(igraph)
crs(sites) <- "+proj=lonlat"
sites <- sites[sample(nrow(sites), 20), ]
ig <- graph_from_adjacency_matrix(
  as.matrix(distance(sites)), weighted=T, mode="undirected")
ig <- set_graph_attr(ig, "layout", crds(sites))

V(ig)$label <- ""
V(ig)$size <- (size+1)*5
V(ig)$frame.color <- "black"
V(ig)$frame.width <- .5
E(ig)$weight <- 1/(E(ig)$weight^2)
E(ig)$width <- exp(E(ig)$weight)+.6

allcomu <- cluster_louvain(ig)$memberships[1,]
V(ig)$color <- allcomu
E(ig)$color <- allcomu[head_of(ig,E(ig))]

plot(height.ras^1.9, col=col_ramp(50), legend=F, reset=F)
plot(height.wat, col="lightblue", add=T, legend=F)
plot(ig, add=T, rescale=F, alpha=.2)
```

Visualise sites on the 3D plot

```{r 3Dnetwork, eval=F}

plot_3d(
  hm_lamb,height_map, zscale=20, fov=0, theta=20, zoom=1, phi=45,
  windowsize=c(1000, 800), water=TRUE, waterdepth=mean(height),
  wateralpha=1, watercolor="lightblue", waterlinecolor="lightblue",
  waterlinealpha=.7
)

render_points(
  extent=raster::extent(raster::raster(height.ras)),
  lat=2.13828, long=-1.685547, altitude=extract(height.ras,sites)[,1]+20,
  zscale=20, size=1, col=categorical_pal(max(allcomu))[allcomu],
  clear_previous=TRUE
)

render_points(
  extent=raster::extent(raster::raster(height.ras)),
  lat=crds(sites)[,"y"], long=crds(sites)[,"x"], altitude=extract(height.ras, sites)[,1]+20,
  zscale=20, size=1, col=categorical_pal(max(allcomu))[allcomu],
  clear_previous=TRUE
)
       
render_highquality(point_radius=5, samples=256, clear=TRUE, light=TRUE)

```

let's say the climate modulate site size in a naive way:


```{r, animation.hook="gifski", interval=0.2, cache=TRUE, eval=F}
for(i in seq(1, tstep, length.out=100)){
	layout(matrix(1:2, nrow=2, ncol=1), heights=c(.8, .2))
	par(mar=c(1, 2, 2, 1))
	image(m, ann=F, axes=F, main=i)
	mtext(round(i), 3, 1)
	nsize <- size*faketemp[i] + size + 1
	points(crds(sites), bg="green", pch=21, cex=nsize)
	par(mar=c(2, 2, 0, 1))
	plot(
	  1:tstep, faketemp, type="l", ylab="temp", xlab="time", ylim=range(faketemp),
	  ann=F, frame.plot=F)
	points(i, faketemp[i], col="red", pch=20)
}
```


