# Population growth, conflicts and migrations {#popgrowth}

## Settlements position



```{r splitArea}
n <- 20 # we will create  20 sites
size <- rexp(n) #assign a size for each site following an exponential distribution 

```

Following what we set up during the previous chapter, we can  use `spatSample` from package `terra` to randomly select `n` points  (`r n` for this simulation) on the viable polygon we created.

```{r samplesites}
sites <- spatSample(viable, n) # generate n random points within the viable polygon and assign to variable sites
plot(height.ras^1.9, col=col_ramp(50), legend=F, reset=F)
plot(height.wat, col="lightblue", add=T, legend=F)
plot(sites, cex=2, pch=21, add=T, bg=rainbow(length(sites), alpha=.6))

# get only the non submerged actual land and plot it
above_level <- height.ras > mean(height.ras[])
coastline <- st_as_sf(as.polygons(above_level))[2,]
plot(coastline, col=NA, bgc=adjustcolor("cyan", 0.1), add=T)
```

In the context of [the archaeoriddle challenge](https://theia.arch.cam.ac.uk/archaeoriddle/), the files that describe this newly generated environment (the DEM and the coastline) as well as the resources raster,  needs to be given to the participant. This can be done as:

```{r writeSharedRessources}
foldname="data_toshare" #We will store all data that can/will be shared with participant in data_toshare
if(!file.exists(foldname))dir.create(foldname)
st_write(dsn=file.path(foldname,"coastline.shp"), coastline,append=F) # Write the coastline as a shapefile
writeRaster(filename=file.path(foldname,"dem_raster.tiff"), height.ras, overwrite=T) #write the DEM as a raster

writeRaster(filename=file.path(foldname,"resources.tiff"), ress, overwrite=T) #write the resources as a raster
```

The original data, saved and shared for the _archaeoriddle_ are stored in `data_original/`

## ~~ Network, Site Size And Climate -- DEPRECATED ~~

The following steps were designed but never used ; they can be quite fun and useful.

----

Create a network between sites  (requires `igraph`)

```{r sitesnetwork, eval=T}

library(igraph)
subsites  <- sites
crs(subsites) <- "+proj=lonlat"
subsites <- sites[sample(nrow(sites), 20), ]
ig <- graph_from_adjacency_matrix(
  as.matrix(distance(subsites)), weighted=T, mode="undirected")
ig <- set_graph_attr(ig, "layout", crds(subsites))

V(ig)$label <- ""
V(ig)$size <- (size+1)*5
V(ig)$frame.color <- "black"
V(ig)$frame.width <- .5
E(ig)$weight <- 1/(E(ig)$weight^2)
E(ig)$width <- exp(E(ig)$weight)+.6

allcomu <- cluster_louvain(ig)$memberships[1,]
V(ig)$color <- allcomu
E(ig)$color <- allcomu[head_of(ig,E(ig))]

plot(height.ras^1.9, col=col_ramp(50), legend=F, reset=F)
plot(height.wat, col="lightblue", add=T, legend=F)
plot(ig, add=T, rescale=F, alpha=.2)
```

Visualise sites on the 3D plot

```{r 3Dnetwork, eval=F}

plot_3d(
  hm_lamb,height_map, zscale=20, fov=0, theta=20, zoom=1, phi=45,
  windowsize=c(1000, 800), water=TRUE, waterdepth=mean(height),
  wateralpha=1, watercolor="lightblue", waterlinecolor="lightblue",
  waterlinealpha=.7
)

render_points(
  extent=raster::extent(raster::raster(height.ras)),
  lat=2.13828, long=-1.685547, altitude=extract(height.ras,sites)[,1]+20,
  zscale=20, size=1, col=categorical_pal(max(allcomu))[allcomu],
  clear_previous=TRUE
)

render_points(
  extent=raster::extent(raster::raster(height.ras)),
  lat=crds(sites)[,"y"], long=crds(sites)[,"x"], altitude=extract(height.ras, sites)[,1]+20,
  zscale=20, size=1, col=categorical_pal(max(allcomu))[allcomu],
  clear_previous=TRUE
)
       
render_highquality(point_radius=5, samples=256, clear=TRUE, light=TRUE)

```

let's say the climate modulate site size in a naive way:


```{r, animation.hook="gifski", interval=0.2, cache=TRUE, eval=F}
for(i in seq(1, tstep, length.out=100)){
	layout(matrix(1:2, nrow=2, ncol=1), heights=c(.8, .2))
	par(mar=c(1, 2, 2, 1))
	image(m, ann=F, axes=F, main=i)
	mtext(round(i), 3, 1)
	nsize <- size*faketemp[i] + size + 1
	points(crds(sites), bg="green", pch=21, cex=nsize)
	par(mar=c(2, 2, 0, 1))
	plot(
	  1:tstep, faketemp, type="l", ylab="temp", xlab="time", ylim=range(faketemp),
	  ann=F, frame.plot=F)
	points(i, faketemp[i], col="red", pch=20)
}
```
Population will then grow, for years and years. This implies multiple things: population need resources and a way to grow with regard to these resources and their own technologies.We initialised the ressource layer in the previous chapter 
in [the  section about ressources](#ressource). In the following sectinos we will use this ressource raster to initialise site size et define how ressource are used by the different culture.

## Initializing Site Population and Culture {#initsize}

Each site is then assigned as either belonging to the hunter-gather (HG) or farmer (F) culture.

To generate a more realistic distribution, all sites are initially, designated as hunter-gatherer sites, except for the one located in the most southwest position. Please note that this decision was based on the specific shape of the initial archaeoriddle's challenge. If needed, you may want to manually select the site culture for this specific location.


```{r setcultures,eval=T}
cultures <- rep("HG",length(sites))
sw <- which.min(sqrt((crds(sites)[,"x"] - min(crds(sites)[,"x"]))^2 + (crds(sites)[,"y"] - min(crds(sites)[,"y"]))^2))
cultures[sw]="F"
sites$culture <- cultures

plotMap(height.ras,height.wat)
points(crds(sites),bg=as.factor(cultures),pch=21,cex=1.3)
```


Each site will then be assigned a carrying capacity (how many people can leave on the site), probabilistically defined using the resource raster and the culture of the site.

```{r computeK}

Kbase <- c("HG"=45,"F"=120) #difference in base-K for the two cultures

# The base Ks  are then modulated given the ressource raster
sites$Ks <- initKs(Kbase, sites, ress, sizeexp="F")
```

Using these carrying capacities, we then set initial number of people, and the population structure for each sites:

```{r computeIN}
# Population size at initialisation, a number close to Ks
INs <- round(runif(length(sites), 0.85, 0.95)*sites$Ks)

# initialise population structure for all sites
Ips <- lapply(INs, initpopstruc)

ts <- 150

# We put the population structure across the years in a matrix: each row will represent a time step of the simulation ; each column a sites

Nts <- initlistsites(Ips, ts=ts)

```

We can look at the population structure of the biggest site:

```{r agestruc,out.width="40%",fig.show="hold"}
biggest <- which.max(INs)
pop.biggest <- Ips[[biggest]]
poplist=sapply(c(F="F",M="M"),function(s)table(cut(pop.biggest[pop.biggest[,2]==s,1],breaks=seq(0,40,4))))
par(mar=c(1,10,1,1))
barplot(-poplist[,1],beside=T,horiz=T,col=c("blue"),axes=F,las=1)
par(mar=c(1,1,1,10))
barplot(unname(poplist[,1]),beside=T,horiz=T,col=c("red"),axes=F)
```


You can see the code of `initKs`, `initpopstruct` and `initlistsites` functions here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
a = knitr::spin('../../R/init_simulation.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/init_simulation.Rmd', '.')

res = knitr::knit_child(input='init_simulation.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('init_simulation.Rmd')
```
</details>



## Base Population $Pop$-protocol {#base-pop-growth}

This is a central aspect of the model, and defines several parameters to control how population growth and reach a maximum population $\sim K$. It is based on three functions: The first function, `Gpd()`, is the core of the protocol. It receives a population `matrix` or `data.frame` with a specific number of individuals (`nrow`), their age (`df[,1]`) and their sex (`df[,2]`) and produces another object with the same format but with a different population based on the current one. Introduced in a loop, this function produces a time series of population, which ensures the Markov property of the stochastic process $p(x_n+1|x_n)$.

However, this function also relies on two the other functions: `death()` and `K_lim()`. `death()` is the process of people dying each year. It is based on an age-structured probability matrix, where infants and old people have more probabilities of deceasing. The matrix is provided, but the user can set its own. `K_lim()` limits the population when it reaches $K$. The user should specify the appropriate value for $K$ according to the size of the site, its type, the group economy, etc. Also, the user can change the probability that exceeding people dies with parameter `prob`. Default is 0.8.




These base function will then be used in higher level functions to update the population size given the size and ages defined and described in the previous [section](#initsize). 

A simple function  `Pop_stoch` implement a simple stochastic growth that uses these functions without the migrations and war mechanismes describe in this chapter. 


We can then use it to check and see the type of growth are generate given the default parameters (prob define the probability of people dying when the population has reached carrying capacity)

```{r test-popstoch,message=F}
set.seed(1234)
plot(1,1,ylim=c(0,180),xlim=c(0,50),type="n",xlab="time",ylab="pop size")
na=replicate(10, lines(Pop_stoch(pop_size=50, ts=50, K=150, W_fert_age = c(10, 45), M_fert_age = c(15, 55), p_offspring = 0.3, prob = 0.8)))
```


To know more about these functions and how their parameters are used, you can look at the manual given in the package:

```{r ,eval=F}
?Gpd
?death
?K_lim
?Pop_stoch
```

You can also use this [shiny app online]() to explore how these parameters interact to generate different growth curves.

### Source {-} 

If you want to delve into the code of the functions you can open it here:

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/population.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/population.Rmd', '.')
```
<details><summary>Show code</summary>
**Code**
```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='population.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('population.Rmd')
```
</details>


`changePopSize`: Change population size by moving population from one to another; or remove a group of people


## Migration

## Conflict and War
`whotouch`


## Run Simulation

We now have sites with two cultures, a environment and function to make population growth, migrate and engages into wars. Each function depends on a lot of paramaters, some have been reviewd before other are explain in the help of each function. Often the parameters can differs between the two cultures. He we gave a list of parameters, close to those used during the Archaeoriddle challenge: 

```{r}

#initialisation

Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures use and defined before ; will be use in simulation to define new site

# spatial penality for cultural extentions, ie for population to move out of initial site : lower, bigger penality
cul_ext <- c("HG"=7, "F"=6)

# penality of occupational area ie how hard it is to come close to your site ; if low, other sites can come close
penal_cul <- c("HG"=4, "F"=5)
# proba to give birth every year
prob_birth <- c("HG"=0.3, "F"=0.5)
# proba to die when pop > K
prob_survive <- c("HG"=0.8, "F"=0.6)
# proba to create new settlement when Ne > K
prob_split <- c("HG"=0.2, "F"=0.6)
# how big the group of migrant should be to create a new city vs
# migrate to a existing one
minimals <- c("HG"=0.14, "F"=0.20)
# prob to migrate to existing settlement when Ne > K
prob_move <- c("HG"=0.2,"F"=0.1)
```

Let's see our raster map again, and put the sites on top, with their ids and showing there initial population size:
```{r}
plotMap(height.ras,height.wat,paste0("year ", 0))
plot(sites, pch=21, add=T, bg=rainbow(2, alpha=0.6)[as.factor(sites$culture)],cex=(1+Nts[1,]/100))
text(sites,pos=3)
```

The `run_simulation` function will take the raster data and parameters for defined before for the two different cultures and run for the specified number of years `ts` (`r ts`).

It will then call the functions described above to model contacts between sites, the outcomes of fights and migrations. If you want 

```{r, eval=F}
?run_simulation
?whotouch
?fightbetterloss
?changePopSize
?warpoints
```

The general organisation of the algorithm is as follow:

> initialisation
> for all sites s:
>     N_s <- growth(N_s)
>     if(N_s>K_s):
>         moving people:
>         if( N_s - K-S > min_s * K_s and probsplit_s):
>             create new city:
>                 find the closest most attractive spot and move there
>         else if(proba move ):
>             migration
>     war(s)


```{r algo-general, echo=TRUE,eval=F}

1. ** initialisation **
  
2. for (i in 2:(ts+1))
    inactives <- (Nts[i-1,]==0)
    for ( s in sample(seq_along(sites)[!inactives]) : #among all active site
        N_s <- Gpd()
        if $N_s> K_s$ :
            migrants <- newN - round(Ks[[s]]*0.9)
            if (migrants >= (minimals[sites$culture[s]]*sites$Ks[s]) & runif(1)<prob_split[sites$culture[s]] ){
          #if subpopulation > 10 people, 10% chance of creation of a new city

          infarea <- (sqrt(tmp)+penal_cul[cultures]) * buffersettl
          buffersize <- rnorm(length(infarea), infarea, infarea * 0.1)
          buffersize[tmp==0] <- 0
          territory <- erase(viable, buffer(sites, buffersize))
          
          if( length(territory)>0 ){
            ##select a new site given its distance to the old one and the ressourcesource available in ressources
            d2 <- logisticdecay( sites[s], dem, x=20000*cul_ext[sites$culture[s]])
            w <- (0.7 * d2 + 0.3*ressources) / (0.7*minmax(d2)[2] + 0.3*minmax(ressources)[2])
            new_site <- spatSample(
              x=mask(
                w * logisticdecay(sites[s], dem, k=0.00002,
                                  x=20000*cul_ext[sites$culture[s]]),
                territory),
              size=1, method="weights", xy=T)[1:2]
            new_site <- vect(new_site, geom=c("x","y"))
            
            if ( length(new_site)>0 & all(!is.na(crds(new_site))) ){
              ##add new site to site listes
              Ips[[length(Ips)+1]] <- initpopstruc(n=migrants) #initialise a fake populaition, will be updated by real migrants later
              new_site$culture <- sites$culture[s]
              new_site$Ks <- round(initKs(
                Kbase, sites=new_site, ressources,
                sizeex="F", rate=0.45))
              sites <- rbind(sites, new_site)
              Ks[length(Ks)+1] <- new_site$Ks
            }
          }
        }
        ## if no creation of new city happen, there is a certain probability that people will move
        if( length(new_site)==0 && runif(1) < prob_move[sites$culture[s]] ){
           #migration to existing site
          att <- extract(ressources,sites)[,2]
          space <- sites$Ks - (Nts[i-1,] + migrants)
          dis <- extract(logisticdecay(sites[s], dem, k=0.00002, x=1), sites)[,2]
          attractivity <- att * space * dis
          #attractivity=attractivity*(1+10*(sites$culture[s]==sites$culture)) #4 times more likely to go to similar culture
          attractivity[s] <- min(attractivity)-1
          attractivity <- exp(attractivity)/sum(exp(attractivity))
          attractivity[Nts[i-1,]<10] <- 0 
          attractivity[sites$culture!=sites$culture[s]] <- 0 
          if(any(is.na(attractivity))){
            print(attractivity)
            attractivity[is.na(attractivity)] <- 0
          }
          
          city <- sample(size=1, x=seq_along(sites), prob=attractivity)
          Nts[i,city] <- Nts[i-1,city] + migrants
        }
        if( havemoved ){
          Ips[c(s,city)] <- changePopSize( loosingPop=Ips[[s]], winingPop=Ips[[city]], size=migrants)
          newN <- newN - migrants
        }
        
      }
      Nts[i,s] <- newN
      
    }
    ## WAR =======================
    potentialfighters <- which(sites$culture=="F" & Nts[i,]>50)
    for (s in sample(x=potentialfighters, size=round(length(potentialfighters)*0.1))){
      buff <- bufferatack
      potentialvictims <- which(sites$culture !=sites$culture[s] & Nts[i,]>0) 
      clash <- whotouch(s, sites, Ne=Nts[i,], buffersize=buff)
      if(length(clash)>0 && !is.na(clash)){
        if(length(clash) == 1){
          attack <- clash
        } else {
          attack <- sample(clash, 1)
        }
        newns <- fightbetterloss(Ne=Nts[i,], a=s, b=attack)
        casualties <- sum(Nts[i, c(s,attack)] - newns[c(s,attack)])
        warcasualties[i] <- casualties
        sizew <- casualties^2/4000
        warpoints(sites, s, attack, Ne=Nts[i,],
                  buffersize=buff, sizewar=sizew+0.5)
        
        #effectively kill people in population (should be done taking into account age pyramid to be more realistic)
        Ips[[s]] <- changePopSize(loosingPop=Ips[[s]],
                                  size=(Nts[i,s] - newns[s]))
        Ips[[attack]] <- changePopSize(loosingPop=Ips[[attack]],
                                       size=(Nts[i, attack] - newns[attack]))
        Nts[i,] <- newns
      }
  }
  return(list(Nts=Nts, warcasualties=warcasualties, Ips=Ips, sites=sites))
}
```

The full code of the function can be visualised here:

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/natural_deposition.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/natural_deposition.Rmd', '.')
```

<details><summary>Show code</summary>
**Code**
```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='natural_deposition.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('natural_deposition.Rmd')
```
</details>


```{r single-run-ex1,eval=T}
onesimu <- run_simulation(
  sites=sites, viable=viable, dem=height.ras,
  ressources=ress,
  water=height.wat,
  foldervid="nan",
  visu=F, visumin=F,
  ts=ts, #length of simulation in year
  Kbase=c("HG"=35, "F"=110), #difference in K for the two cultures
  cul_ext=c("HG"=7, "F"=6), #spatial penality to extent: lower, bigger penality
  penal_cul=c("HG"=4, "F"=5), #penality of occupational area: low, other sites can cam close
  prob_birth=c("HG"=0.3, "F"=0.5), #proba of giving birth every year
  prob_survive=c("HG"=0.8, "F"=0.65), #proba of dying when pop > K
  prob_split=c("HG"=0.5, "F"=0.6), #proba of creating a new settlement when Ne > K
  minimals=c("HG"=0.14,"F"=0.20), #how big the group of migrant should be to create a new city vs migrate to a existing one 
  bufferatack=300, #distance max around which settlement can fight
  prob_move=c("HG"=0.2, "F"=0.1) #proba of migrating to existing settlement when Ne > K
)
```


```{r , echo=FALSE, message=FALSE}
a = knitr::spin('../../R/run_simulation.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/run_simulation.Rmd', '.')
```
### Source code {-}

Check the code of the simulation functions here:
<details><summary>Show code</summary>
**Code**
```{r print-run_sim, echo=FALSE, results='asis'}
res = knitr::knit_child(input='run_simulation.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('run_simulation.Rmd')
```
</details>


### Explore results:

The list `onesimu`  contains a few interesting info:


```{r explore-ex1-1,cache=F}
Nts <- onesimu$Nts # population at each timestep
warcasualties <- onesimu$warcasualties #death by war at each time step
```


Use ressource to adjust the Ks:
```{r explore-ex1-b,eval=T}
plot(sites, cex=(1+Nts[1,]/100), pch=21, bg=rainbow(2, alpha=0.6)[as.factor(sites$culture)])
```

We look at the simulation step by stepk
Step by step to allow interaction between sites:

```{r explore-ex1-trajpersite,eval=T,messages=F}
## visualise trajectories for each site
i <- ts
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)), xlab="time", ylab="popsize")
lapply(1:ncol(Nts),
       function(j)lines(Nts[,j],col=rainbow(2)[as.factor(sites$culture)[j]])
)

```

Visualise, Record deposit and loss for the 5 first sites (see next chapters to learn about  `A_rates` and `Rec_c`).
```{r,out.width="25%"}
alldeposit <- lapply(
  1:5,
  function(i){
    Rec_c(sapply(Nts[,i], A_rates), InitBP = 15000, ts=ts, r=0.2, max_bone_thickness="m")
  }
)
maxy <- max(sapply(alldeposit, function(i)max(apply(i,2,sum))))

nill <- lapply(alldeposit, function(depo)barplot(t(depo), col=viridis(ts+1), ylim=c(0,maxy)))
```

Population by time and culture

```{r,eval=T}
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)),
     xlab="time", ylab="popsize")
lapply(1:ncol(Nts),
       function(i)lines(Nts[,i], col=rainbow(2)[as.factor(sites$culture)[i]]))
plot(apply(Nts, 1, sum)[1:i], xlab="time", ylab="popsize")

```

Exploring some trajectories of the simulations

```{r, eval=T}
i <- nrow(Nts)  # Get the number of rows in Nts
plot(apply(Nts[1:i, sites$culture=="F",drop=F], 1, sum), col="red", type="l",
     lwd=2, ylim=c(0, max(apply(Nts, 1, sum))))  # Plot sum of 'F' culture values
points(apply(Nts[1:i, sites$culture=="HG"], 1, sum),
       col="blue", lwd=2, type="l")  # Add points for 'HG' culture values
plot(warcasualties[1:(i-1)], lwd=2, col="green", type="h", yaxt="n", ylab="")  # Plot war casualties
axis(4)  # Add an axis on the right side
par(new=T)  # Allow a new plot to be drawn on the existing plot
growF <- apply(Nts[1:(i-1), sites$culture=="F",drop=F], 1, sum)  # Sum of 'F' culture values excluding last row
growHG <- apply(Nts[1:(i-1), sites$culture=="HG",drop=F], 1, sum)  # Sum of 'HG' culture values excluding last row
plot(growF, col="red", type="l", lwd=2, ylim=c(0, max(growF, growHG)))  # Plot growth of 'F' culture
points(growHG, col="blue", lwd=2, type="l")  # Add points for growth of 'HG' culture
```
