# Population structure & growth {#popgrowth}

## Settlements position


```{r splitArea}
n <- 20 # we will create  20 sites
size <- rexp(n) #assign a size for each site following an exponential distribution 

```

Following what we set up during the previous chapter, we can  use `spatSample` from package `terra` to randomly select `n` points  (`r n` for this simulation) on the viable polygon we created.

```{r samplesites}
sites <- spatSample(viable, n) # generate n random points within the viable polygon and assign to variable sites
plot(height.ras, col=col_ramp(20), legend=F, reset=F)
plot(height.wat, col="lightblue", add=T, legend=F)
plot(sites, cex=2, pch=21, add=T, bg=rainbow(length(sites), alpha=.6))

# get only the non submerged actual land and plot it
above_level <- height.ras > mean(height.ras[])
coastline <- st_as_sf(as.polygons(above_level))[2,]
plot(coastline, col=NA, bgc=adjustcolor("cyan", 0.1), add=T)
```

In the context of [the archaeoriddle challenge](https://theia.arch.cam.ac.uk/archaeoriddle/), the files that describe this newly generated environment (the DEM and the coastline) as well as the resources raster,  needs to be given to the participant. This can be done as:

```{r writeSharedRessources}
foldname="data_toshare" #We will store all data that can/will be shared with participant in data_toshare
if(!file.exists(foldname))dir.create(foldname)
st_write(dsn=file.path(foldname,"coastline.shp"), coastline,append=F) # Write the coastline as a shapefile
writeRaster(filename=file.path(foldname,"dem_raster.tiff"), height.ras, overwrite=T) #write the DEM as a raster

writeRaster(filename=file.path(foldname,"resources.tiff"), ress, overwrite=T) #write the resources as a raster
```

The original data, saved and shared for the _archaeoriddle_ are stored in `data_original/`

## Initializing Site Population and Culture {#initsize}

Each site is then assigned as either belonging to the hunter-gather (HG) or farmer (F) culture.

To generate a more realistic distribution, all sites are initially, designated as hunter-gatherer sites, except for the one located in the most southwest position. Please note that this decision was based on the specific shape of the initial archaeoriddle's challenge. If needed, you may want to manually select the site culture for this specific location.


```{r setcultures,eval=T}
cultures <- rep("HG",length(sites))
sw <- which.min(sqrt((crds(sites)[,"x"] - min(crds(sites)[,"x"]))^2 + (crds(sites)[,"y"] - min(crds(sites)[,"y"]))^2))
cultures[sw]="F"
sites$culture <- cultures

plotMap(height.ras,height.wat)
points(crds(sites),bg=as.factor(cultures),pch=21,cex=1.3)
```


Each site will then be assigned a carrying capacity (how many people can leave on the site), probabilistically defined using the resource raster and the culture of the site.

```{r computeK}

Kbase <- c("HG"=45,"F"=120) #difference in base-K for the two cultures

# The base Ks  are then modulated given the ressource raster
sites$Ks <- initKs(Kbase, sites, ress, sizeexp="F")
```

Using these carrying capacities, we then set initial number of people, and the population structure for each sites:

```{r computeIN}
# Population size at initialisation, a number close to Ks
INs <- round(runif(length(sites), 0.85, 0.95)*sites$Ks)

# initialise population structure for all sites
Ips <- lapply(INs, initpopstruc)

ts <- 250

# We put the population structure across the years in a matrix: each row will represent a time step of the simulation ; each column a sites

Nts <- initlistsites(Ips, ts=ts)

```

We can look at the population structure of the biggest site:

```{r agestruc,out.width="40%",fig.show="hold"}
biggest <- which.max(INs)
pop.biggest <- Ips[[biggest]]
poplist=sapply(c(F="F",M="M"),function(s)table(cut(pop.biggest[pop.biggest[,2]==s,1],breaks=seq(0,40,4))))
par(mar=c(1,10,1,1))
barplot(-poplist[,1],beside=T,horiz=T,col=c("blue"),axes=F,las=1)
par(mar=c(1,1,1,10))
barplot(unname(poplist[,1]),beside=T,horiz=T,col=c("red"),axes=F)
```


You can see the code of `initKs`, `initpopstruct` and `initlistsites` functions here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
a = knitr::spin('../../R/init_simulation.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/init_simulation.Rmd', '.')

res = knitr::knit_child(input='init_simulation.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('init_simulation.Rmd')
```
</details>



## Population Growth : $Pop$-protocol {#base-pop-growth}

This is a central aspect of the model, and defines several parameters to control how population growth and reach a maximum population $\sim K$. It is based on three functions: The first function, `Gpd()`, is the core of the protocol. It receives a population `matrix` or `data.frame` with a specific number of individuals (`nrow`), their age (`df[,1]`) and their sex (`df[,2]`) and produces another object with the same format but with a different population based on the current one. Introduced in a loop, this function produces a time series of population, which ensures the Markov property of the stochastic process $p(x_n+1|x_n)$.

However, this function also relies on two the other functions: `death()` and `K_lim()`. `death()` is the process of people dying each year. It is based on an age-structured probability matrix, where infants and old people have more probabilities of deceasing. The matrix is provided, but the user can set its own. `K_lim()` limits the population when it reaches $K$. The user should specify the appropriate value for $K$ according to the size of the site, its type, the group economy, etc. Also, the user can change the probability that exceeding people dies with parameter `prob`. Default is 0.8.




These base function will then be used in higher level functions to update the population size given the size and ages defined and described in the previous [section](#initsize). 

A simple function  `Pop_stoch` implement a simple stochastic growth that uses these functions without the migrations and war mechanismes describe in this chapter. 


We can then use it to check and see the type of growth are generate given the default parameters (prob define the probability of people dying when the population has reached carrying capacity)

```{r test-popstoch,message=F}
set.seed(1234)
plot(1,1,ylim=c(0,180),xlim=c(0,50),type="n",xlab="time",ylab="pop size")
na=replicate(10, lines(Pop_stoch(pop_size=50, ts=50, K=150, W_fert_age = c(10, 45), M_fert_age = c(15, 55), p_offspring = 0.3, prob = 0.8)))
```


To know more about these functions and how their parameters are used, you can look at the manual given in the package:

```{r ,eval=F}
?Gpd
?death
?K_lim
?Pop_stoch
```

You can also use this [shiny app online]() to explore how these parameters interact to generate different growth curves.

### Source {-} 

If you want to delve into the code of the functions you can open it here:

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/population.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/population.Rmd', '.')
```
<details><summary>Show code</summary>
**Code**
```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='population.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('population.Rmd')
```
</details>


