# Population growth on multiple sites

## Environmental resources
We generate zones of high resources, that will attract settlements and support higher K.
The resources will be located at given hotspots with an area around in which they decay
with a logistic function.

```{r, echo=FALSE}
?logisticdecay
```


```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/logistic_decay.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/logistic_decay.Rmd', 'spin_functions/')
```

Here's the code for the generation of resources under logistic decay 
<details><summary>Show code</summary>
**Code**
```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='spin_functions/logistic_decay.Rmd', quiet=TRUE)
cat(res, sep = '\n')
```
</details>
<br>

```{r}

##ressource geolocalisation set manually
goodresources <- vect(
  cbind(
    x=c(-0.2300711, -3.1455282, -0.5086485, -1.9639755,
        -0.4077843, 0.019688, -3.116710),
    y=c(3.6423000, -0.2551019, -0.7440748, 1.1303214,
        1.0248567, 0.2194895, 2.0267718)
  )
)
#spread of resources
areas <- 4 * c(100, 2000, 200, 300, 100, 200, 400)
#speed of ressource decay:
ks <- c(0.0002, 0.0001000, 0.0001600, 0.0001800, 0.00040, .0002, 0.0002)/4

crs(goodresources) <- crs(height.ras)

allres <- lapply(
  seq_along(goodresources),
  function(i){
    logisticdecay(goodresources[i], mask(height.ras, viable),
                  x=areas[i], k=ks[i]
    )
  }
)

allna <- sapply(allres, function(i)any(is.na(values(i))))
allres <- allres[!allna]

ress <- allres[[1]]
for(i in 2:length(allres))
    ress <- ress + allres[[i]]
ress <- mask(ress,viable)
plot(ress)
```

## Initial sites and size:
We initialize the sites of 2 cultires, `HG` and `F`.

<br>

```{r, echo=FALSE}
?initKs
```
<br>

```{r, echo=FALSE}
?initpopstruc
```
<br>

```{r, echo=FALSE}
?initlistsites
```
<br>

```{r,eval=T}
ts <- 250
#a manual way to set cultures given general geography
cultures <- rep("HG",length(sites))
cultures[(crds(sites)[,"x"] < -1 & crds(sites)[,"y"] < 1)]="F"
sites$culture <- cultures
Kbase <- c("HG"=45,"F"=120) #difference in K for the two cultures
# The initial Ks depends on the resources in the area
sites$Ks <- initKs(Kbase, sites, ress, sizeexp="F")
#saveRDS(file="data_original/sitesinitialposition/", sites)

# Optionally we can intialize here our list of sites and population structures
# Otherwise, they are initialized within the run_simulation function
# Population size at initialisation, a number close to Ks
INs <- round(runif(length(sites), 0.85, 0.95)*sites$Ks)
# initialise population structure for all sites
Ips <- lapply(INs, initpopstruc)
# Not we put the population structure across the years in a list
Nts <- initlistsites(Ips, ts=ts)

```

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/init_simulation.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/init_simulation.Rmd', 'spin_functions/')
```

You can see the code of `initKs`, `initpopstruct` and `initlistsites` functions here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='spin_functions/init_simulation.Rmd', quiet=TRUE)
cat(res, sep = '\n')
```
</details>


Now we have sites with two cultures, initialise the parameters we will use:

```{r}

#(read sites saved before)
# sites=readRDS(file="data_original/sitesinitialposition/",sites)


#initialisation

ts <- 250
# Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures

# spatial penality to extent: lower, bigger penality
cul_ext <- c("HG"=7, "F"=6)
# penality of occupational area: low, other sites can come close
penal_cul <- c("HG"=4, "F"=5)
# proba to give birth every year
prob_birth <- c("HG"=0.3, "F"=0.5)
# proba to die when pop > K
prob_survive <- c("HG"=0.8, "F"=0.6)
# proba to create new settlement when Ne > K
prob_split <- c("HG"=0.2, "F"=0.6)
# how big the group of migrant should be to create a new city vs
# migrate to a existing one
minimals <- c("HG"=0.14, "F"=0.20)
# prob to migrate to existing settlement when Ne > K
prob_move <- c("HG"=0.2,"F"=0.1)
```

Let's create again our raster map from the `height.ras` from the previous chapter
```{r}
height.wat <- height.ras
height.wat[height.wat>mean(height.wat[])] <- NA
height.groups <- height.ras
maxh <- max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])] <- NA
height.groups[height.groups<(maxh*.7)] <- 1
height.groups[height.groups>(maxh*.7)] <- 200
height.groups[is.na(height.groups)] <- -1
height.poly <- as.polygons(height.groups)
viable <- makeValid(height.poly[2,])
plotMap(height.ras,height.wat,paste0("year ", 0))
plot(sites, pch=21, add=T, bg=rainbow(2, alpha=0.6)[as.factor(sites$culture)])
text(sites)
```

The `run_simulation` function will take the raster data and parameters for the
different populations and run for the specified number of years.

<br>

```{r, echo=FALSE}
?run_simulation
```
<br>

It uses a number of other functions to model contacts between sites, the outcomes of fights and migrations

<br>

```{r, echo=FALSE}
?whotouch
```
<br>

<br>

```{r, echo=FALSE}
?fightbetterloss
```
<br>

<br>

```{r, echo=FALSE}
?changePopSize
```
<br>

<br>

```{r, echo=FALSE}
?warpoints
```
<br>

Or we can run a simple simulations using the saved elevation data.

```{r,eval=T}

height.ras <- rast("data_original/east_narnia4x.tif")
height.wat <- height.ras
height.wat[height.wat>mean(height.wat[])] <- NA
height.groups <- height.ras
maxh <- max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])] <- NA
height.groups[height.groups<(maxh*.7)] <- 1
height.groups[height.groups>(maxh*.7)] <- 200
height.groups[is.na(height.groups)] <- -1
height.poly <- as.polygons(height.groups)
viable <- makeValid(height.poly[2,])
sites <- vect("data_original/sitesinitialposition/")
ts <- 10

print(paste0("Starting simulation ","nan"))

onesimu <- run_simulation(
  sites=sites, viable=viable, dem=height.ras,
  ressources=rast("data_original/resources.tiff"),
  water=height.wat,
  foldervid="nan",
  visu=F, visumin=F,
  ts=ts, #length of simulation in year
  Kbase=c("HG"=35, "F"=110), #difference in K for the two cultures
  cul_ext=c("HG"=7, "F"=6), #spatial penality to extent: lower, bigger penality
  penal_cul=c("HG"=4, "F"=5), #penality of occupational area: low, other sites can cam close
  prob_birth=c("HG"=0.3, "F"=0.5), #proba of giving birth every year
  prob_survive=c("HG"=0.8, "F"=0.65), #proba of dying when pop > K
  prob_split=c("HG"=0.5, "F"=0.6), #proba of creating a new settlement when Ne > K
  minimals=c("HG"=0.14,"F"=0.20), #how big the group of migrant should be to create a new city vs migrate to a existing one 
  bufferatack=300, #distance max around which settlement can fight
  prob_move=c("HG"=0.2, "F"=0.1) #proba of migrating to existing settlement when Ne > K
)
```

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/run_simulation.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/run_simulation.Rmd', 'spin_functions/')
```

Check the code of the simulation functions here:
<details><summary>Show code</summary>
**Code**
```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='spin_functions/run_simulation.Rmd', quiet=TRUE)
cat(res, sep = '\n')
```
</details>


The list `onesimu`  contains a few interesting info:


```{r,cache=F}
Nts <- onesimu$Nts # population at each timestep
warcasualties <- onesimu$warcasualties #death by war at each time step
```


Use ressource to adjust the Ks:
```{r,eval=T}
plot(sites, cex=(1+Nts[1,]/100), pch=21, bg=rainbow(2, alpha=0.6)[as.factor(sites$culture)])
```


We look at the simulation step by stepk
Step by step to allow interaction between sites:

```{r,eval=T,messages=F}
## Run stochastic process  


#par(mfrow=c(2,1))
i <- ncol(Nts)
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)), xlab="time", ylab="popsize")
lapply(1:ncol(Nts),
       function(j)lines(Nts[,j],col=rainbow(2)[as.factor(sites$culture)[j]])
)

```

Visualise, Record deposit and  loss. for the 5 first sites
```{r,out.width="25%"}
alldeposit <- lapply(
  1:5,
  function(i){
    Rec_c(sapply(Nts[,i], A_rates), InitBP = 15000, ts=ts, r=0.2, max_bone_thickness="m")
  }
)
maxy <- max(sapply(alldeposit, function(i)max(apply(i,2,sum))))

nill <- lapply(alldeposit, function(depo)barplot(t(depo), col=viridis(ts+1), ylim=c(0,maxy)))
```

Population by time and culture

```{r,eval=T}
par(mfrow=c(2,1))
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)),
     xlab="time", ylab="popsize")
lapply(1:ncol(Nts),
       function(i)lines(Nts[,i], col=rainbow(2)[as.factor(sites$culture)[i]]))
plot(apply(Nts, 1, sum)[1:i], xlab="time", ylab="popsize")

```


Below some usefull code to manually select the ressource origins on the map (by clicking on the screen) and explore different decay around them

```{r, eval=F}
plotMap(height.ras, height.wat, paste0("year ", 0))
new_site <- vect(as.data.frame(locator()), geom=c("x", "y"))  # select point on the sscreep
crs(new_site) <- crs(height.ras)
d2 <- logisticdecay(new_site, height.ras, x=20000) # generate a logistic decay around the point

plot(d2)
w <- (0.7*d2 + 0.3*ress) / (0.7*minmax(d2)[2] + 0.3*minmax(ress)[2])
plot(w)
plot(mask(w*logisticdecay(new_site, height.ras, k=0.00002, x=1), territory))
multisamp <- sapply(
  1:1000,
  function(i){
    spatSample(x=w, size=1, method="weights", xy=T)[1:2]
  }
)


```

Exploring some trajectories of the simulations

```{r, eval=T}
i <- nrow(Nts)
plot(apply(Nts[1:i, sites$culture=="F"], 1, sum), col="red", type="l",
     lwd=2, ylim=c(0, max(apply(Nts, 1, sum))))
points(apply(Nts[1:i, sites$culture=="HG"], 1, sum),
       col="blue", lwd=2, type="l")
plot(warcasualties[1:(i-1)], lwd=2, col="green", type="h", yaxt="n", ylab="")
axis(4)
par(new=T)
growF <- apply(Nts[1:(i-1), sites$culture=="F"], 1, sum)
growHG <- apply(Nts[1:(i-1), sites$culture=="HG"], 1, sum)
plot(growF, col="red", type="l", lwd=2, ylim=c(0, max(growF, growHG)))
points(growHG, col="blue", lwd=2, type="l")
```


