# Populatin grow on multiple sites

All the function below are mere illustration, the exact functions, used for the final simulations are in `tools.R`

```{r,eval=F}
## Function 11. Logistic decay
#' @title Logistic decay
#' @param pt a vector point around witch decay is computed
#' @param rast a raster to compute distances
#' @param L starting point of decay
logisticdecay <- function(pt, rast, L = 1, k = 0.0001, x0 = 60000){
  ds <- distance(rast, pt)
  decay <- L-L/(1+exp(-k*(ds-x0)))
  return(decay)
}

## Function 12. Create population matrix
#' @title Create initial population matrix
#' @param n number of individuals
#' @param ages initial ages
#' @param p_sex proportion of the different sex
#' @return a dataframe with age and sex of every individual of the population
initpopstruc <- function(n=100, ages=10:30, p_sex=c(0.5,0.5)){
  data.frame("Age" = sample(ages, n, ages, replace = TRUE),
             "Sex" = sample(c("M", "F"), n, prob=p_sex, replace=TRUE))
}

## Function 13. Create initial list of sites
#' @title Create initial list of sites
#' @param list_sites a list of population for all initial sites of the simulation
#' @param ts length of simulation
initlistsites <- function(list_sites, ts = 200){
  Nts <- matrix(0, nrow=ts + 1, ncol=length(list_sites))
  Nts[1,] <- sapply(list_sites, nrow)
  return(Nts)
}


#
# This should be change by a funciton that allow to:
# sample the existing pop and kill some, swapp between two pop

## Function 14. Samples through an existing population
#' @title Sample through an existing population
#'
#' @param loosingPop population matrix that will loose members
#' @param winingPop population matrix that will gain members (optional)
#' @param size size of moving/killed population
#' @param new 
#' @param method 
#' @param probs Density function to generate probabilities of dying
#' according to age. Default is dnorm with age as quantiles 
#' @param prob.option List with elements `mean` and `sd` to use in the `prob` function
#' @return Updated loosing and winning populations
#' @export
#'
#' @examples
changePopSize <- function(
  loosingPop, winingPop = NULL, size, new = F, method = "random",
  probs = dnorm, prob.option = list("sd"=10, "mean"=22)
  ) {
  #print(dim(loosingPop))
  #if(!is.null(winingPop))
  #    print(dim(winingPop))
  #if(length(size)==0 || size==0)return(data.frame(Age=numeric(),Sex=character()))
  if(nrow(loosingPop)==0){ 
    kill <- 0
  } else if(method=="random"){
    kill <- sample(
      x=1:nrow(loosingPop), size=size,
      prob=probs(
        loosingPop$Age, mean=prob.option$mean, sd=prob.option$sd))
  }
  #print(paste("diff",nrow(popdistrib)-size,"new",size))
  if(!is.null(winingPop)){
    winingPop <- rbind(winingPop, loosingPop[kill, ])
  }
  loosingPop=loosingPop[-kill, ]
  if(!is.null(winingPop)) {
    return(list(loosingPop, winingPop))
  } else {
    return(loosingPop)
  }
}

# ## Function 15. Initial carrying capacity
# #' @param Kbase baseline carrying capacity or different culturesj
# #' @param sites raster with site and culture
# #' @param resources a raster type for fall sites
# initKs <- function(Kbase = c("HG"=30, "F"=120), sites, resources, sizeexp=NULL){
#   Ks <- round( Kbase[sites$culture] + rnorm(length(sites), 0, 10) )
#   while(any(Ks<1)) {
#     Ks <- round( Kbase[sites$culture] + rnorm(length(sites), 0, 10) )
#   }
#   #Ks[sites$culture=="F"]=Ks[sites$culture=="F"]*runif(sum(sites$culture=="F"),1,1)
#   tmp <- Ks*( 1 + extract(resources, sites)[,2] )
#   if(!is.null(sizeexp)){
#     tmp[sites$culture==sizeexp] <- (
#       Ks[sites$culture==sizeexp] *
#         (1 + rexp(sum(sites$culture==sizeexp), rate=.3) *
#           extract(resources, sites[sites$culture==sizeexp])[, 2])
#     )
#   }
#   return(tmp)
# }


initKs <- function(Kbase=c("HG"=30,"F"=120),sites,ressources,sizeexp=NULL){
    Ks=round(Kbase[sites$culture]+rnorm(length(sites),0,10))
    while(any(Ks<1)) Ks=round(Kbase[sites$culture]+rnorm(length(sites),0,10))
    #Ks[sites$culture=="F"]=Ks[sites$culture=="F"]*runif(sum(sites$culture=="F"),1,1)
    tmp=Ks*(1+extract(ressources,sites)[,2])
    if(!is.null(sizeexp))tmp[sites$culture==sizeexp]=Ks[sites$culture==sizeexp]*(1+rexp(sum(sites$culture==sizeexp),rate=.3)*extract(ressources,sites[sites$culture==sizeexp])[,2])
    tmp
}



plotMap <- function(height, water, maintitle=""){
  plot(height^1.9, col=col_ramp(50), legend=F, reset=F, main=maintitle)
  plot(water, col="lightblue", add=T, legend=F)
}


#' @title Check who is touching a given site
#' @param i index of the site checked
#' @param sites raster with site coordinates,  cultures and karying capacityes
#' @param homophily if true, return all sites that touch (even same culture)
#' @param Ne population size of all sites (could be remove if Ne was stored in the raster)
#' @return A list of sites touching site i
#' @export
whotouch <- function(i, sites, homophily=F, buffersize=200, Ne){
  touch <- st_intersects(
    st_make_valid(st_as_sf(buffer(sites[i], Ne[i]*buffersize))),
    st_make_valid(st_as_sf(buffer(sites, Ne*buffersize))))
  # If there are 1 or more sites touching i
  if(length(touch)>0){
    enemies <- unlist(touch)
    if(homophily) {
      enemies <- enemies[enemies!=i]
    } else {
      enemies <- enemies[sites$culture[enemies]!=sites$culture[i]]
    }
  }
  else {
    enemies <- NA
  }
  return(enemies)
}

#' @tile A simple fight between sites
#' A function to compute lost during a fighting.
#' Who's the winner and looser depend on population size
#' Decrease in population size is simply calculated with a binomial and a prob
#' of individual survival for the winner and the looser
#' @param Ne list of population size for the fighting settlement
#' @param a indice of the first settlement
#' @param b indice of the second settlement
#' @return the updated population size of both settlements engaged in the fight
simplefight <- function(Ne, a, b){
  # Compute who wins randomly, but based on population sizes
  if(runif(1)<Ne[a]/(Ne[a]+Ne[b])){
    v <- a
    l <- b
  }
  else{
    v <- b
    l <- a
  }
  one <- Ne
  Ne[v] <- rbinom(n=1, prob=.9, size=Ne[v])
  Ne[l] <- rbinom(n=1, prob=.4, size=Ne[l])
  print(
    paste(
      "victory", v, "(", one[v], "-", Ne[v], ") over",
      l,"(", one[l], "-", Ne[l], "), total of:",
      (one[v]-Ne[v]) + (one[l]-Ne[l]), "people"))
  return(Ne)
}   


#' @tile A simple fight between sites
#' A function to compute lost during a fighting.
#' Who's the winner and looser depend on population size.
#' Decrease in population size is simply calculated with a binomial and a prob
#' of individual survival for the winner and the looser. The probabilities depend
#' on the the relative population size of each size.
#' @param Ne list of population size for the fighting settlement
#' @param a indice of the first settlement
#' @param b indice of the second settlement
#' @return the updated population size of both settlements engaged in the fight
fightbetterloss <- function(Ne,a,b){
  if (runif(1)<Ne[a]/(Ne[a]+Ne[b])) {
    v=a
    l=b
  }
  else{
    v=b
    l=a
  }
  one <- Ne
  Ne[v] <- rbinom(n=1, prob=1-Ne[l]/(Ne[v] + Ne[l]), size=Ne[v])
  Ne[l] <- rbinom(n=1, prob=1-Ne[v]/(Ne[v] + Ne[l]), size=Ne[l])
  print(
    paste(
      "victory", v, "(", one[v], "-", Ne[v], ") over",
      l, "(", one[l], "-", Ne[l], "), total of:",
      (one[v]-Ne[v]) + (one[l]-Ne[l]), "people"))
  return(Ne)
}   


##draw a war symbole where two clans are fighting
warpoints <- function(sites, a, b, Ne, buffersize=300, plot=T, sizewar=2) {
    meetpoints <- crop(
      buffer(sites[a], 1+Ne[a]*buffersize),
      buffer(sites[b], 1+Ne[b]*buffersize)
    )
    if (length(meetpoints)>0) {
      p <- spatSample(meetpoints,1)
      if (plot & length(p)>0) {
        plot(p, add=T, bg="red", pch="🔥", cex=sizewar,
             col=adjustcolor("yellow", 0.1))
        
        plot(p, add=T, bg="yellow", pch="⚔️",cex=sizewar)
      }
      return(p)
    } else NULL
}




```


## Environmental resources
We generate zone of hight resource, that will attract settlements and support higher K

```{r}

source("tools.R")
##ressource geolocalisation set manually
goodresources=vect(
                    cbind(
                     x=c(-0.2300711,-3.1455282,-0.5086485,-1.9639755,-0.4077843,0.019688,-3.116710),
                     y=c(3.6423000,-0.2551019,-0.7440748,1.1303214,1.0248567,0.2194895,2.0267718)
                     )
                    )
#spread of resources
areas=4*c(100,2000,200,300,100,200,400)
#speed of ressource decay:
ks=c(0.0002,0.0001000,0.0001600,0.0001800,0.00040,.0002,0.0002)/4

crs(goodresources)=crs(height.ras)

allres=lapply(seq_along(goodresources),function(i)logisticdecay(goodresources[i],mask(height.ras,viable),x=areas[i],k=ks[i]))

allna=sapply(allres,function(i)any(is.na(values(i))))
allres=allres[!allna]

ress=allres[[1]]
for(i in 2:length(allres))
    ress=ress+allres[[i]]
ress=mask(ress,viable)
plot(ress)
```
## Initial sites and size:

```{r,eval=T}
#a manual way to set cultures given general geography
cultures=rep("HG",length(sites))
cultures[(crds(sites)[,"x"] < -1 & crds(sites)[,"y"] < 1)]="F"
sites$culture=cultures
Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures
sites$Ks=initKs(Kbase,sites,ress,sizeexp="F")
#saveRDS(file="data_original/sitesinitialposition/",sites)
```

Now we have sites with two cultures, initialise the parameters we will use:

```{r}

#(read sites saved before)
# sites=readRDS(file="data_original/sitesinitialposition/",sites)


#initialisation

ts=250
Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures
cul_ext=c("HG"=7,"F"=6) #spatial penality to extent: lower, bigger penality
penal_cul=c("HG"=4,"F"=5) #penality of occupational area: low, other sites can cam close
prob_birth=c("HG"=0.3,"F"=0.5) #proba to give birth every year
prob_survive=c("HG"=0.8,"F"=0.6) #proba to die when pop > K
prob_split=c("HG"=.2,"F"=.6) #proba to create new settlement when Ne > K
minimals=c("HG"=.14,"F"=.20) #how big the group of migrant should be to create a new city vs migrate to a existing one 
prob_move=c("HG"=0.2,"F"=0.1) #proba to migrate to existing settlement when Ne > K

height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA
height.groups=height.ras
maxh=max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])]=NA
height.groups[height.groups<(maxh*.7)]=1
height.groups[height.groups>(maxh*.7)]=200
height.groups[is.na(height.groups)]=-1
height.poly=as.polygons(height.groups)
viable=makeValid(height.poly[2,])
plotMap(height.ras,height.wat,paste0("year ",0))
plot(sites,pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
text(sites)
```

We can run a simple simulations using this:


```{r,eval=T}


height.ras=rast("data_original/east_narnia4x.tif")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA
height.groups=height.ras
maxh=max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])]=NA
height.groups[height.groups<(maxh*.7)]=1
height.groups[height.groups>(maxh*.7)]=200
height.groups[is.na(height.groups)]=-1
height.poly=as.polygons(height.groups)
viable=makeValid(height.poly[2,])
sites=vect("data_original/sitesinitialposition/")
ts=10

print(paste0("Starting simulation ","nan"))

onesimu=run_simulation (
                           sites=sites,
                           viable=viable,
                           dem=height.ras,
                           ressources=rast("data_original/resources.tiff"),
                           water=height.wat,
                           foldervid="nan",
                           visu=F,visumin=F,
                           ts=ts,#length of simulation in year
                           Kbase=c("HG"=35,"F"=110),#difference in K for the two cultures
                           cul_ext=c("HG"=7,"F"=6),#spatial penality to extent: lower, bigger penality
                           penal_cul=c("HG"=4,"F"=5),#penality of occupational area: low, other sites can cam close
                           prob_birth=c("HG"=0.3,"F"=0.5),#proba to give birth every year
                           prob_survive=c("HG"=0.8,"F"=0.65),#proba to die when pop > K
                           prob_split=c("HG"=.5,"F"=.6),#proba to create new settlement when Ne > K
                           minimals=c("HG"=.14,"F"=.20),#how big the group of migrant should be to create a new city vs migrate to a existing one 
                           bufferatack=300,#distance max around which settlement can fight
                           prob_move=c("HG"=0.2,"F"=0.1) #proba to migrate to existing settlement when Ne > K
                           )
```

The list `onesimu`  contains a few interesting info:


```{r,cache=F}
Nts <- onesimu$Nts # population at each timestep
warcasualties <- onesimu$warcasualties #death by war at each time step
```


Use ressource to adjust the Ks:
```{r,eval=T}
plot(sites,cex=(1+Nts[1,]/100),pch=21,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
```


We look at the simulation step by stepk
Step by step to allow interaction between sites:

```{r,eval=T,messages=F}
## Run stochastic process  


#par(mfrow=c(2,1))
i=ncol(Nts)
plot(1,1,type="n",xlim=c(0,i),ylim=c(0,max(Nts)),xlab="time",ylab="popsize")
lapply(1:ncol(Nts),function(j)lines(Nts[,j],col=rainbow(2)[as.factor(sites$culture)[j]]))

```

Visualise, Record deposit and  loss. for the 5 first sites
```{r,out.width="25%"}
alldeposit=lapply(1:5,function(i)Rec_c(sapply(Nts[,i],A_rates), InitBP = 15000,ts=ts,r = 0.2, Max_bone_thickness = "m"))
maxy=max(sapply(alldeposit,function(i)max(apply(i,2,sum))))

nill=lapply(alldeposit,function(depo)barplot(t(depo),col=viridis(ts+1),ylim=c(0,maxy)))
```

Population by time and culture

```{r,eval=T}
par(mfrow=c(2,1))
plot(1,1,type="n",xlim=c(0,i),ylim=c(0,max(Nts)),xlab="time",ylab="popsize")
lapply(1:ncol(Nts),function(i)lines(Nts[,i],col=rainbow(2)[as.factor(sites$culture)[i]]))
plot(apply(Nts,1,sum)[1:i],xlab="time",ylab="popsize")

```


Below some usefull code to manually select the ressource origins on the map (by clicking on the screen) and explore different decay around them

```{r,eval=F}
plotMap(height.ras,height.wat,paste0("year ",0))
new_site=vect(as.data.frame(locator()),geom=c("x","y"))  # select point on the sscreep
crs(new_site)=crs(height.ras)
d2=logisticdecay(new_site,height.ras,x=20000) # generate a logistic decay around the point

plot(d2)
w=(.7*d2+.3*ress)/(.7*minmax(d2)[2] + .3*minmax(ress)[2])
plot(w)
plot(mask(w*logisticdecay(new_site,height.ras,k=0.00002,x=1),territory))
multisamp=sapply(1:1000,function(i)spatSample(x=w,size=1,method="weights",xy=T)[1:2])


```

Exploring some trajectories of the simulations

```{r,eval=T}
i=nrow(Nts)
plot(apply(Nts[1:i,sites$culture=="F"],1,sum),col="red",type="l",lwd=2,ylim=c(0,max(apply(Nts,1,sum))))
 points(apply(Nts[1:i,sites$culture=="HG"],1,sum),col="blue",lwd=2,type="l")
plot(warcasualties[1:(i-1)],lwd=2,col="green",type="h",yaxt="n",ylab="")
axis(4)
par(new=T)
growF=apply(Nts[1:(i-1),sites$culture=="F"],1,sum)
growHG=apply(Nts[1:(i-1),sites$culture=="HG"],1,sum)
plot(growF,col="red",type="l",lwd=2,ylim=c(0,max(growF,growHG)))
 points(growHG,col="blue",lwd=2,type="l")
```

Population density:

overal: .7 

or: >1 and 1.2 for Farmers
or: <1  for Hunter Gather

HG: 0.08 - 0.16


