# Main Simulation {#main-simu}

## Initial Parameters
We now have sites with two cultures, a environment and function to make population growth, migrate and engages into wars. Each function depends on a lot of paramaters, some have been reviewd before other are explain in the help of each function. Often the parameters can differs between the two cultures. He we gave a list of parameters, close to those used during the Archaeoriddle challenge: 

## Preparation

We will create antoher older `data_tmp` where we will store useful file, that won't be shared to everyone but will be used at some point or another. 

```{r foldcreate}
foldtmp="data_tmp" #We will store all data that can/will be shared with participant in data_toshare
if(!file.exists(foldtmp))dir.create(foldtmp)
```

We setup all paramaters

```{r initparam}

#initialisation

Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures use and defined before ; will be use in simulation to define new site

# spatial penality for cultural extentions, ie for population to move out of initial site : lower, bigger penality
cul_ext <- c("HG"=7, "F"=6)

# penality of occupational area ie how hard it is to come close to your site ; if low, other sites can come close
penal_cul <- c("HG"=4, "F"=5)
# proba to give birth every year
prob_birth <- c("HG"=0.3, "F"=0.5)
# proba to die when pop > K
prob_survive <- c("HG"=0.8, "F"=0.6)
# proba to create new settlement when Ne > K
prob_split <- c("HG"=0.2, "F"=0.6)
# how big the group of migrant should be to create a new city vs
# migrate to a existing one
minimals <- c("HG"=0.14, "F"=0.20)
# prob to migrate to existing settlement when Ne > K
prob_move <- c("HG"=0.2,"F"=0.1)
```

Let's see our raster map again, and put the sites on top, with their ids and showing there initial population size:
```{r mapagain}
plotMap(height.ras,height.wat,paste0("year ", 0))
plot(sites, pch=21, add=T, bg=rainbow(2, alpha=0.6)[as.factor(sites$culture)],cex=(1+Nts[1,]/100))
text(sites,pos=3)
```

The `run_simulation` function will take the raster data and parameters for defined before for the two different cultures and run for the specified number of years `ts` (`r ts`).

It will then call the functions described above to model contacts between sites, the outcomes of fights and migrations. If you want 

```{r, eval=F}
?run_simulation
?whotouch
?fightbetterloss
?changePopSize
?warpoints
```


## Algorithm

The general organisation of the algorithm is as follow:

> initialisation
> for all sites s:
>     N_s <- growth(N_s)
>     if(N_s>K_s):
>         moving people:
>         if( N_s - K-S > min_s * K_s and probsplit_s):
>             create new city:
>                 find the closest most attractive spot and move there
>         else if(proba move ):
>             migration
>     war(s)


```{r algo-general, echo=TRUE,eval=F}

1. ** initialisation **
  
2. for (i in 2:(ts+1))
    inactives <- (Nts[i-1,]==0)
    for ( s in sample(seq_along(sites)[!inactives]) : #among all active site
        N_s <- Gpd()
        if $N_s> K_s$ :
            migrants <- newN - round(Ks[[s]]*0.9)
            if (migrants >= (minimals[sites$culture[s]]*sites$Ks[s]) & runif(1)<prob_split[sites$culture[s]] ){
          #if subpopulation > 10 people, 10% chance of creation of a new city

          infarea <- (sqrt(tmp)+penal_cul[cultures]) * buffersettl
          buffersize <- rnorm(length(infarea), infarea, infarea * 0.1)
          buffersize[tmp==0] <- 0
          territory <- erase(viable, buffer(sites, buffersize))
          
          if( length(territory)>0 ){
            ##select a new site given its distance to the old one and the ressourcesource available in ressources
            d2 <- logisticdecay( sites[s], dem, x=20000*cul_ext[sites$culture[s]])
            w <- (0.7 * d2 + 0.3*ressources) / (0.7*minmax(d2)[2] + 0.3*minmax(ressources)[2])
            new_site <- spatSample(
              x=mask(
                w * logisticdecay(sites[s], dem, k=0.00002,
                                  x=20000*cul_ext[sites$culture[s]]),
                territory),
              size=1, method="weights", xy=T)[1:2]
            new_site <- vect(new_site, geom=c("x","y"))
            
            if ( length(new_site)>0 & all(!is.na(crds(new_site))) ){
              ##add new site to site listes
              Ips[[length(Ips)+1]] <- initpopstruc(n=migrants) #initialise a fake populaition, will be updated by real migrants later
              new_site$culture <- sites$culture[s]
              new_site$Ks <- round(initKs(
                Kbase, sites=new_site, ressources,
                sizeex="F", rate=0.45))
              sites <- rbind(sites, new_site)
              Ks[length(Ks)+1] <- new_site$Ks
            }
          }
        }
        ## if no creation of new city happen, there is a certain probability that people will move
        if( length(new_site)==0 && runif(1) < prob_move[sites$culture[s]] ){
           #migration to existing site
          att <- extract(ressources,sites)[,2]
          space <- sites$Ks - (Nts[i-1,] + migrants)
          dis <- extract(logisticdecay(sites[s], dem, k=0.00002, x=1), sites)[,2]
          attractivity <- att * space * dis
          #attractivity=attractivity*(1+10*(sites$culture[s]==sites$culture)) #4 times more likely to go to similar culture
          attractivity[s] <- min(attractivity)-1
          attractivity <- exp(attractivity)/sum(exp(attractivity))
          attractivity[Nts[i-1,]<10] <- 0 
          attractivity[sites$culture!=sites$culture[s]] <- 0 
          if(any(is.na(attractivity))){
            print(attractivity)
            attractivity[is.na(attractivity)] <- 0
          }
          
          city <- sample(size=1, x=seq_along(sites), prob=attractivity)
          Nts[i,city] <- Nts[i-1,city] + migrants
        }
        if( havemoved ){
          Ips[c(s,city)] <- changePopSize( loosingPop=Ips[[s]], winingPop=Ips[[city]], size=migrants)
          newN <- newN - migrants
        }
        
      }
      Nts[i,s] <- newN
      
    }
    ## WAR =======================
    potentialfighters <- which(sites$culture=="F" & Nts[i,]>50)
    for (s in sample(x=potentialfighters, size=round(length(potentialfighters)*0.1))){
      buff <- bufferatack
      potentialvictims <- which(sites$culture !=sites$culture[s] & Nts[i,]>0) 
      clash <- whotouch(s, sites, Ne=Nts[i,], buffersize=buff)
      if(length(clash)>0 && !is.na(clash)){
        if(length(clash) == 1){
          attack <- clash
        } else {
          attack <- sample(clash, 1)
        }
        newns <- fightbetterloss(Ne=Nts[i,], a=s, b=attack)
        casualties <- sum(Nts[i, c(s,attack)] - newns[c(s,attack)])
        warcasualties[i] <- casualties
        sizew <- casualties^2/4000
        warpoints(sites, s, attack, Ne=Nts[i,],
                  buffersize=buff, sizewar=sizew+0.5)
        
        #effectively kill people in population (should be done taking into account age pyramid to be more realistic)
        Ips[[s]] <- changePopSize(loosingPop=Ips[[s]],
                                  size=(Nts[i,s] - newns[s]))
        Ips[[attack]] <- changePopSize(loosingPop=Ips[[attack]],
                                       size=(Nts[i, attack] - newns[attack]))
        Nts[i,] <- newns
      }
  }
  return(list(Nts=Nts, warcasualties=warcasualties, Ips=Ips, sites=sites))
}
```

We wrapped all this in a function, `run_simulation` that return a list of object. Let see one simple example:

```{r single-run-ex1,animation.hook="gifski", interval=0.2, eval=T,fig.cap="quick animation showing the simulation run in the previous chunk. Size o the dot represent the size of the settlement, the color its culture, where red represents farmers and blue hunter-gatherer. Flame represent fight between settlement, the size of the flame being proportional to the total number of casualties during the conflict (both sides included).",messages=F,message=F}
ts <- 150 # we keep using 250 time steps.
onesimu <- run_simulation(
  sites=sites, viable=viable, dem=height.ras,
  ressources=ress,
  water=height.wat,
  foldervid=NULL,
  visu=F, visumin=T,
  log=F,
  ts=ts, #length of simulation in year
  Kbase=c("HG"=35, "F"=110), #difference in K for the two cultures
  cul_ext=c("HG"=7, "F"=6), #spatial penality to extent: lower, bigger penality
  penal_cul=c("HG"=4, "F"=5), #penality of occupational area: low, other sites can cam close
  prob_birth=c("HG"=0.4, "F"=0.75), #proba of giving birth every year
  prob_survive=c("HG"=0.8, "F"=0.65), #proba of dying when pop > K
  prob_split=c("HG"=0.5, "F"=0.6), #proba of creating a new settlement when Ne > K
  minimals=c("HG"=0.14,"F"=0.20), #how big the group of migrant should be to create a new city vs migrate to a existing one 
  probfight=0.5, #this has been added for the bookdown to increase conflict occurences ; it is of 0.1 in the original challenge, it allows to modulate the probability that any settlement in a position of conflict actually goes to war.
  bufferatack=1000, #distance max around which settlement can fight 
  prob_move=c("HG"=0.2, "F"=0.1) #proba of migrating to existing settlement when Ne > K
)
```

Note that we here use a larger bufferatack that generates more conflicts and higher birth rates to make this simulation 

To create your own video of the simulation you can replace `foldervid=NULL` by `foldervid=foldtmp` or any other folder you want. this will save all output in the folder you specify, and you can use the image to generate your video using, for exemple, `ffmpeg`.


We also save the different object generated by `run_simulation` to be able to re-load it, share it, analyse it, later. 

```{r write-main-simu}
#write the terra SpatVect that stores all site sites location
terra::writeVector(onesimu$sites,file.path(foldtmp,"allsites.shp"),overwrite=TRUE)
#write the matrix with all the population for all istes and all time step 
saveRDS(file=file.path(foldtmp,"popSizeMatrix.RDS"),onesimu$Nts)
#write a list that stores  all the population structure for the sites at the end of the simulation
saveRDS(file=file.path(foldtmp,"popStructList.RDS"),onesimu$Ips)

#write a vector that stores  the number of death due to conflict at each time step
saveRDS(file=file.path(foldtmp,"wardeath.RDS"),onesimu$warcasualties)
```


### Source code {-}

Check the code of the simulation functions here:
<details><summary>Show code</summary>
**Code**
```{r print-run_sim, echo=FALSE, results='asis'}
a = knitr::spin('../../R/run_simulation.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/run_simulation.Rmd', '.')
res = knitr::knit_child(input='run_simulation.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('run_simulation.Rmd')
```
</details>


## Explore simulation:

The list `onesimu`  contains a few interesting info:


```{r explore-ex1-1}
Nts <- onesimu$Nts # population at each timestep
warcasualties <- onesimu$warcasualties #death by war at each time step
```
Nts is a $n \times m$ matrix where n is the number of time step of the simulaiton, and m the total number of sites that appears during the simulation, including the one that diseapers.

We can thus already see a couple of thing: `ncol(Nts)=` `r ncol(Nts)` 
place where occupied, the biggest settlement was the settlement `which(max(Nts)==Nts,ind.array=T)`

```{r exploreNts}
indmax=which(max(Nts)==Nts,arr.ind = T)
indmax[1] #timestep when biggest settlement was the biggest
indmax[2] #id of the biggest settlement 
max(Nts) #max number of people during the simulation
```



We can compare the initial condition with the  ressource to adjust the Ks:
```{r explore-ex1-b,out.width="45%",fig.cap="left: sites at initialisation, right: sites at the end of the simulation",fig.show="hold"}
plot(onesimu$site, cex=(Nts[1,]/max(Nts))*10, pch=21, bg=rainbow(2, alpha=0.6)[as.factor(onesimu$site$culture)])
plot(onesimu$site, cex=(Nts[nrow(Nts),]/max(Nts))*10, pch=21, bg=rainbow(2, alpha=0.6)[as.factor(onesimu$site$culture)])
```

We can look at the dynamics throught time, here how each individual sites are growings:

```{r explore-ex1-trajpersite,eval=T,messages=F}
## visualise trajectories for each site
i <- ts
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)), xlab="time", ylab="popsize")
nill <- lapply(1:ncol(Nts),
       function(j)lines(Nts[,j],col=rainbow(2)[as.factor(sites$culture)[j]])
)
```

Population by time and culture

```{r explore-ex1-timeandcult,out.width="45%",fig.cap="",fig.show="hold"}
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)),
     xlab="time", ylab="popsize")
na=lapply(1:ncol(Nts),
       function(i)lines(Nts[,i], col=rainbow(2)[as.factor(sites$culture)[i]]))
legend("topright",legend=c("F","HG"),col= rainbow(2),lwd=1,title="culture")
plot(apply(Nts, 1, sum)[1:i], xlab="time", ylab="popsize")
```

Exploring some trajectories of the simulations

```{r, eval=T}
i <- nrow(Nts)  # Get the number of rows in Nts
plot(apply(Nts[1:i, sites$culture=="F",drop=F], 1, sum), col="red", type="l",
     lwd=2, ylim=c(0, max(apply(Nts, 1, sum))))  # Plot sum of 'F' culture values
points(apply(Nts[1:i, sites$culture=="HG"], 1, sum),
       col="blue", lwd=2, type="l")  # Add points for 'HG' culture values
plot(warcasualties[1:(i-1)], lwd=2, col="green", type="h", yaxt="n", ylab="")  # Plot war casualties
axis(4)  # Add an axis on the right side
par(new=T)  # Allow a new plot to be drawn on the existing plot
growF <- apply(Nts[1:(i-1), sites$culture=="F",drop=F], 1, sum)  # Sum of 'F' culture values excluding last row
growHG <- apply(Nts[1:(i-1), sites$culture=="HG",drop=F], 1, sum)  # Sum of 'HG' culture values excluding last row
plot(growF, col="red", type="l", lwd=2, ylim=c(0, max(growF, growHG)))  # Plot growth of 'F' culture
points(growHG, col="blue", lwd=2, type="l")  # Add points for growth of 'HG' culture
```

