# Main Simulation {#main-simu}

## Initial Parameters
We now have sites with two cultures, an environment and a function to make population growth, migrate and engages into wars. Each function depends on a lot of parameters, most of which have been reviewed and are explained in the `help` of each function. Often, the parameters can differ between the two cultures. Here we give a list of parameters close to those used during the Archaeoriddle challenge: 

## Preparation

We will create another older `data_tmp` where we will store valuable files that will not be shared with everyone but will be used at some point. 

```{r foldcreate}
foldtmp="data_tmp" #We will store all data that can/will be shared with participant in data_toshare
if(!file.exists(foldtmp))dir.create(foldtmp)
```

We setup all parameters

```{r initparam}

#initialisation

Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures use and defined before ; will be use in simulation to define new site

# spatial penality for cultural extentions, ie for population to move out of initial site : lower, bigger penality
cul_ext <- c("HG"=7, "F"=6)

# penality of occupational area ie how hard it is to come close to your site ; if low, other sites can come close
penal_cul <- c("HG"=4, "F"=5)
# proba to give birth every year
prob_birth <- c("HG"=0.3, "F"=0.5)
# proba to die when pop > K
prob_survive <- c("HG"=0.8, "F"=0.6)
# proba to create new settlement when Ne > K
prob_split <- c("HG"=0.2, "F"=0.6)
# how big the group of migrant should be to create a new city vs
# migrate to a existing one
minimals <- c("HG"=0.14, "F"=0.20)
# prob to migrate to existing settlement when Ne > K
prob_move <- c("HG"=0.2,"F"=0.1)
```

We see our raster map again, and put the sites on top, with their IDs and showing their initial population size:
```{r mapagain}
plotMap(height.ras,height.wat,paste0("year ", 0))
plot(sites, pch=21, add=T, bg=rainbow(2, alpha=0.6)[as.factor(sites$culture)],cex=(1+Nts[1,]/100))
text(sites,pos=3)
```

The `run_simulation` function will take the raster data and the parameters defined before for the two different cultures and run for the specified number of years `ts` (`r ts`).

It will then call the functions described above to model contacts between sites, the outcomes of fights and migrations. If you want 

```{r, eval=F}
?run_simulation
?whotouch
?fightbetterloss
?changePopSize
?warpoints
```


## Algorithm

The general organisation of the algorithm is as follows:

> initialisation
> for all sites s:
>     N_s <- growth(N_s)
>     if(N_s>K_s):
>         moving people:
>         if( N_s - K-S > min_s * K_s and probsplit_s):
>             create new city:
>                 find the closest most attractive spot and move there
>         else if(proba move ):
>             migration
>     war(s)


```{r algo-general, echo=TRUE,eval=F}

1. ** initialisation **
  
2. for (i in 2:(ts+1))
    inactives <- (Nts[i-1,]==0)
    for ( s in sample(seq_along(sites)[!inactives]) : #among all active site
        N_s <- Gpd()
        if $N_s> K_s$ :
            migrants <- newN - round(Ks[[s]]*0.9)
            if (migrants >= (minimals[sites$culture[s]]*sites$Ks[s]) & runif(1)<prob_split[sites$culture[s]] ){
          #if subpopulation > 10 people, 10% chance of creation of a new city

          infarea <- (sqrt(tmp)+penal_cul[cultures]) * buffersettl
          buffersize <- rnorm(length(infarea), infarea, infarea * 0.1)
          buffersize[tmp==0] <- 0
          territory <- erase(viable, buffer(sites, buffersize))
          
          if( length(territory)>0 ){
            ##select a new site given its distance to the old one and the ressourcesource available in ressources
            d2 <- logisticdecay( sites[s], dem, x=20000*cul_ext[sites$culture[s]])
            w <- (0.7 * d2 + 0.3*ressources) / (0.7*minmax(d2)[2] + 0.3*minmax(ressources)[2])
            new_site <- spatSample(
              x=mask(
                w * logisticdecay(sites[s], dem, k=0.00002,
                                  x=20000*cul_ext[sites$culture[s]]),
                territory),
              size=1, method="weights", xy=T)[1:2]
            new_site <- vect(new_site, geom=c("x","y"))
            
            if ( length(new_site)>0 & all(!is.na(crds(new_site))) ){
              ##add new site to site listes
              Ips[[length(Ips)+1]] <- initpopstruc(n=migrants) #initialise a fake populaition, will be updated by real migrants later
              new_site$culture <- sites$culture[s]
              new_site$Ks <- round(initKs(
                Kbase, sites=new_site, ressources,
                sizeex="F", rate=0.45))
              sites <- rbind(sites, new_site)
              Ks[length(Ks)+1] <- new_site$Ks
            }
          }
        }
        ## if no creation of new city happen, there is a certain probability that people will move
        if( length(new_site)==0 && runif(1) < prob_move[sites$culture[s]] ){
           #migration to existing site
          att <- extract(ressources,sites)[,2]
          space <- sites$Ks - (Nts[i-1,] + migrants)
          dis <- extract(logisticdecay(sites[s], dem, k=0.00002, x=1), sites)[,2]
          attractivity <- att * space * dis
          #attractivity=attractivity*(1+10*(sites$culture[s]==sites$culture)) #4 times more likely to go to similar culture
          attractivity[s] <- min(attractivity)-1
          attractivity <- exp(attractivity)/sum(exp(attractivity))
          attractivity[Nts[i-1,]<10] <- 0 
          attractivity[sites$culture!=sites$culture[s]] <- 0 
          if(any(is.na(attractivity))){
            print(attractivity)
            attractivity[is.na(attractivity)] <- 0
          }
          
          city <- sample(size=1, x=seq_along(sites), prob=attractivity)
          Nts[i,city] <- Nts[i-1,city] + migrants
        }
        if( havemoved ){
          Ips[c(s,city)] <- changePopSize( loosingPop=Ips[[s]], winingPop=Ips[[city]], size=migrants)
          newN <- newN - migrants
        }
        
      }
      Nts[i,s] <- newN
      
    }
    ## WAR =======================
    potentialfighters <- which(sites$culture=="F" & Nts[i,]>50)
    for (s in sample(x=potentialfighters, size=round(length(potentialfighters)*0.1))){
      buff <- bufferatack
      potentialvictims <- which(sites$culture !=sites$culture[s] & Nts[i,]>0) 
      clash <- whotouch(s, sites, Ne=Nts[i,], buffersize=buff)
      if(length(clash)>0 && !is.na(clash)){
        if(length(clash) == 1){
          attack <- clash
        } else {
          attack <- sample(clash, 1)
        }
        newns <- fightbetterloss(Ne=Nts[i,], a=s, b=attack)
        casualties <- sum(Nts[i, c(s,attack)] - newns[c(s,attack)])
        warcasualties[i] <- casualties
        sizew <- casualties^2/4000
        warpoints(sites, s, attack, Ne=Nts[i,],
                  buffersize=buff, sizewar=sizew+0.5)
        
        #effectively kill people in population (should be done taking into account age pyramid to be more realistic)
        Ips[[s]] <- changePopSize(loosingPop=Ips[[s]],
                                  size=(Nts[i,s] - newns[s]))
        Ips[[attack]] <- changePopSize(loosingPop=Ips[[attack]],
                                       size=(Nts[i, attack] - newns[attack]))
        Nts[i,] <- newns
      }
  }
  return(list(Nts=Nts, warcasualties=warcasualties, Ips=Ips, sites=sites))
}
```

We wrapped all this in a function, `run_simulation` which returns an object of class `list()`. Let's see one simple example:

```{r single-run-ex1,animation.hook="gifski", interval=0.2, eval=T,fig.cap="quick animation showing the simulation run in the previous chunk. Sizes of the dots represent the size of the settlement, the color its culture, where red represents farmers and blue hunter-gatherers. Flame represents fight between settlements, the size of the flame being proportional to the total number of casualties during the conflict (both sides included).",messages=F,message=F}
ts <- 150 # we keep using 250 time steps.
onesimu <- run_simulation(
  sites=sites, viable=viable, dem=height.ras,
  ressources=ress,
  water=height.wat,
  foldervid=NULL,
  visu=F, visumin=T,
  log=F,
  ts=ts, #length of simulation in year
  Kbase=c("HG"=35, "F"=110), #difference in K for the two cultures
  cul_ext=c("HG"=7, "F"=6), #spatial penality to extent: lower, bigger penality
  penal_cul=c("HG"=4, "F"=5), #penality of occupational area: low, other sites can cam close
  prob_birth=c("HG"=0.4, "F"=0.75), #proba of giving birth every year
  prob_survive=c("HG"=0.8, "F"=0.65), #proba of dying when pop > K
  prob_split=c("HG"=0.5, "F"=0.6), #proba of creating a new settlement when Ne > K
  minimals=c("HG"=0.14,"F"=0.20), #how big the group of migrant should be to create a new city vs migrate to a existing one 
  probfight=0.5, #this has been added for the bookdown to increase conflict occurences ; it is of 0.1 in the original challenge, it allows to modulate the probability that any settlement in a position of conflict actually goes to war.
  bufferatack=1000, #distance max around which settlement can fight 
  prob_move=c("HG"=0.2, "F"=0.1) #proba of migrating to existing settlement when Ne > K
)
```

Note that we here use a larger bufferatack that generates more conflicts and higher birth rates to make this simulation 

To create your own video of the simulation you can replace `foldervid=NULL` by `foldervid=foldtmp` or any other folder you want. This will save all outputs in the folder you specify, and you can use the image to generate your video using, for example, `ffmpeg`.


We also save the different objects generated by `run_simulation` to be able to re-load them, share them, and analyse them later. 

```{r write-main-simu}
#write the terra SpatVect that stores all site sites location
terra::writeVector(onesimu$sites,file.path(foldtmp,"allsites.shp"),overwrite=TRUE)
#write the matrix with all the population for all sites and all time steps
saveRDS(file=file.path(foldtmp,"popSizeMatrix.RDS"),onesimu$Nts)
#write a list that stores  all the population structure for the sites at the end of the simulation
saveRDS(file=file.path(foldtmp,"popStructList.RDS"),onesimu$Ips)
#write a vector that stores  the number of death due to conflict at each time step
saveRDS(file=file.path(foldtmp,"wardeath.RDS"),onesimu$warcasualties)
```


### Source code {-}

Check the code of the simulation functions here:
<details><summary>Show code</summary>
**Code**
```{r print-run_sim, echo=FALSE, results='asis'}
a = knitr::spin('../../R/run_simulation.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/run_simulation.Rmd', '.')
res = knitr::knit_child(input='run_simulation.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('run_simulation.Rmd')
```
</details>


## Explore simulation

The list `onesimu`  contains a few interesting info:

```{r explore-ex1-1}
Nts <- onesimu$Nts # population at each timestep
warcasualties <- onesimu$warcasualties #death by war at each time step
```
Nts is a $n \times m$ matrix where n is the number of time steps of the simulation, and m the total number of sites that appears during the simulation, including the ones that disappear.

With this `Nts` matrix we can already see a couple of things: `ncol(Nts)=` `r ncol(Nts)`, the total number of place occupied, and others (next chunk):

```{r exploreNts}
indmax=which(max(Nts)==Nts,arr.ind = T)
indmax[1] #timestep when biggest settlement was the biggest
indmax[2] #id of the biggest settlement 
max(Nts) #max number of people during the simulation
```

We can also look at the dynamics through time, and explore how each individual site is growing:

```{r explore-ex1-trajpersite,fig.show="hold",out.width="45%",fig.cap="left: Growth of each site, right: overall growth"}
i <- ts
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)), xlab="time", ylab="popsize")
nill <- lapply(1:ncol(Nts),function(j)lines(Nts[,j]))
plot(apply(Nts, 1, sum)[1:i], xlab="time", ylab="popsize")
```

Combining `onesimu$site` with `Nts` we can compare the initial conditions with the state of the world at the end:

```{r explore-ex1-b,out.width="45%",fig.cap="left: sites at initialisation, right: sites at the end of the simulation",fig.show="hold"}
plot(onesimu$site, cex=(Nts[1,]/max(Nts))*10, pch=21, bg=rainbow(2, alpha=0.6)[as.factor(onesimu$site$culture)])
plot(onesimu$site, cex=(Nts[nrow(Nts),]/max(Nts))*10, pch=21, bg=rainbow(2, alpha=0.6)[as.factor(onesimu$site$culture)])
```

Or look at the dynamics through time but divided by culture, and comparing with the war casualties

```{r explore-ex1-timeandcult,out.width="",fig.cap="left: Growth of each site for both cultures, right: combined growth",fig.show="hold",out.width="45%"}
plot(warcasualties[1:(i-1)], lwd=2, col="green", type="h", yaxt="n", ylab="")  # Plot war casualties
par(new=T) 
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)), xlab="time", ylab="popsize")
na=lapply(1:ncol(Nts),
       function(i)lines(Nts[,i], col=c("red","blue")[as.factor(sites$culture)[i]]))
plot(warcasualties[1:(i-1)], lwd=2, col="green", type="h", yaxt="n", ylab="")  # Plot war casualties
par(new=T)
growF <- apply(Nts[1:(i-1), sites$culture=="F",drop=F], 1, sum)  # Sum of 'F' culture values excluding last row
growHG <- apply(Nts[1:(i-1), sites$culture=="HG",drop=F], 1, sum)  # Sum of 'HG' culture values excluding last row
plot(growF, col="red", type="l", lwd=2, ylim=c(0, max(growF, growHG)))  # Plot growth of 'F' culture
points(growHG, col="blue", lwd=2, type="l")  # Add points for growth of 'HG' culture
legend("topleft",legend=c("F","HG","war"),col=c("red","blue","green"),lwd=1,title="culture")
```


The dynamics of the simulations will heavily depend on the initial conditions and the parameters. As an example, with these parameters, it is not rare that the farmers, due to the conflict mechanisms that we described in Chapter&nbsp;\@ref(conflict), will destroy themselves by going to war too often, as one can see in the Figure&nbsp;\@ref(fig:explore-mult) below. This can be avoided with the current implementation by setting a small `probfight` value to 0.1; however, it is important to remember that this could have been achieved in various ways. A more clever approach would have been for farmers to 'estimate' the likelihood of them losing the battle, perhaps by fighting only _safe_ battles, against much smaller settlements.

In order to find the right simulation, the one that will show the dynamics you want your participants to detect, you will need to play with these parameters and run several replications of the simulation.


The script `scriptmini.R` was designed to do so for the original challenge, here we adapted it for a more generic need, to re-run simulation using the world you have generated throughout this book. 

```{r multi-run,out.width="33%",fig.cap="Final state of multiple simulations with the exact same parameters.",fig.show="hold"}

par(mar=c(0,0,0,0))
height.ras=rast("data_toshare/dem_raster.tiff")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA
height.groups=height.ras
maxh=max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])]=NA
height.groups[height.groups<(maxh*.7)]=1
height.groups[height.groups>(maxh*.7)]=200
height.groups[is.na(height.groups)]=-1
height.poly=as.polygons(height.groups)
viable=makeValid(height.poly[2,])
sites=vect("data_tmp/allsites.shp")
ts=10

allsim=list()  #we will store te multiplesimulation, in this list to play with it, with a proper setup we stored results in files as simulation are too big to be kept in memory.
for(expname in paste0("exp",1:9)){

    print(paste0("Starting simulation ",expname))

    onesimu=run_simulation (
                            sites=sites, viable=viable, dem=height.ras,
                            ressources=ress,
                            water=height.wat,
                            foldervid=NULL,
                            visu=F, visumin=F,
                            log=F,
                            ts=ts, 
                            Kbase=c("HG"=35, "F"=110), 
                            cul_ext=c("HG"=7, "F"=6), 
                            penal_cul=c("HG"=4, "F"=5), 
                            prob_birth=c("HG"=0.4, "F"=0.75), 
                            prob_survive=c("HG"=0.8, "F"=0.65), 
                            prob_split=c("HG"=0.5, "F"=0.6), 
                            minimals=c("HG"=0.14,"F"=0.20), 
                            probfight=0.5, 
                            bufferatack=1000, 
                            prob_move=c("HG"=0.2, "F"=0.1) 
    )

    Nts=onesimu$Nts
    warcasualties=onesimu$warcasualties
    sites=onesimu$sites
    i=min(ts,which(apply(Nts,1,sum)==0))
    saveRDS(file=file.path(foldtmp,paste0(expname,"_all.RDS")),onesimu)
    saveRDS(file=file.path(foldtmp,paste0(expname,"_sitesRast.RDS")),onesimu$sites)
    plotMap(height.ras,height.wat,paste0("year ",i))
    plot(sites,cex=(as.integer(Nts[i,]>0)*0.3+Nts[i,]/200),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
    growF=apply(Nts[1:i,sites$culture=="F"],1,sum)
    growHG=apply(Nts[1:i,sites$culture=="HG"],1,sum)
    growT=apply(Nts[1:i,],1,sum)

    allsim[[expname]]=cbind.data.frame(growF,growHG,growT,warcasualties[1:i],rep(expname,i))
}
```

```{r explore-mult,out.width="33%",fig.cap="Comparing multiple runs",fig.show="hold"}
    gnmax=max(sapply(allsim,function(i)max(i$growT)))
    plot(1,i,xlim=c(0,10),ylim=c(0,gnmax),type="n",main="Overal growth",ylab="number of individual",xlab="time")
    na=lapply(allsim,function(i)lines(i$growT))
    gnmax=max(sapply(allsim,function(i)max(i[,c("growF","growHG")])))
    plot(1,i,xlim=c(0,10),ylim=c(0,gnmax),type="n",main="Growth per culture",ylab="number of individual",xlab="time")
    na=lapply(allsim,function(i){lines(i$growHG,col="red");lines(i$growF,col="green")})
    gnmax=max(sapply(allsim,function(i)max(i$warcasualties)))
    plot(1,i,xlim=c(0,10),ylim=c(0,gnmax),type="n",main="War Casualties",ylab="number of individual",xlab="time")
    na=lapply(allsim,function(i)lines(i$warcasualties))
```



Once you find the right simulation, we can use the populations this simulation created to generate an archaeological record based on the mechanisms that we will describe in the following chapters.

