# Final Choice and simulation

As I (simon) am stupid I didn't generate seeds for each experiment so we cannot re-run it EXACTLY the same. But I kept lot of informations and output about it.

They are stored in the folder `general_results_selected_simu/`

and will allow us to reconstruct most of what happened. We basically run thousands of simulations and choose the one who looked the cool 😆. We used the script `scriptmini.R` that allow to run one simulation and ran in in parallel as described in `smallscript.md`.

What scriptmini.R doesn basically is

1. We first load the maps and the sites:

```{r}
library(terra)
library(sf)
source("tools.R")

height.ras=rast("east_narnia4x.tif")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA
height.groups=height.ras
maxh=max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])]=NA
height.groups[height.groups<(maxh*.7)]=1
height.groups[height.groups>(maxh*.7)]=200
height.groups[is.na(height.groups)]=-1
height.poly=as.polygons(height.groups)
viable=makeValid(height.poly[2,])
sites=vect(readRDS("testsites.RDS"))
```

2. run one simulation


```{r,eval=F}

ts=1000
print(paste0("Starting simulation ",expname))

onesimu=run_simulation (
                           sites=sites,
                           viable=viable,
                           dem=height.ras,
                           ressources=rast("ressources.tiff"),
                           water=height.wat,
                           foldervid=expname,
                           visu=F,visumin=TRUE,
                           ts=ts,#length of simulation in year
                           Kbase=c("HG"=35,"F"=110),#difference in K for the two cultures
                           cul_ext=c("HG"=7,"F"=6),#spatial penality to extent: lower, bigger penality
                           penal_cul=c("HG"=4,"F"=5),#penality of occupational area: low, other sites can cam close
                           prob_birth=c("HG"=0.3,"F"=0.5),#proba to give birth every year
                           prob_survive=c("HG"=0.8,"F"=0.65),#proba to die when pop > K
                           prob_split=c("HG"=.5,"F"=.6),#proba to create new settlement when Ne > K
                           minimals=c("HG"=.14,"F"=.20),#how big the group of migrant should be to create a new city vs migrate to a existing one 
                           bufferatack=300,#distance max around which settlement can fight
                           prob_move=c("HG"=0.2,"F"=0.1) #proba to migrate to existing settlement when Ne > K
                           )
```



2. generate a few output, including a video that helps to select the good simulation

```{r,eval=F}


Nts=onesimu$Nts
warcasualties=onesimu$warcasualties
sites=onesimu$sites
i=min(ts,which(apply(Nts,1,sum)==0))
pdf(paste0(expname,"_mapFinal.pdf"))
plotMap(height.ras,height.wat,paste0("year ",i))
plot(sites,cex=(as.integer(Nts[i,]>0)*0.3+Nts[i,]/200),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
text(sites)
dev.off()


pdf(paste0(expname,"growth_utils.pdf"))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="green",type="h",yaxt="n",ylab="",xlim=c(0,i))
axis(4)
par(new=T)
growF=apply(Nts[1:i,sites$culture=="F"],1,sum)
growHG=apply(Nts[1:i,sites$culture=="HG"],1,sum)
plot(growF,col="red",type="l",lwd=2,ylim=c(0,max(growF,growHG)),xlim=c(0,i))
points(growHG,col="blue",lwd=2,type="l")
dev.off()

pdf(paste0(expname,"growth_tot.pdf"))
plot(warcasualties[1:i-1],lwd=2,col="green",type="h",yaxt="n",ylab="")
axis(4)
par(new=T)
growT=apply(Nts[1:i,],1,sum)
plot(growT,col="black",type="l",lwd=2,ylim=c(0,max(growT)))
dev.off()
```

Let's redo that with the selected simulation:


```{r initialcondition}

onesimu=readRDS("general_results_selected_simu/buffattack300_K110_PSU065_3_all.RDS")
Nts=onesimu$Nts
warcasualties=onesimu$warcasualties
sites=terra::vect(readRDS("general_results_selected_simu/buffattack300_K110_PSU065_3_sitesRast.RDS"))

i=1
plotMap(height.ras,height.wat,paste0("year ",i,", when it all begins"))
plot(sites,cex=(as.integer(Nts[1,]>0)*0.5+Nts[1,]/100),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
text(sites[Nts[1,]>0],pos=3)

```

we can look at the war and the number of death for each battle

```{r}
i=1000
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",ylab="# dead",xlim=c(0,i))
```

I comparison to the total population growth of each culture

```{r}
i=1000
par(mar=c(5,5,1,5))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",yaxt="n",ylab="",xlim=c(0,i),xlab="time")
axis(4)
mtext("# dead",4,2.5)
par(new=T)
growF=apply(Nts[1:i,sites$culture=="F"],1,sum)
growHG=apply(Nts[1:i,sites$culture=="HG"],1,sum)
plot(growF,col="orange",type="l",lwd=5,ylim=c(0,max(growF,growHG)),xlim=c(0,i),ylab="pop",xlab="")
points(growHG,col="blue",lwd=5,type="l")
```

If we look at the total population size, we can see how the clashes slowed down the population growth:
```{r}
par(mar=c(5,5,1,5))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",yaxt="n",ylab="",xlim=c(0,i),xlab="time")
axis(4)
mtext("# dead",4,2.5)
par(new=T)
growT=apply(Nts,1,sum)
plot(growT,col="black",type="l",lwd=5,ylim=c(0,max(growT)),xlim=c(0,i),ylab="pop",xlab="")

# Function to calculate simple moving average
sma <- function(x, n) {
      s <- stats::filter(x, rep(1/n, n), sides=2)
  return(s)
}
```

This show if we looka at the growth rate:

```{r}
plot(diff(growT,lag=100),type="l")
```


Replaying all that will give this:

```{r,echo=False}
htmltools::tags$video(src="general_results_selected_simu/buffattack300_K110_PSU065_3_out.mp4")
```





