---
output:
  html_document: default
  pdf_document: default
  always_allow_html: true
  code_folding: show
---


# Record formation {#recgen}

## Theoretical model

We have produced a computational model, replicating the process of formation of the archaeological record in a specific site, accounting for different variables. In order to consider the possible relation between those variables, we have decided to express them as a Directed Acyclic Graph (DAG), following the ECS-DAG methodology proposed by Ferguson et al. (2020). In this case, however, and because our goal is to obtain an accurate initial template for the development of the model, we have not considered setting an exposure variable, but rather to propose different theoretical directional and causal relationships without relying on the conventional exposure-outcome assessment. Therefore, highlights in the graph below do not intend to differentiate among confounders, mediators, etc., but to emphasize the possible weight that each variable has for producing the outcome (produced waste by $m^2$).

```{r, echo = FALSE}
require(DiagrammeR)

grViz("
  digraph record_formation {
  
    graph [layout = dot, ## Set title and layout
           label = 'Record formation process',
           labelloc = 't'
           fontname = Helvetica,
           fontsize = 30]
    
    # node definitions with substituted label text
    node [fontname = Helvetica, ## General node definition
          style = filled,
          color = MistyRose]
          
    ## Per node definition
    a [label = '@@1',
       fillcolor = orange]
    b [label = '@@2',
       fillcolor = cadetblue1]
    c [label = '@@3',
       fillcolor = orange]
    d [label = '@@4',
       fillcolor = cadetblue1]
    e [label = '@@5',
       fillcolor = orange]
    f [label = '@@6',
       fillcolor = cadetblue,
       fixedsize = shape,
       width = 4.5,
       height = 1,
       fontsize = 30]
    
    ## General edge definition
    edge [color = grey,
           arrowhead = vee,
           arrowtail = vee]
    
    ## Specific edges
    a -> {c}
    b -> {c}
    c -> {f}
    d -> {e}
    e -> {f}
  }
  
  ## Label definition
  [1]: 'Pop (group size)'
  [2]: 'Site function'
  [3]: 'Anthropogenic deposition rates'
  [4]: 'Site type'
  [5]: 'Natural deposition rates'
  [6]: 'Waste/sqrm'

")

```

The above graph depicts a very schematic way to represent the formation of the archaeological record ($W/m^2$). In this case, we are interested in radiocarbon dating, focusing on bone samples; that is the number (and range) of potential 14C dates produced in a site, and this is key to understand the way in which we produced the model. We are aware that generating a specific archaeological palimpsest depends on a large quantity of variables, some of which may be controlled by field archaeologists and some of which may not (or are not). Therefore, the graph above must be read as follows. Asides from the outcome variable, the exposure variables, *Pop (group size)*, *Anthropogenic deposition rates* and *Natural deposition rates* (depicted in orange), are designed in a way that they can be activated for the model to infer them, they can be provided by the archaeologist, or they can be simply ignored. In the most elementary case, the archaeologist could introduce directly the outcome variable ($W/m^2$) without taking into account anything else, if she/he is confident to provide an accurate value.

However, in most cases, this information will not be known to the field researcher, and in this situation, these variables can be modeled to different extents of complexity. For example, in the case of the most simple simulation, the field archaeologist could ignore deposition rates. In this case, and because we are only considering 14C dates, then the quantity of the sample becomes broadly proportional to the number of people present in a site, and thus it could be modeled like this.

```{r, echo = FALSE, out.width="50%"}
grViz("
  digraph record_formation {
  
    graph [layout = dot, ## Set title and layout
           label = 'Record formation process',
           labelloc = 't'
           fontname = Helvetica,
           fontsize = 30]
    
    # node definitions with substituted label text
    node [fontname = Helvetica, ## General node definition
          style = filled,
          color = MistyRose]
           
    ## Per node definition
    a [label = '@@1',
       fillcolor = orange]
    b [label = '@@2',
       fillcolor = cadetblue,
       fixedsize = shape,
       width = 4.5,
       height = 1,
       fontsize = 30]
    
    ## General edge definition
    edge [color = grey,
           arrowhead = vee,
           arrowtail = vee]
    
    ## Specific edges
    a -> {b}
  }
  
  ## Label definition
  [1]: 'Pop (group size)'
  [2]: 'Waste/sqrm'

")

```

However, deposition rates are difficult to ignore if we want to account for the formation process of the archaeological record. If these are known to the archaeologists that developed fieldwork, then this would be the behaviour of the model for creating $W/m^2$.

```{r, echo = FALSE, out.width="60%"}
grViz("
  digraph record_formation {
  
    graph [layout = dot, ## Set title and layout
           label = 'Record formation process',
           labelloc = 't'
           fontname = Helvetica,
           fontsize = 30]
    
    # node definitions with substituted label text
    node [fontname = Helvetica, ## General node definition
          style = filled,
          color = MistyRose]
          
    ## Per node definition
    a [label = '@@1',
       fillcolor = orange]
    b [label = '@@2',
       fillcolor = orange]
    c [label = '@@3',
       fillcolor = cadetblue,
       fixedsize = shape,
       width = 4.5,
       height = 1,
       fontsize = 30]
    
    ## General edge definition
    edge [color = grey,
           arrowhead = vee,
           arrowtail = vee]
    
    ## Specific edges
    a -> {c}
    b -> {c}
  }
  
  ## Label definition
  [1]: 'Anthropogenic deposition rates'
  [2]: 'Natural deposition rates'
  [3]: 'Waste/sqrm'

")

```

Finally, if these are not known to the archaeologist, but they want to model them, then we would go back to the first DAG. In this case, keep reading for the specificities of the model.

This is how each variable works specifically:

-   **Waste produced** ($W/m^2$ or just $W$) **(outcome variable)**: Archaeological record present per $m^2$. Since we are considering C14 dates coming from bone, this is the quantity of C14 samples present a time *t* and at a depth *d*. Thus, this variable depends on time and depth.

$$W(t,d)=function(t,d)$$.

We have considered this a discrete variable (number of potential samples) with range $[0,\propto K]$. *Defined in code as* `W`.

The variables time ($T$) and depth ($D$) must also be defined. Thus

-   **T** describes the time, discretely, in years, of existence of the site. For the simulation, this time will be known. Thus, in the simulation $T_{birth}=(T=0)$ and $T_{max}=T_{death}$. Values of $T$ are represented by $t$.
-   **D** is the depth, in cms, where each specific sample is deposited at a time $t$. This is a continuous variable with range $[0,\infty]$, and for any $t$, we know that $0\leq D(t)\leq d_0$ and that

$$D(t)=max\{D(t-1)-r,0\} \tag{1},\\ t=1,2,...,T_{max},$$

where $r$ is the natural deposition rates.

-   **Population (group size)** ($P$): Discrete variable with range $[0,\sim K]$. Population group size is key to determine anthropogenic deposition rates (which in turn determine $W$) but it can also be used as a proxy to determine $W$ in case deposition rates are not accounted for since, in any case, $W \propto Pop$. The model includes the possibility to create a stochastic demographic process (see below), but the user can also set fix values for population or set the parameters in a way that the stochastic process is constrained. In any case, the user does have to provide a value for the carrying capacity ($K$) as this will limit population growth. *Defined in code as* `Pop`.
-   **Anthropogenic deposition** ($A$): Amount of bone kilogram/year deposited by each group at a time $t$. This is a continuous variable, with range $[0,\infty]$, measured in *Kg*. See *Anthropogenic deposition protocol* for full definition.
-   **Natural deposition rates** ($r$): This is a continuous variable, with range $[0,0.5]$, measured in *cm/year*. Because, under normal circumstances, bones are thicker than any possible amount of earth deposited per year, even in a perfectly uncorrupted archaeological sequence (which is what is being modelled here), those would occupy microstrats belonging to different years. This variable is one of the responsibles for how many microstrats will each bone be occupying. See *Depth protocol* to understand how it works. *Defined in code as* `r`.
-   **Site function**: Has not been considered as a variable *per se* but, rather, as a group of variables, which are defined by the functionality of the site, and that condition the outcome of $A$. See *Anthropogenic deposition protocol* to understand how these are modelled.
-   **Site type**: Site type (whether a site is a cave, a rockshelter or an open-air camp) can heavily condition the deposition rates at any archaeological location. However, in order to reduce uncertainty and not make an overcomplex model, we have decided to embed this condition, and the possible expert knowledge related to it, within the definition of the variable $r$.

## Implementation

This simulation for the creation of the archaeological record has been divided in three phases, or protocols.  

The first one has been described in the [previous chapter](#popgrowth), where we populated the site, according to a stochastic demographic process on top of which migration and war events occurred. Here we focus on how these populations generate waste, and how this waste will be stored in the archaeological record and where. This is done in two steps: we first generate a sample of waste for each time $t$ and, in the second step, we distribute the amount of waste for each depth $d$.

In order to fully understand the process, first we go through the protocols and functions used. 

### Population $Pop$-protocol

Before anything, in order to account for the chronological occupation of the site, a demographic process must be produced _ie_ sites need to be occupied. From these occupations a record sample is generated accordingly. This is what his done and described in the [previous chapter](#popgrowth), but here we will use the function `Pop_stoch`  to generate a simple stochastic growth and apply our loss function to it in order to describe and test the other protocols.

```{r, eval=FALSE}
?Pop_stoch
```
<br>

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/population.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/population.Rmd', '.')
```

### Anthropogenic deposition $A$-protocol

This generates the amount of human waste produced for each $t$ (In kilograms of bone). It is defined as:

$$A(t)= Pop(t)\times O(t)\times F(t) \times G(t),\\ \text{with}\ t = 1,2,...,T_{death} \tag{2}$$ Where:

-   $Pop(t)$ is a discrete variable representing the number of living people at a time $t$.
-   $O(t)$ is the proportion of time of the year that the group spends in the camp for each time $t$. It is a continuous variable with range $[0,1]$. In the code, this is defined as the parameter `in_camp_stay` and it is provided as the number of weeks, considering 52 the maximum number of weeks for a year.
-   $F(t)$ is the proportion of bone for the full weight of each Kg of an animal consumed for a time $t$. This is a unit interval variable $[0,1]$.
-   $G(t)$ is the quantity, in Kg, of a (full-weight) animal consumed in a site by each individual at a time $t$. This is a continuous variable with range $[0,\propto K]$. This is a hierarchical variable, defined as follows:

$$G(t)=S(t)\times\frac{C(t)}{R(t)} = S(t)\times\frac{B(t)\times M(t)}{R(t)} \tag{3}$$

Where:

-   $S(t)$ is the proportion of meat consumed within the camp. $[0,1]$

-   $R(t)$ is the quantity of kilocalories produced for each kilogram of meat. This is a continuous with range $[1,2.5]$. Total kilocalories have been divided by 1000 for easier computation.

-   And $C(t)$ is the quantity of kilocalories, extracted from meat, consumed by person for each time $t$. It is in turn defined as $C(t)=B(t) \times M(t)$, where

    -   $B(t)$ is the quantity of kilocalories consumed by an individual for each $t$, and it has a range $[1.5,2.5]\times 365$. For easier computation, the actual average of kilocalories consumed per person/day has been divided by 1000.

    -   $M(t)$ is the proportion of kilocalories coming from meat relating to the total calories consumed per each individual. $[0,1]$.

As it can be seen, the above definitions embed the functionality of the site, which can be considered by the user. In practice, setting the variables $O(t) = 1$ and $S(t) = 1$ would eliminate site functionality. This would assume that a site is inhabited by the full group during the full year, and that this group consumes all the food in the site.

Up to now, we have obtained the value for the variable $A$, which is the amount of kilograms of bone produced at each time $t$. If we consider 4 grams of non-heated bones per sample, we can extract the number of samples at time $t$, which we cal $W(t)$

$$W(t)=\mbox{Integer part of } \Bigg(\frac{1000 \times A(t)}{4} \Bigg) \tag{4}$$

This protocol is captured in the function `A_rates`:

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/anthropogenic_deposition.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/anthropogenic_deposition.Rmd', '.')
```

You can inspect the `A_rates` code here:
<details><summary>Show code</summary>
**Code**

```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='anthropogenic_deposition.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('anthropogenic_deposition.Rmd')
```
</details>

### Depth $D$-protocol

This protocol is designed to respond to the fact that, despite a bone is deposited at time $t$, it can occupy the estratigraphic space of many subsequent years until it is fully buried. The amount of years it takes to be fully buried depends on the thickness of the bone and the natural deposition rates $r$.

Since the thickest bones are usually the minority within a sample, while the majority is usually conformed of smaller bones, we have considered that bone presence decreases exponentially from $t_0$, the year when it was deposited. Thus, this process is modeled as follows:

$$W_b(t_u,t_l)=W(t=t_l)\times(1-e^{-\lambda(t_u-t_l)}),\ t_u < t_l \tag{5}$$

Where

-   $W$ as defined before, it is the number of samples deposited at a time $t_l$
-   $W_b$: is the number of samples buried at a time $t_u$ from the ones generated at $t_l$. Thus, $t_u \geq t_l$
-   $t_l$ is the moment in time $t$ when the samples $W$ are deposited.
-   $t_u$ is each time $t$, $t \geq t_l$, when the sample is being buried according to the parameter $r$.
-   At $t_{u_{max}}$, the sample is completely buried. $t_{u_{max}}>=t_u>=t_l$

$t_{u_{max}}$ depends on two factors:

-   $L$ is maximum thickness of the thickest bone in the direction of the deposition, in cm.
-   $r$ is the natural deposition rate, in $\mbox{cm/year}$. For this model, we have considered the values 0.1, 0.2, 0.3, 0.4 and 0.5

Thus we can calculate the time it takes for a sample to be fully buried as

$$r = \frac{L}{t_{u_{max}}} \Rightarrow t_{u_{max}}=\frac{L}{r} \tag{6}$$

**Estimation of** $\lambda$

Now we need to estimate the exponential rate $\lambda$ in equation $(5)$. In order to do so, we consider the state of the model when $t_l=0$ and $t_u=t_{u_{max}}$, which marks the time at which the first samples deposited are completely buried. And consider the following

$$\frac{W_b(t_u,t_l)}{W(t=t_l)} \tag{7}$$

is the proportion of buried samples in $t_{u}$ from the ones that were deposited in $t_l$. Thus, if we substitute the time points specified above and plug in equation 5, we have that

$$\frac{W_b(t_u=t_{u_{max}},t_l)}{W(t_l=1)}=1-e^{-t_{u_{max}}\lambda} \tag{8}$$

However, since we consider that all of the samples are buried in $t_{u_{max}}$, then the proportion on the left hand side is 1, and in this situation $\lambda \rightarrow \infty$ so we cannot use the equation. We can especify an error margin, $\theta_e$ to the proportion of buried samples in $t_{u_{max}}$, say $99\%$. In this case, $\lambda$ can be easily computed as follows

$$\theta_e=1-e^{-\lambda t_{u_{max}}}$$

and thus, solving for $\lambda$ and taking logarithms on both sides,

$$\lambda=\frac{-log(1-\theta_e)}{t_{u_{max}}} \tag{9}$$

Where $\theta_b<1$ is the proportion of buried samples in $t_{u_{max}}$, which is calculated in equation 6.


At the end, the function `Rec_c` runs `D_along` over a period of time, effectively producing an archaeological record. In the function `Rec_c` the user can consider whether to divide the total sample by the area of the site, or not. In addition, the user can use the modeled values (with the functions above) as an input to this function, or introduce them directly, for example, in the case that fix population rates per year were preferred.

The function `D_along` captures this process.

As usual, you can learn all about the parameters of `D_rates` and `Rec_c`

```{r, eval=FALSE}
?D_along
?Rec_c
```
<br>

```{r, echo=FALSE, message=FALSE}
a = knitr::spin('../../R/natural_deposition.R', format='Rmd', comment=c("#'","#'"),
            precious=TRUE, knit=FALSE, report=FALSE)
fs::file_move('../../R/natural_deposition.Rmd', '.')
```

As before, you can check the code here:
<details><summary>Show code</summary>
**Code**
```{r, echo=FALSE, results='asis'}
res = knitr::knit_child(input='natural_deposition.Rmd', quiet=TRUE)
cat(res, sep = '\n')
fs::file_delete('natural_deposition.Rmd')
```
</details>



## Minimal example and simulation

With the above functions, the user can apply to any generative demographic process: (1) how many samples would a group of people produce in a site, according to different constraints and (2) how those dates distribute along the depth of a site under the assumption that there has not been any kind of record loss or post-depositional process (this will be modelled in a second layer of modelling).

We can create a very basic and simple simulation to illustrate these processes following the parameters seen below (or defaults in function definitions above) following these steps:

1.  $Pop$-protocol. Using `Pop_stoch`, we generate a stochastic population dynamic that will give us the number of individuals for each year.
2.  $A$-protocol. Knowing the population size per year, we apply `A_rates` to each of them and generate the anthropogenic deposition, and get the number of samples per year.
3.  $D$-protocol. Knowing how many samples are generated each year, using `Rec_c`, we distribute them along the depth record.

```{r mini-simu}
set.seed(1234)
# 1. Pop-protocol
Ps <- Pop_stoch(pop_size=100, ts=100, K=150)
# 2. A-protocol
samp <- unlist(lapply(Ps, A_rates)) ## Extract sample per time steps
# 3. D-protocol
Rec <- Rec_c(samp, InitBP = 7500, ts = 100, r = 0.2, max_bone_thickness = "m")
```

With this we obtain, among other things, the distribution of the samples deposited for each date along the depth of the site, and its comparison to the total population. For better comprehension, we have decided not to include all of the dates in the plot below, but only some of them as an example. The user can also check how the model works and its results [here](https://acortell3.shinyapps.io/archaeo_record_model). *The total number of population has been adapted for visualisation*

```{r visu-record, echo = FALSE, message=FALSE}

## Plot dates
sdate <- 1
edate <- 5
sdepth <- 1
edepth <- 5
barplot(Rec[sdepth:edepth,sdate:edate], col = viridis(10), legend = rownames(Rec)[sdate:edate],
       xlab = "Years BP", ylab = "# samples", main = "Distribution of samples for each year at different depths")

tRec <- t(Rec)
barplot(tRec[c(1:4,77:79),], col = viridis(10), legend = colnames(Rec)[c(1:4,77:79)],
        xlab = "Depth", ylab = "# samples", main = "Distribution of specific years along depth with population")
lines(Ps*300, col = "darkred", lwd = 1.5)
```

Notice the exponential decay in the amount of samples per depth layer for the same year.
