# Archaeoriddle: the original challenge {#original-challenge}

We will here go through the whole steps we presented throught all Chapter \@ref(env) to \@ref(final-output), but in a condensed way and with the parameters use to generate the data shared during the archaeoriddle original challenge. 

We will also discuss some of the proposal submitted for the challenge and you will be able to try the challenge too!



## Context 


## Implementation


As I (simon) am stupid I didn't generate seeds for each experiment so we cannot re-run it EXACTLY the same. But I kept lot of informations and output about it.

They are stored in the folder `general_results_selected_simu/`

and will allow us to reconstruct most of what happened. We basically run thousands of simulations and choose the one who looked the cool 😆. We used the script `scriptmini.R` that allow to run one simulation and ran in in parallel as described in `smallscript.md`.


This summarise all the previous steps but using the original map and sites.

The original raster:

```{r}

original.ras <- rast("data_original/east_narnia4x.tif")
plot(original.ras, col=col_ramp(20), legend=F, reset=F)
original.wat <- original.ras
original.wat[original.wat>mean(original.wat[])] <- NA
plot(original.wat, col="lightblue", add=T, legend=F)

original.groups <- original.ras # taking back the raster created before
original.groups[original.groups<mean(original.groups[])] <- NA #replace values that are below mean level (under water)
original.groups[original.groups<(maxh*.7)] <- 1 # values below 70% of the maximum original are set as 1
original.groups[original.groups>(maxh*.7)] <- 200 # value above at 200 (high mountains)
original.groups[is.na(original.groups)] <- -1 #  NA values (below waters) are set to -1
original.poly <- as.polygons(original.groups) # convert original.groups as polygons that will be of three type: -1 => under water, 1, viable, and 200, high mountaines
original.viable <- makeValid(original.poly[2,]) # select polygon with '1' and make it a valid polygon as it may have some loop and problems
# get only the non submerged actual land and plot it
above_level <- original.ras > mean(original.ras[])
coastline <- st_as_sf(as.polygons(above_level))[2,]
plot(coastline, col=NA, bgc=adjustcolor("cyan", 0.1), add=T)
```

Adding ressource

```{r}
##ressource geolocalisation set manually
goodresources <- vect(
  cbind(
    x=c(-0.2300711, -3.1455282, -0.5086485, -1.9639755,
        -0.4077843, 0.019688, -3.116710),
    y=c(3.6423000, -0.2551019, -0.7440748, 1.1303214,
        1.0248567, 0.2194895, 2.0267718)
  )
)
#spread of resources
areas <- 4 * c(100, 2000, 200, 300, 100, 200, 400)
#speed of ressource decay:
ks <- c(0.0002, 0.0001000, 0.0001600, 0.0001800, 0.00040, .0002, 0.0002)/4

crs(goodresources) <- crs(original.ras)

original.allres <- lapply(
  seq_along(goodresources),
  function(i){
    logisticdecay(goodresources[i], mask(original.ras, original.viable),
                  x=areas[i], k=ks[i]
    )
  }
)

allna <- sapply(original.allres, function(i)any(is.na(values(i))))
original.allres <- original.allres[!allna]
original.ress <- original.allres[[1]]
for(i in 2:length(original.allres))
    original.ress <- original.ress + original.allres[[i]]
original.ress <- mask(original.ress,original.viable)
plot(original.ress)
```

This, should be exactly what is stored in `data_original/resources.tiff`

```{r comptiff,out.width="45%" }
plot(original.ress)
plot(rast("data_original/resources.tiff"))
```

We then take back the sites position defined for the original archaeoriddle:

```{r}
original.sites=vect("data_original/sitesinitialposition/")


Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures

# spatial penality to extent: lower, bigger penality
cul_ext <- c("HG"=7, "F"=6)
# penality of occupational area: low, other sites can come close
penal_cul <- c("HG"=4, "F"=5)
# proba to give birth every year
prob_birth <- c("HG"=0.3, "F"=0.5)
# proba to die when pop > K
prob_survive <- c("HG"=0.8, "F"=0.6)
# proba to create new settlement when Ne > K
prob_split <- c("HG"=0.2, "F"=0.6)
# how big the group of migrant should be to create a new city vs
# migrate to a existing one
minimals <- c("HG"=0.14, "F"=0.20)
# prob to migrate to existing settlement when Ne > K
prob_move <- c("HG"=0.2,"F"=0.1)
```

Let's check again our raster and the sites and there cutlures etc...
```{r}
plotMap(original.ras,original.wat,paste0("year ", 0))
points(crds(original.sites), pch=21, bg=rainbow(2, alpha=0.6)[as.factor(original.sites$culture)],col=1)
text(original.sites,pos=3)
```


Let's run one simple simulation on the archaeoriddle:

```{r,eval=T}

ts <- 100

print(paste0("Starting simulation ","nan"))

onesimu <- run_simulation(
  sites=original.sites, viable=original.viable, dem=original.ras,
  ressources=original.ress,
  water=original.wat,
  foldervid="nan",
  visu=F, visumin=F,
  ts=ts, #length of simulation in year
  Kbase=c("HG"=35, "F"=110), #difference in K for the two cultures
  cul_ext=c("HG"=7, "F"=6), #spatial penality to extent: lower, bigger penality
  penal_cul=c("HG"=4, "F"=5), #penality of occupational area: low, other sites can cam close
  prob_birth=c("HG"=0.3, "F"=0.5), #proba of giving birth every year
  prob_survive=c("HG"=0.8, "F"=0.65), #proba of dying when pop > K
  prob_split=c("HG"=0.5, "F"=0.6), #proba of creating a new settlement when Ne > K
  minimals=c("HG"=0.14,"F"=0.20), #how big the group of migrant should be to create a new city vs migrate to a existing one 
  bufferatack=300, #distance max around which settlement can fight
  prob_move=c("HG"=0.2, "F"=0.1) #proba of migrating to existing settlement when Ne > K
)
```

```{r,cache=F}
Nts <- onesimu$Nts # population at each timestep
warcasualties <- onesimu$warcasualties #death by war at each time step
plot(sites, cex=(1+Nts[1,]/100), pch=21, bg=rainbow(2, alpha=0.6)[as.factor(original.sites$culture)])
```



```{r,eval=T,messages=F}

i <- nrow(Nts)
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)), xlab="time", ylab="popsize")
lapply(1:ncol(Nts),
       function(j)lines(Nts[,j],col=rainbow(2)[as.factor(original.sites$culture)[j]])
)

```

Visualise, Record deposit and  loss. for the 5 first sites (see next chapters for `A_rates`).
```{r,out.width="25%"}
alldeposit <- lapply(
  1:5,
  function(i){
    Rec_c(sapply(Nts[,i], A_rates), InitBP = 15000, ts=nrow(Nts), r=0.2, max_bone_thickness="m")
  }
)
maxy <- max(sapply(alldeposit, function(i)max(apply(i,2,sum))))

nill <- lapply(alldeposit, function(depo)barplot(t(depo), col=viridis(ts+1), ylim=c(0,maxy)))
```

Population by time and culture

```{r,eval=T}
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)),
     xlab="time", ylab="popsize")
lapply(1:ncol(Nts),
       function(i)lines(Nts[,i], col=rainbow(2)[as.factor(original.sites$culture)[i]]))
plot(apply(Nts, 1, sum)[1:i], xlab="time", ylab="popsize")

```


Exploring some trajectories of the simulations

```{r, eval=T}
i <- nrow(Nts)  # Get the number of rows in Nts
plot(apply(Nts[1:i, original.sites$culture=="F"], 1, sum), col="red", type="l",
     lwd=2, ylim=c(0, max(apply(Nts, 1, sum))))  # Plot sum of 'F' culture values
points(apply(Nts[1:i, original.sites$culture=="HG"], 1, sum),
       col="blue", lwd=2, type="l")  # Add points for 'HG' culture values
plot(warcasualties[1:(i-1)], lwd=2, col="green", type="h", yaxt="n", ylab="")  # Plot war casualties
axis(4)  # Add an axis on the right side
par(new=T)  # Allow a new plot to be drawn on the existing plot
growF <- apply(Nts[1:(i-1), original.sites$culture=="F"], 1, sum)  # Sum of 'F' culture values excluding last row
growHG <- apply(Nts[1:(i-1), original.sites$culture=="HG"], 1, sum)  # Sum of 'HG' culture values excluding last row
plot(growF, col="red", type="l", lwd=2, ylim=c(0, max(growF, growHG)))  # Plot growth of 'F' culture
points(growHG, col="blue", lwd=2, type="l")  # Add points for growth of 'HG' culture
```



### Generate and share final simulation

No we have the full setup, we want to generate a record people will uses :


1. We first load the maps and the sites:

```{r loadressources}
sites=vect("data_original/sitesinitialposition/")
```

2. run one simulation


```{r fakesimu,eval=F}

ts=1000
expname="basar"
print(paste0("Starting simulation ",expname))

onesimu=run_simulation (
                           sites=sites,
                           viable=viable,
                           dem=height.ras,
                           ressources=rast("data_original/resources.tiff"),
                           water=height.wat,
                           foldervid=expname,
                           visu=F,visumin=TRUE,
                           ts=ts,#length of simulation in year
                           Kbase=c("HG"=35,"F"=110),#difference in K for the two cultures
                           cul_ext=c("HG"=7,"F"=6),#spatial penality to extent: lower, bigger penality
                           penal_cul=c("HG"=4,"F"=5),#penality of occupational area: low, other sites can cam close
                           prob_birth=c("HG"=0.3,"F"=0.5),#proba to give birth every year
                           prob_survive=c("HG"=0.8,"F"=0.65),#proba to die when pop > K
                           prob_split=c("HG"=.5,"F"=.6),#proba to create new settlement when Ne > K
                           minimals=c("HG"=.14,"F"=.20),#how big the group of migrant should be to create a new city vs migrate to a existing one 
                           bufferatack=300,#distance max around which settlement can fight
                           prob_move=c("HG"=0.2,"F"=0.1) #proba to migrate to existing settlement when Ne > K
                           )
```


2. generate a few output, including a video that helps to select the good simulation, these things are also done by `scriptmini.R`:

```{r getfakeres,eval=F}


Nts=onesimu$Nts
warcasualties=onesimu$warcasualties
sites=onesimu$sites
i=min(ts,which(apply(Nts,1,sum)==0))
pdf(paste0(expname,"_mapFinal.pdf"))
plotMap(height.ras,height.wat,paste0("year ",i))
plot(sites,cex=(as.integer(Nts[i,]>0)*0.3+Nts[i,]/200),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
text(sites)
dev.off()


pdf(paste0(expname,"growth_utils.pdf"))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="green",type="h",yaxt="n",ylab="",xlim=c(0,i))
axis(4)
par(new=T)
growF=apply(Nts[1:i,sites$culture=="F"],1,sum)
growHG=apply(Nts[1:i,sites$culture=="HG"],1,sum)
plot(growF,col="red",type="l",lwd=2,ylim=c(0,max(growF,growHG)),xlim=c(0,i))
points(growHG,col="blue",lwd=2,type="l")
dev.off()

pdf(paste0(expname,"growth_tot.pdf"))
plot(warcasualties[1:i-1],lwd=2,col="green",type="h",yaxt="n",ylab="")
axis(4)
par(new=T)
growT=apply(Nts[1:i,],1,sum)
plot(growT,col="black",type="l",lwd=2,ylim=c(0,max(growT)))
dev.off()
```

Let's redo that with the selected simulation:


```{r initialcondition}

onesimu=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_all.RDS")
#Nts=readRDS("nts")
#warcasualties=readRDS("war")
#sites=terra::vect(readRDS("general_results_selected_simu/buffattack300_K110_PSU065_3_sitesRast.RDS"))
Nts=onesimu$Nts
warcasualties=onesimu$warcasualties

```

Plot the sites at the beginning 

```{r plotinitialcondition,cache=F}
# source("tools.R")

#expname="general_results_selected_simu/buffattack300_K110_PSU065_3"
#sites=vect(readRDS(paste0(expname,"_sitesRast.RDS")))

sites=vect("data_original/sitesinitialposition/")
height.ras=rast("data_original/east_narnia4x.tif")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA

i=1
plotMap(height.ras,height.wat,paste0("year ",i,", when it all begins"))
plot(sites,cex=(as.integer(Nts[1,]>0)*0.5+Nts[1,]/100),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
text(sites[Nts[1,]>0],pos=3)

```

We can look at the war and the number of deaths for each battle

```{r plotwar}
i=1000
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",ylab="# dead",xlim=c(0,i))
```

Comparison to the total population growth of each culture

```{r plotwarWRTgrowths}
i=1000
par(mar=c(5,5,1,5))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",yaxt="n",ylab="",xlim=c(0,i),xlab="time")
axis(4)
mtext("# dead",4,2.5)
par(new=T)
growF=apply(Nts[1:i,sites$culture=="F"],1,sum)
growHG=apply(Nts[1:i,sites$culture=="HG"],1,sum)
plot(growF,col="orange",type="l",lwd=5,ylim=c(0,max(growF,growHG)),xlim=c(0,i),ylab="pop",xlab="")
points(growHG,col="blue",lwd=5,type="l")
```

If we look at the total population size, we can see how the clashes slowed down the population growth:
```{r plotwarTotgrowths}
par(mar=c(5,5,1,5))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",yaxt="n",ylab="",xlim=c(0,i),xlab="time")
axis(4)
mtext("# dead",4,2.5)
par(new=T)
growT=apply(Nts,1,sum)
plot(growT,col="black",type="l",lwd=5,ylim=c(0,max(growT)),xlim=c(0,i),ylab="pop",xlab="")

```

This show if we looka at the growth rate:

```{r rategrowth}
plot(diff(growT,lag=100),type="l")
```


Replaying all that will give this:


```{r finalvideo,echo=F,results="asis"}
cat('<video controls> <source  src="interfinal.mp4" type="video/mp4"></video>')
```


```{r getRaster,eval=T,cache=F}
height.ras=rast("data_toshare/dem_raster.tiff")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA
height.groups=height.ras
maxh=max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])]=NA
height.groups[height.groups<(maxh*.7)]=1
height.groups[height.groups>(maxh*.7)]=200
height.groups[is.na(height.groups)]=-1
height.poly=as.polygons(height.groups)
viable=makeValid(height.poly[2,])
expname="data_original/general_results_selected_simu/buffattack300_K110_PSU065_3"
onesimu=readRDS(paste0(expname,"_all.RDS"))
sites=readRDS(paste0(expname,"_sitesRast.RDS"))
if(class(sites)[1] == "PackedSpatVector")
    sites=terra::vect(sites)
Nts=onesimu$Nts
#alldeposit=lapply(1:ncol(Nts),function(s)Rec_c(sapply(Nts[1:i,s],A_rates), InitBP = 7500,ts=ts,r = 0.2, Max_bone_thickness = "m"))
#allShortLoss=lapply(1:10,function(ind){Rec=alldeposit[[ind]];print(paste0("settlement #",ind));apply(Rec,2,short_loss,.6)})
#allShortLoss=lapply(allShortLoss,function(Rec)long_loss(Rec,.9997,7500))
```

Using `Rec_c` , `short_loss` and `long_loss` we generate deposit. Generating everything can take long time but  it's basically done like this:


```{r doloss,eval=F}

#All in one:
alldeposit=lapply(1:ncol(Nts),function(s)Rec_c(sapply(Nts[,s],A_rates), InitBP = 7500,ts=ncol(Nts),r = 0.2, Max_bone_thickness = "m"))
allLosses=lapply(1:length(alldeposit),function(ind){
                 st=Sys.time()
                    print(paste0("settlement #",ind));
                    Rec=alldeposit[[ind]];                    
                    Rec=apply(Rec,2,short_loss,.6)  #apply short loss
                    Rec=long_loss(Rec,.9997,7500) #apply long loss
                    print(Sys.time()-st)
                    return(Rec)
})
rm(alldeposit)
```

Remove where to small

```{r removesmall, eval=F}


maxSites=max(sapply(allLosses,sum))
nsample=round(sapply(allLosses,sum)*30/maxSites)
allRemainingDates=lapply(seq_along(allLosses),function(r)extractDates(allLosses[[r]],n=nsample[r]))
rm(allLosses) #object are big, need to free memory

```



```{r getadates,eval=F}

#pdf(paste0(expname,"_mapFound.pdf"))
#plotMap(height.ras,height.wat,paste0("final after losses"))
#plot(sites[1:10,],cex=3*(lengths(allRemainingDates)-1)/(29),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture[1:10])])
#dev.off()

dates=unique(unlist(allRemainingDates))
dates=rev(sort(dates[!is.na(dates)]))
plot(table(unlist(allRemainingDates)))
totallDatesRemains=sapply(allRemainingDates,function(i)table(factor(i,levels=dates)))
saveRDS(allRemainingDates,file=paste0(expname,"_dates.RDS"))

```

From this we have, for every site, a list of date(s)

This has been done using the script `exploreDate.R` but will be explained here:


Loading simulation results, green are the sites who survived, dots are all the site that existed.

```{r getfakeresults,cache=F}
height.ras=rast("data_original/east_narnia4x.tif")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA


Nts=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_all.RDS")$Nts
ress=rast("data_original/resources.tiff")
allsites=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_sitesRast.RDS")
if(class(allsites)[1] == "PackedSpatVector")
    allsites=terra::vect(allsites)
plotMap(height.ras,height.wat)
plot(allsites,add=T)
plot(allsites[Nts[nrow(Nts),]>0,],add=T,pch=21,bg="green",cex=Nts[nrow(Nts),]/60)
```

Loading dates:

```{r getfakedates,cache=F}
dates=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_dates.RDS")
```

Keep sites with foundable dates

```{r extractdates,cache=F}

foundsites=allsites[lengths(dates)>0,]
foundsites$numdates=unlist(lengths(dates[lengths(dates)>0]))
founddates=dates[lengths(dates)>0]
```

Remove BP to the dates, add some random variation

```{r addstdev,cache=F}
stdpool=c(20,30,40,50,60,80,100,120)
founddates=lapply(founddates,sort)
founddates=lapply(founddates,gsub,pattern=" BP",replacement="")
founddates=lapply(founddates,rev)
founddates=lapply(founddates,function(i)paste0(i," ± ",sample(stdpool,length(i),replace=T,prob=c(3,3,3,3,2,2,1,1))," BP"))
foundsites$dates=sapply(founddates,paste0,collapse=" | ")
plotMap(height.ras,height.wat)
plot(foundsites,cex=foundsites$numdates/20+1,pch=21,bg=as.factor(foundsites$culture),add=T)
```

### Split environment in a grid

Split the environment in grids and extract date for each grid

```{r plotallfoundiste,cache=F}

leftdates=dates[lengths(dates)>0]
plotMap(height.ras,height.wat)
plot(foundsites,cex=foundsites$numdates/20+.1,pch=21,bg=1,add=T)
squares=st_make_grid(height.ras,.5)
plot(squares,add=T,col=adjustcolor(rainbow(length(squares)),.35))
text(st_coordinates(st_centroid(squares)),label=1:length(squares),col="white")
text(st_coordinates(st_centroid(squares)),label=1:length(squares),col="white")
```

Public selection

```{r getpublicsele,cache=F}
plotMap(height.ras,height.wat)
plot(foundsites,cex=foundsites$numdates/20+1,pch=21,bg=as.factor(foundsites$culture),add=T)
selection=c(14,30,45,65,66)
plot(squares[selection],add=T,col=adjustcolor("blue",.3))
inter=st_intersection(st_as_sf(foundsites),squares[selection])
plotMap(height.ras,height.wat)
plot(st_geometry(inter),add=T,bg=rainbow(2,alpha=.6)[as.factor(inter$culture)],pch=21,cex=1+inter$numdates/10)
```

### Giving name to the publicly available data {#naming}

The names have bee ngenerated via: [fantasynamegenerators.com](https://www.fantasynamegenerators.com/fantasy-town-names.php)

```{r namesites,cache=F}

site_dist=st_distance(inter)
#min(site_dist[as.numeric(site_dist)>units(0)])

sitesnames=1:nrow(inter)

sitesnames[c(1:4, 20)]=c("Farwallow" ,"Bearcall" ,"Dustscar" ,"Clearreach" ,"Rabbithole")

fr=c("Épibéliard" ,"Cololuçon" ,"Pulogne" ,"Haguemasse" ,"Auriteaux" ,"Bourville" ,"Banau" ,"Montnesse" ,"Bannet" ,"Alenlon", "Roullac" ,"Genneville" ,"Vinlès" ,"Antonnet" ,"Courtou" ,"Beaulogne" ,"Coloville" ,"Sarsart" ,"Soilon" ,"Cololimar")
sitesnames[5:19]=fr[1:(19-4)]
spain=c("Zava" ,"Catadrid" ,"Tegon" ,"Alicia" ,"Mulid" ,"Zararbella" ,"Malid" ,"Cásca" ,"Granalejos" ,"Segorez" ,"Terteixo" ,"Astumanca" ,"Galle" ,"Talona" ,"Girovega" ,"Albanada" ,"Nadoba" ,"Senca" ,"Vallanca" ,"Taville")

sitesnames[21:length(sitesnames)]=spain[1:(length(sitesnames)-20)]
inter=cbind(inter,sitesnames)
plotMap(height.ras,height.wat)
plot(st_geometry(inter),add=T,bg=rainbow(2,alpha=.6)[as.factor(inter$culture)],pch=21,cex=1+inter$numdates/10)
text(st_coordinates(inter),inter$sitesnames,cex=.8)
```

Export csv and data for each public square

```{r exportpublic,eval=F}


inter$sitesnames=sitesnames
for(g in selection){
curr=inter[inter$ID==g,]
coords=st_coordinates(curr)
    write.csv(file=paste0("square_",g,".csv"),cbind.data.frame(sitename=curr$sitesnames,lon=coords[,1],lat=coords[,2],dates=curr$dates,economy=curr$culture))
}

```

Export csv and data for all ever square

```{r exportprivate,eval=F}

allsites=st_intersection(st_as_sf(foundsites),squares[-selection])

for(g in (1:length(squares))[-selection]){
curr=allsites[allsites$ID==g,]
coords=st_coordinates(curr)
    write.csv(file=paste0("square_",g,".csv"),cbind.data.frame(lon=coords[,1],lat=coords[,2],dates=curr$dates,economy=curr$culture))
}

ld=strsplit(inter$dates," \\| ")
ld=lapply(ld,function(i)gsub(" ± .*","",i))
inter$start=sapply(ld,max)
inter$end=sapply(ld,min)

```


Some maps we use and did for fake papers and social media etc... 


```{r randommaps,out.width="30%",fig.show="hold"}

## plotting oldschool map

par(mar=c(0,0,0,0),oma=c(0,0,0,0))

plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=.8,col=as.factor(inter$culture),alpha=.8)
box(which="outer")

old=inter[inter$start>7180,]
ne=inter[inter$start>6800 & inter$end<6900,]

plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(old),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(old),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(old$culture),alpha=.8)
box()


plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
box()


plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
box()

plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(ne),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(ne),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(ne$culture),alpha=.8)
box()
```

More maps for twitter:

```{r randommaps2}
plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.2,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
box()
sel=inter[c(20,2,15,36),]
text(st_coordinates(sel),sel$sitesnames,cex=.8,pos=c(1,2,2,4))
text(x = -1, y = .7, labels = "❤️", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -1, y = 1.2, labels = "🔥 ", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -1, y = 1.2, labels = "⚔️ ", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -1.2, y = 1.8/2, labels = "???? ", adj=c(0,0), cex = 3, col = "black", font = 1)
```
More maps!! (the coolest)


```{r randommaps3}
plotMap(height.ras,height.wat)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.2,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
coastline=st_cast(st_as_sf(as.polygons(height.ras>mean(height.ras[])))[2,],"MULTILINESTRING")
plot(coastline,add=T,lwd=1.1,col="black")
box()
sel=inter[c(20,2,15,36),]
text(st_coordinates(sel),sel$sitesnames,cex=.8,pos=c(1,2,2,4))

```


```{r}

plotMap(height.ras,height.wat)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.2,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
coastline=st_cast(st_as_sf(as.polygons(height.ras>mean(height.ras[])))[2,],"MULTILINESTRING")
plot(coastline,add=T,lwd=1.1,col="black")
box()
sel=inter[c(20,2,15,36),]
text(st_coordinates(sel),sel$sitesnames,cex=.8,pos=c(1,2,2,4))
#text(x = -1, y = .7, labels = "🍻", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -3.4, y = .7, labels = "🍻 ", adj = c(0, 0), cex = 10, col = "black", font = 1)
text(x = -1.8, y = .7, labels = "👍 ", adj = c(0, 0), cex = 8, col = "black", font = 1)
text(x = -.5, y = .7, labels = "🍔 ", adj = c(0, 0), cex = 6, col = "black", font = 1)
#text(x = -1, y = 1.2, labels = "⚔️ ", adj = c(0, 0), cex = 3, col = "black", font = 1)
#text(x = -1.2, y = 1.8/2, labels = "???? ", adj=c(0,0), cex = 3, col = "black", font = 1)


```

## The challenge

### Research Question 

RQ1. The Poppy-chewers and Rabbit-skinners had a hostile relationship.
RQ2. Poppy-chewers followed an exponential population growth, while Rabbit-Skinners eventually stalled, and as the Poppy-chewers population size started to increase, their population declined (Fig. 3).
RQ3. The rate of dispersal varied across the landscape, with an average of ~0.62 Km/year, faster for the sea crossing (~1.11 Km/year) and a slight slow-down to ~0.57 Km/year for the Northeastern quadrant of the map.

### Challenge & EAA workshop


## Proposals



### P1

P1 used Agent-Based modelling (ABM) combined with exploratory data analysis. It first studied the land of Rabbithole and calibrated the dates provided, from which they computed trajectories of dispersal and studied site preference using ArcGIS pro and R. The suggested dispersal rates over land and water were close to the Archaeoriddle solution. The proposal then used the results of its analyses to fit the ABM, which was built using NetLogo. For the ABM, moving groups of hunter-gatherers and farmers, as well as the different sites were treated as agents, with starting values obtained from the exploratory data analysis and relevant literature. Additionally, it introduced behavioural rules including movement and site preference, reaction to population threshold or reaction to interaction. After running the model several times with different parameters, it correctly addressed a hostile relation between the two groups and inferred an initial location of the Rabbit-skinners at the Northeast and the Poppy-chewers at the Southwest, detecting their northward movement. The expansion rate was not captured by the initial assumption of logistic growth in the areas already densely inhabited by Poppy-chewers, which makes sense considering the different population trajectories of the two groups. The full proposal can be found here.


### P2 

P2 used Point Process Modeling (PPM) to assess the problem. First, it focused on the sampling process and built first-order point process models combining fitness and the available grids to predict potential occupation. It focused on the question of whether there was conflict between the two groups or not. To do this, after building different archaeological phases, it computed the clustering patterns of the groups under the assumption that higher clustering could lead to higher conflict (Field 2003). Following this, it computed the interaction distance between groups of sites of hunter-gatherers and farmers through a Multitype Strauss Model. According to the results, hostilities increased as time passed. In the original model, the rules for hostility were not time-dependent. However, it does make sense that, even if the rules for conflict do not change, as the population grows and sites increase there is more probability of contact and, thus more probability of conflict. The full proposal can be found here.

### P3

P3 (full proposal https://doi.org/10.5281/zenodo.8260754) used Species Distribution Modelling instead. Using R, it proposed a four-stage research design. The first stage focused on determining which additional cells would result in a representative sample of the range variables (elevation and resource quality) present within the data. The second stage focused on data exploration to identify internal temporal, spatial, and farmer-forager patterns within the data. In the third stage, it generated summed probability distributions (SPDs) from the calibrated radiocarbon dates to get relative estimates of population size through time for foragers and farmers. The fourth stage combined these data into a spatiotemporal species distribution model, where both time and space were explicit predictors, to estimate the distribution of farmers and foragers in 100-years intervals. The spatiotemporal species distribution was successful in reproducing the directionality of the farming dispersal (from South to North) as well as the decline of hunter-gatherer populations.

### P4 

P4 first developed a strategy to aid the selection of additional data using a friction calculation considering the distance from a putative origin region, the elevation of the region and the environmental fitness of the region. Also using R, the analyses focused on capturing local complexity in the dispersal of Poppychewers in the study area. To track this, it used a Hierarchical Bayesian Phase Model, informed by all the selected sites (the common ones, and the ones obtained after calculations for additional sampling). This method allowed for the introduction of uncertainty within her outcome. It divided the map of Rabbithole into 25 areas, and calculated the times of arrival of the Poppy-chewers for each of these areas, including their High Probability Density Intervals (HDPI). The full approach can be found at: https://github.com/AlexesMes/Archeaoriddle_RabbitWorld.


### P5

P5 used a qualitative approach. In his case, it rather provided a theoretical framework. It considered two cultures A and B initially established in regions R_A and R_B separated by a geographical barrier (the sea from Rabbithole). After this, culture A gains the ability to cross or bypass the barrier entering the region R_B and getting into contact with B. Under the condition that all other parameters are more or less equal (e.g. environmental fitness in both regions, no new pathogens or predators hitherto unbeknown to A), a significantly shorter settlement persistence of distribution of settlements of culture A in R_B as compared with A’s settlement persistence distribution assessed in other regions, implies a hostile relation between A and B.

