# Archaeoriddle: the original challenge {#original-challenge}

We will here go through the whole steps we presented throughout Chapter&nbsp;\@ref(env) to&nbsp;\@ref(final-output), but in a condensed way and with the parameters used to generate the data shared during the archaeoriddle original challenge. 

We will also discuss some of the proposal submitted for the challenge and you will be able to try the challenge too.

## Context 

After discussion within the member of the CDAL, we decided that a challenge-like event, with a cash-prize, would be the best way to convince people to participate. After the B/A grant has been awarded to Simon Carrignon, Alfredo Cortell and Enrico Crema, we thus decided to announce this challenge through various social media video and a website: https://theia.arch.cam.ac.uk/archaeoriddle/

As this website will disappear anytime soon, here is the first page:


The source to reproduce the website are also available in the github Archaeoriddle repository [here](https://github.com/acortell3/archaeoriddle/tree/main/doc/shinyapp).


## Implementation

The Challenge is basically built on top of a manually selected environment, generated using the tools describe in the previous chapter, on top of which a bunch of simulation have been run, and one has been selected.

But as [Simon Carrinon](sc.fram.io) has do a poor job when he did the simulation he didn't generate seeds, nor stored them, for each experiment ; so we cannot re-run it EXACTLY the same. But he kept a lot of informations and output about it which, are stored in the folder `general_results_selected_simu/`

This should be enough to allow anyone to reconstruct most of what happened. Thus, a quick recap of the implementation of the original challenge:

From all the function developped and describ previously in this book:

1. Generate an environment that look cools
2. Choose a series of parameters and
3. Run hundreds of simulations and choose the one who looked the coolest 😆.
  
We used the script [scriptmini.R](https://github.com/acortell3/archaeoriddle/blob/main/doc/bookdown/scriptmini.R) to run one simulation, and then used a very simple and inelegant bash loop to parallelise muleiple run:

```bash
for t in {1..5} ; do Rscript scriptmini.R newset_$t > log_new_set$t 2> log_new_set$t.err &  done
```

On can see, on line 52 until the end of the `scriptmini.R`. that some graph are saved. This will be used to quickly compare the different simulation and choose the good one.

The simulation are also run with the parameter: `visumin=TRUE `, which allow to save time serie of image at every time step while the model is running.

```bash
for i in test*_? ; do ffmpeg -i "$i/map_000%3d.png" "${i}_out.mp4" ; done
```

## Reminder

Here we will summarise all the previous steps but using the original map and sites.

Load the original raster:

```{r originalrast,fig.show="hold"}

original.ras <- rast("data_original/east_narnia4x.tif")
plot(original.ras, col=col_ramp(20), legend=F, reset=F)
original.wat <- original.ras
original.wat[original.wat>mean(original.wat[])] <- NA
plot(original.wat, col="lightblue", add=T, legend=F)

original.groups <- original.ras # taking back the raster created before
original.groups[original.groups<mean(original.groups[])] <- NA #replace values that are below mean level (under water)
original.groups[original.groups<(maxh*.7)] <- 1 # values below 70% of the maximum original are set as 1
original.groups[original.groups>(maxh*.7)] <- 200 # value above at 200 (high mountains)
original.groups[is.na(original.groups)] <- -1 #  NA values (below waters) are set to -1
original.poly <- as.polygons(original.groups) # convert original.groups as polygons that will be of three type: -1 => under water, 1, viable, and 200, high mountaines
original.viable <- makeValid(original.poly[2,]) # select polygon with '1' and make it a valid polygon as it may have some loop and problems
# get only the non submerged actual land and plot it
above_level <- original.ras > mean(original.ras[])
coastline <- st_as_sf(as.polygons(above_level))[2,]
plot(coastline, col=NA, bgc=adjustcolor("cyan", 0.1), add=T)
```

Adding ressource to it using original geolocalistaiont

```{r originalress,fig.show="hold"}
##ressource geolocalisation set manually
goodresources <- vect(
  cbind(
    x=c(-0.2300711, -3.1455282, -0.5086485, -1.9639755,
        -0.4077843, 0.019688, -3.116710),
    y=c(3.6423000, -0.2551019, -0.7440748, 1.1303214,
        1.0248567, 0.2194895, 2.0267718)
  )
)
#spread of resources
areas <- 4 * c(100, 2000, 200, 300, 100, 200, 400)
#speed of ressource decay:
ks <- c(0.0002, 0.0001000, 0.0001600, 0.0001800, 0.00040, .0002, 0.0002)/4

crs(goodresources) <- crs(original.ras)

original.allres <- lapply(
  seq_along(goodresources),
  function(i){
    logisticdecay(goodresources[i], mask(original.ras, original.viable),
                  x=areas[i], k=ks[i]
    )
  }
)

allna <- sapply(original.allres, function(i)any(is.na(values(i))))
original.allres <- original.allres[!allna]
original.ress <- original.allres[[1]]
for(i in 2:length(original.allres))
    original.ress <- original.ress + original.allres[[i]]
original.ress <- mask(original.ress,original.viable)
plot(original.ress)
```

This, should be exactly what is stored in `data_original/resources.tiff`

```{r comptiff,out.width="45%",fig.show="gold" }
plot(original.ress)
plot(rast("data_original/resources.tiff"))
```

We then take back the sites position defined for the original archaeoriddle and all the original parameters:

```{r originalinit}
original.sites=vect("data_original/sitesinitialposition/")


Kbase=c("HG"=45,"F"=120) #difference in K for the two cultures

# spatial penality to extent: lower, bigger penality
cul_ext <- c("HG"=7, "F"=6)
# penality of occupational area: low, other sites can come close
penal_cul <- c("HG"=4, "F"=5)
# proba to give birth every year
prob_birth <- c("HG"=0.3, "F"=0.5)
# proba to die when pop > K
prob_survive <- c("HG"=0.8, "F"=0.6)
# proba to create new settlement when Ne > K
prob_split <- c("HG"=0.2, "F"=0.6)
# how big the group of migrant should be to create a new city vs
# migrate to a existing one
minimals <- c("HG"=0.14, "F"=0.20)
# prob to migrate to existing settlement when Ne > K
prob_move <- c("HG"=0.2,"F"=0.1)
```

This is `Rabbithole`  home of the Poppychewer and Rabbitskinner!
`

```{r rabbithole}
plotMap(original.ras,original.wat,paste0("year ", 0))
points(crds(original.sites), pch=21, bg=rainbow(2, alpha=0.6)[as.factor(original.sites$culture)],col=1)
text(original.sites,pos=3)
```


Let's run one simple simulation on Rabbithole:

```{r exsimurh,eval=T}

ts <- 100

print(paste0("Starting simulation ","nan"))

onesimu <- run_simulation(
  sites=original.sites, viable=original.viable, dem=original.ras,
  ressources=original.ress,
  water=original.wat,
  foldervid="nan",
  visu=F, visumin=F,
  ts=ts, #length of simulation in year
  Kbase=c("HG"=35, "F"=110), #difference in K for the two cultures
  cul_ext=c("HG"=7, "F"=6), #spatial penality to extent: lower, bigger penality
  penal_cul=c("HG"=4, "F"=5), #penality of occupational area: low, other sites can cam close
  prob_birth=c("HG"=0.3, "F"=0.5), #proba of giving birth every year
  prob_survive=c("HG"=0.8, "F"=0.65), #proba of dying when pop > K
  prob_split=c("HG"=0.5, "F"=0.6), #proba of creating a new settlement when Ne > K
  minimals=c("HG"=0.14,"F"=0.20), #how big the group of migrant should be to create a new city vs migrate to a existing one 
  bufferatack=300, #distance max around which settlement can fight
  prob_move=c("HG"=0.2, "F"=0.1) #proba of migrating to existing settlement when Ne > K
)
```

```{r plotsimurh,cache=F}
Nts <- onesimu$Nts # population at each timestep
warcasualties <- onesimu$warcasualties #death by war at each time step
plot(sites, cex=(1+Nts[1,]/100), pch=21, bg=rainbow(2, alpha=0.6)[as.factor(original.sites$culture)])
```



```{r,eval=T,messages=F}

i <- nrow(Nts)
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)), xlab="time", ylab="popsize")
nill <- lapply(1:ncol(Nts),
       function(j)lines(Nts[,j],col=rainbow(2)[as.factor(original.sites$culture)[j]])
)

```

Visualise, Record deposit and  loss. for the 5 first sites (see next chapters for `A_rates`).

```{r,out.width="25%"}
alldeposit <- lapply(
  1:5,
  function(i){
    Rec_c(sapply(Nts[,i], A_rates), InitBP = 15000, ts=nrow(Nts), r=0.2, max_bone_thickness="m")
  }
)
maxy <- max(sapply(alldeposit, function(i)max(apply(i,2,sum))))

nill <- lapply(alldeposit, function(depo)barplot(t(depo), col=viridis(ts+1), ylim=c(0,maxy)))
```

Population by time and culture

```{r,eval=T}
plot(1, 1, type="n", xlim=c(0,i), ylim=c(0,max(Nts)),
     xlab="time", ylab="popsize")
nill  <- lapply(1:ncol(Nts),
       function(i)lines(Nts[,i], col=rainbow(2)[as.factor(original.sites$culture)[i]]))
plot(apply(Nts, 1, sum)[1:i], xlab="time", ylab="popsize")

```

Exploring some trajectories of the simulations

```{r, eval=T}
i <- nrow(Nts)  # Get the number of rows in Nts
plot(apply(Nts[1:i, original.sites$culture=="F"], 1, sum), col="red", type="l",
     lwd=2, ylim=c(0, max(apply(Nts, 1, sum))))  # Plot sum of 'F' culture values
points(apply(Nts[1:i, original.sites$culture=="HG"], 1, sum),
       col="blue", lwd=2, type="l")  # Add points for 'HG' culture values
plot(warcasualties[1:(i-1)], lwd=2, col="green", type="h", yaxt="n", ylab="")  # Plot war casualties
axis(4)  # Add an axis on the right side
par(new=T)  # Allow a new plot to be drawn on the existing plot
growF <- apply(Nts[1:(i-1), original.sites$culture=="F"], 1, sum)  # Sum of 'F' culture values excluding last row
growHG <- apply(Nts[1:(i-1), original.sites$culture=="HG"], 1, sum)  # Sum of 'HG' culture values excluding last row
plot(growF, col="red", type="l", lwd=2, ylim=c(0, max(growF, growHG)))  # Plot growth of 'F' culture
points(growHG, col="blue", lwd=2, type="l")  # Add points for growth of 'HG' culture
```



### Generate and share final simulation

Now we have the full setup, we want to generate a record people will uses :


1. We first load the maps and the sites:

```{r loadressources}
sites=vect("data_original/sitesinitialposition/")
```

2. run one simulation


```{r fakesimu,eval=F}

ts=1000
expname="basar"
print(paste0("Starting simulation ",expname))

onesimu=run_simulation (
                           sites=sites,
                           viable=viable,
                           dem=height.ras,
                           ressources=rast("data_original/resources.tiff"),
                           water=height.wat,
                           foldervid=expname,
                           visu=F,visumin=TRUE,
                           ts=ts,#length of simulation in year
                           Kbase=c("HG"=35,"F"=110),#difference in K for the two cultures
                           cul_ext=c("HG"=7,"F"=6),#spatial penality to extent: lower, bigger penality
                           penal_cul=c("HG"=4,"F"=5),#penality of occupational area: low, other sites can cam close
                           prob_birth=c("HG"=0.3,"F"=0.5),#proba to give birth every year
                           prob_survive=c("HG"=0.8,"F"=0.65),#proba to die when pop > K
                           prob_split=c("HG"=.5,"F"=.6),#proba to create new settlement when Ne > K
                           minimals=c("HG"=.14,"F"=.20),#how big the group of migrant should be to create a new city vs migrate to a existing one 
                           bufferatack=300,#distance max around which settlement can fight
                           prob_move=c("HG"=0.2,"F"=0.1) #proba to migrate to existing settlement when Ne > K
                           )
```


2. generate a few output, including a video that helps to select the good simulation, these things are also done by `scriptmini.R`:

```{r getfakeres,eval=F}


Nts=onesimu$Nts
warcasualties=onesimu$warcasualties
sites=onesimu$sites
i=min(ts,which(apply(Nts,1,sum)==0))
pdf(paste0(expname,"_mapFinal.pdf"))
plotMap(height.ras,height.wat,paste0("year ",i))
plot(sites,cex=(as.integer(Nts[i,]>0)*0.3+Nts[i,]/200),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
text(sites)
dev.off()


pdf(paste0(expname,"growth_utils.pdf"))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="green",type="h",yaxt="n",ylab="",xlim=c(0,i))
axis(4)
par(new=T)
growF=apply(Nts[1:i,sites$culture=="F"],1,sum)
growHG=apply(Nts[1:i,sites$culture=="HG"],1,sum)
plot(growF,col="red",type="l",lwd=2,ylim=c(0,max(growF,growHG)),xlim=c(0,i))
points(growHG,col="blue",lwd=2,type="l")
dev.off()

pdf(paste0(expname,"growth_tot.pdf"))
plot(warcasualties[1:i-1],lwd=2,col="green",type="h",yaxt="n",ylab="")
axis(4)
par(new=T)
growT=apply(Nts[1:i,],1,sum)
plot(growT,col="black",type="l",lwd=2,ylim=c(0,max(growT)))
dev.off()
```

Let's redo that with the selected simulation:


```{r initialcondition}

onesimu=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_all.RDS")
#Nts=readRDS("nts")
#warcasualties=readRDS("war")
#sites=terra::vect(readRDS("general_results_selected_simu/buffattack300_K110_PSU065_3_sitesRast.RDS"))
Nts=onesimu$Nts
warcasualties=onesimu$warcasualties

```

Plot the sites at the beginning 

```{r plotinitialcondition,cache=F}
# source("tools.R")

#expname="general_results_selected_simu/buffattack300_K110_PSU065_3"
#sites=vect(readRDS(paste0(expname,"_sitesRast.RDS")))

sites=vect("data_original/sitesinitialposition/")
height.ras=rast("data_original/east_narnia4x.tif")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA

i=1
plotMap(height.ras,height.wat,paste0("year ",i,", when it all begins"))
plot(sites,cex=(as.integer(Nts[1,]>0)*0.5+Nts[1,]/100),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture)])
text(sites[Nts[1,]>0],pos=3)

```

We can look at the war and the number of deaths for each battle

```{r plotwar}
i=1000
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",ylab="# dead",xlim=c(0,i))
```

Comparison to the total population growth of each culture

```{r plotwarWRTgrowths}
i=1000
par(mar=c(5,5,1,5))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",yaxt="n",ylab="",xlim=c(0,i),xlab="time")
axis(4)
mtext("# dead",4,2.5)
par(new=T)
growF=apply(Nts[1:i,sites$culture=="F"],1,sum)
growHG=apply(Nts[1:i,sites$culture=="HG"],1,sum)
plot(growF,col="orange",type="l",lwd=5,ylim=c(0,max(growF,growHG)),xlim=c(0,i),ylab="pop",xlab="")
points(growHG,col="blue",lwd=5,type="l")
```

If we look at the total population size, we can see how the clashes slowed down the population growth:
```{r plotwarTotgrowths}
par(mar=c(5,5,1,5))
plot(2:i,warcasualties[1:(i-1)],lwd=2,col="dark red",type="h",yaxt="n",ylab="",xlim=c(0,i),xlab="time")
axis(4)
mtext("# dead",4,2.5)
par(new=T)
growT=apply(Nts,1,sum)
plot(growT,col="black",type="l",lwd=5,ylim=c(0,max(growT)),xlim=c(0,i),ylab="pop",xlab="")

```

This show if we looka at the growth rate:

```{r rategrowth}
plot(diff(growT,lag=100),type="l")
```


Replaying all that will give this:


```{r finalvideo,echo=F,results="asis"}
cat('<video controls> <source  src="interfinal.mp4" type="video/mp4"></video>')
```


```{r getRaster,eval=T,cache=F}
height.ras=rast("data_toshare/dem_raster.tiff")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA
height.groups=height.ras
maxh=max(height.ras[],na.rm=T)
height.groups[height.groups<mean(height.groups[])]=NA
height.groups[height.groups<(maxh*.7)]=1
height.groups[height.groups>(maxh*.7)]=200
height.groups[is.na(height.groups)]=-1
height.poly=as.polygons(height.groups)
viable=makeValid(height.poly[2,])
expname="data_original/general_results_selected_simu/buffattack300_K110_PSU065_3"
onesimu=readRDS(paste0(expname,"_all.RDS"))
sites=readRDS(paste0(expname,"_sitesRast.RDS"))
if(class(sites)[1] == "PackedSpatVector")
    sites=terra::vect(sites)
Nts=onesimu$Nts
#alldeposit=lapply(1:ncol(Nts),function(s)Rec_c(sapply(Nts[1:i,s],A_rates), InitBP = 7500,ts=ts,r = 0.2, Max_bone_thickness = "m"))
#allShortLoss=lapply(1:10,function(ind){Rec=alldeposit[[ind]];print(paste0("settlement #",ind));apply(Rec,2,short_loss,.6)})
#allShortLoss=lapply(allShortLoss,function(Rec)long_loss(Rec,.9997,7500))
```

Using `Rec_c` , `short_loss` and `long_loss` we generate deposit. Generating everything can take long time but  it's basically done like this:


```{r doloss,eval=F}

#All in one:
alldeposit=lapply(1:ncol(Nts),function(s)Rec_c(sapply(Nts[,s],A_rates), InitBP = 7500,ts=ncol(Nts),r = 0.2, Max_bone_thickness = "m"))
allLosses=lapply(1:length(alldeposit),function(ind){
                 st=Sys.time()
                    print(paste0("settlement #",ind));
                    Rec=alldeposit[[ind]];                    
                    Rec=apply(Rec,2,short_loss,.6)  #apply short loss
                    Rec=long_loss(Rec,.9997,7500) #apply long loss
                    print(Sys.time()-st)
                    return(Rec)
})
rm(alldeposit)
```

Remove where to small

```{r removesmall, eval=F}


maxSites=max(sapply(allLosses,sum))
nsample=round(sapply(allLosses,sum)*30/maxSites)
allRemainingDates=lapply(seq_along(allLosses),function(r)extractDates(allLosses[[r]],n=nsample[r]))
rm(allLosses) #object are big, need to free memory

```



```{r getadates,eval=F}

#pdf(paste0(expname,"_mapFound.pdf"))
#plotMap(height.ras,height.wat,paste0("final after losses"))
#plot(sites[1:10,],cex=3*(lengths(allRemainingDates)-1)/(29),pch=21,add=T,bg=rainbow(2,alpha=.6)[as.factor(sites$culture[1:10])])
#dev.off()

dates=unique(unlist(allRemainingDates))
dates=rev(sort(dates[!is.na(dates)]))
plot(table(unlist(allRemainingDates)))
totallDatesRemains=sapply(allRemainingDates,function(i)table(factor(i,levels=dates)))
saveRDS(allRemainingDates,file=paste0(expname,"_dates.RDS"))

```

From this we have, for every site, a list of date(s)

This has been done using the script `exploreDate.R` but will be explained here:


Loading simulation results, green are the sites who survived, dots are all the site that existed.

```{r getfakeresults,cache=F}
height.ras=rast("data_original/east_narnia4x.tif")
height.wat=height.ras
height.wat[height.wat>mean(height.wat[])]=NA


Nts=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_all.RDS")$Nts
ress=rast("data_original/resources.tiff")
allsites=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_sitesRast.RDS")
if(class(allsites)[1] == "PackedSpatVector")
    allsites=terra::vect(allsites)
plotMap(height.ras,height.wat)
plot(allsites,add=T)
plot(allsites[Nts[nrow(Nts),]>0,],add=T,pch=21,bg="green",cex=Nts[nrow(Nts),]/60)
```

Loading dates:

```{r getfakedates,cache=F}
dates=readRDS("data_original/general_results_selected_simu/buffattack300_K110_PSU065_3_dates.RDS")
```

Keep sites with foundable dates

```{r extractdates,cache=F}

foundsites=allsites[lengths(dates)>0,]
foundsites$numdates=unlist(lengths(dates[lengths(dates)>0]))
founddates=dates[lengths(dates)>0]
```

Remove BP to the dates, add some random variation

```{r addstdev,cache=F}
stdpool=c(20,30,40,50,60,80,100,120)
founddates=lapply(founddates,sort)
founddates=lapply(founddates,gsub,pattern=" BP",replacement="")
founddates=lapply(founddates,rev)
founddates=lapply(founddates,function(i)paste0(i," ± ",sample(stdpool,length(i),replace=T,prob=c(3,3,3,3,2,2,1,1))," BP"))
foundsites$dates=sapply(founddates,paste0,collapse=" | ")
plotMap(height.ras,height.wat)
plot(foundsites,cex=foundsites$numdates/20+1,pch=21,bg=as.factor(foundsites$culture),add=T)
```

### Split environment in a grid

Split the environment in grids and extract date for each grid

```{r plotallfoundiste,cache=F}

leftdates=dates[lengths(dates)>0]
plotMap(height.ras,height.wat)
plot(foundsites,cex=foundsites$numdates/20+.1,pch=21,bg=1,add=T)
squares=st_make_grid(height.ras,.5)
plot(squares,add=T,col=adjustcolor(rainbow(length(squares)),.35))
text(st_coordinates(st_centroid(squares)),label=1:length(squares),col="white")
text(st_coordinates(st_centroid(squares)),label=1:length(squares),col="white")
```

Public selection

```{r getpublicsele,cache=F}
plotMap(height.ras,height.wat)
plot(foundsites,cex=foundsites$numdates/20+1,pch=21,bg=as.factor(foundsites$culture),add=T)
selection=c(14,30,45,65,66)
plot(squares[selection],add=T,col=adjustcolor("blue",.3))
inter=st_intersection(st_as_sf(foundsites),squares[selection])
plotMap(height.ras,height.wat)
plot(st_geometry(inter),add=T,bg=rainbow(2,alpha=.6)[as.factor(inter$culture)],pch=21,cex=1+inter$numdates/10)
```

### Giving name to the publicly available data {#naming}

To name publicly available sites we used: [fantasynamegenerators.com](https://www.fantasynamegenerators.com/fantasy-town-names.php). We manually positionned them with cultural relevent group, although this had no impact on the results. 

```{r namesites,cache=F}

site_dist=st_distance(inter)
#min(site_dist[as.numeric(site_dist)>units(0)])

sitesnames=1:nrow(inter)

sitesnames[c(1:4, 20)]=c("Farwallow" ,"Bearcall" ,"Dustscar" ,"Clearreach" ,"Rabbithole")

fr=c("Épibéliard" ,"Cololuçon" ,"Pulogne" ,"Haguemasse" ,"Auriteaux" ,"Bourville" ,"Banau" ,"Montnesse" ,"Bannet" ,"Alenlon", "Roullac" ,"Genneville" ,"Vinlès" ,"Antonnet" ,"Courtou" ,"Beaulogne" ,"Coloville" ,"Sarsart" ,"Soilon" ,"Cololimar")
sitesnames[5:19]=fr[1:(19-4)]
spain=c("Zava" ,"Catadrid" ,"Tegon" ,"Alicia" ,"Mulid" ,"Zararbella" ,"Malid" ,"Cásca" ,"Granalejos" ,"Segorez" ,"Terteixo" ,"Astumanca" ,"Galle" ,"Talona" ,"Girovega" ,"Albanada" ,"Nadoba" ,"Senca" ,"Vallanca" ,"Taville")

sitesnames[21:length(sitesnames)]=spain[1:(length(sitesnames)-20)]
inter=cbind(inter,sitesnames)
plotMap(height.ras,height.wat)
plot(st_geometry(inter),add=T,bg=rainbow(2,alpha=.6)[as.factor(inter$culture)],pch=21,cex=1+inter$numdates/10)
text(st_coordinates(inter),inter$sitesnames,cex=.8,pos=3)
```

Export csv and data for each public square

```{r exportpublic,eval=F}


inter$sitesnames=sitesnames
for(g in selection){
curr=inter[inter$ID==g,]
coords=st_coordinates(curr)
    write.csv(file=paste0("square_",g,".csv"),cbind.data.frame(sitename=curr$sitesnames,lon=coords[,1],lat=coords[,2],dates=curr$dates,economy=curr$culture))
}

```

Export csv and data for all ever square

```{r exportprivate,eval=F}

allsites=st_intersection(st_as_sf(foundsites),squares[-selection])

for(g in (1:length(squares))[-selection]){
curr=allsites[allsites$ID==g,]
coords=st_coordinates(curr)
    write.csv(file=paste0("square_",g,".csv"),cbind.data.frame(lon=coords[,1],lat=coords[,2],dates=curr$dates,economy=curr$culture))
}

ld=strsplit(inter$dates," \\| ")
ld=lapply(ld,function(i)gsub(" ± .*","",i))
inter$start=sapply(ld,max)
inter$end=sapply(ld,min)

```


Some maps we use and did for fake papers and social media etc... 


```{r randommaps,out.width="30%",fig.show="hold"}

## plotting oldschool map

par(mar=c(0,0,0,0),oma=c(0,0,0,0))

plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=.8,col=as.factor(inter$culture),alpha=.8)

old=inter[inter$start>7180,]
ne=inter[inter$start>6800 & inter$end<6900,]

plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(old),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(old),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(old$culture),alpha=.8)


plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)


plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)

plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.1,labels="",bg="light blue",add=T)
plot(st_geometry(ne),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(ne),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(ne$culture),alpha=.8)
```

More maps for twitter:

```{r randommaps2,out.width="50%",fig.show="hold"}
plot(height.wat,col=adjustcolor("light blue",.4),reset=F,legend=F,axes=F)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.2,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
sel=inter[c(20,2,15,36),]
text(st_coordinates(sel),sel$sitesnames,cex=.8,pos=c(1,2,2,4))
text(x = -1, y = .7, labels = "❤️", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -1, y = 1.2, labels = "🔥 ", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -1, y = 1.2, labels = "⚔️ ", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -1.2, y = 1.8/2, labels = "???? ", adj=c(0,0), cex = 3, col = "black", font = 1)
```

More maps!! (the coolest)


```{r randommaps3,out.width="50%",fig.show="hold"}
plotMap(height.ras,height.wat)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.2,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
coastline=st_cast(st_as_sf(as.polygons(height.ras>mean(height.ras[])))[2,],"MULTILINESTRING")
plot(coastline,add=T,lwd=1.1,col="black")
sel=inter[c(20,2,15,36),]
text(st_coordinates(sel),sel$sitesnames,cex=.8,pos=c(1,2,2,4))

```


```{r randommaps4,out.width="50%",fig.show="hold"}

plotMap(height.ras,height.wat)
contour(height.ras,levels=seq(5,300,15),axes=F,ann=F,lwd=.2,labels="",bg="light blue",add=T)
plot(st_geometry(inter),pch=4,add=T,lwd=.5,cex=3)
plot(st_geometry(inter),pch=20,add=T,lwd=.2,cex=1.8,col=as.factor(inter$culture),alpha=.8)
coastline=st_cast(st_as_sf(as.polygons(height.ras>mean(height.ras[])))[2,],"MULTILINESTRING")
plot(coastline,add=T,lwd=1.1,col="black")
sel=inter[c(20,2,15,36),]
text(st_coordinates(sel),sel$sitesnames,cex=.8,pos=c(1,2,2,4))
#text(x = -1, y = .7, labels = "🍻", adj = c(0, 0), cex = 3, col = "black", font = 1)
text(x = -3.4, y = .7, labels = "🍻 ", adj = c(0, 0), cex = 10, col = "black", font = 1)
text(x = -1.8, y = .7, labels = "👍 ", adj = c(0, 0), cex = 8, col = "black", font = 1)
text(x = -.5, y = .7, labels = "🍔 ", adj = c(0, 0), cex = 6, col = "black", font = 1)
#text(x = -1, y = 1.2, labels = "⚔️ ", adj = c(0, 0), cex = 3, col = "black", font = 1)
#text(x = -1.2, y = 1.8/2, labels = "???? ", adj=c(0,0), cex = 3, col = "black", font = 1)
```

## The challenge

### Research Questions

RQ1. The Poppy-chewers and Rabbit-skinners had a hostile relationship.

RQ2. Poppy-chewers followed an exponential population growth, while Rabbit-Skinners eventually stalled, and as the Poppy-chewers population size started to increase, their population declined (Fig. 3).

RQ3. The rate of dispersal varied across the landscape, with an average of ~0.62 Km/year, faster for the sea crossing (~1.11 Km/year) and a slight slow-down to ~0.57 Km/year for the Northeastern quadrant of the map.

### Challenge & EAA workshop


## Proposals

### P1 

P1 (https://github.com/dpriss/Archaeoriddle_Kahlenberg_Priss) was based on agent-based modelling (ABM) combined with exploratory data analysis. It first studied the land of Rabbithole and calibrated the dates provided, which were then used to compute trajectories of dispersal and study site preference using ArcGIS pro and R. The suggested dispersal rates over land and water were close to the Archaeoriddle solution. The proposal then used the results of its analyses to fit the ABM, which was built using NetLogo. For the ABM, moving groups of hunter–gatherers and farmers, as well as the different settlements, were treated as agents, with starting values obtained from the exploratory data analysis and relevant literature. Additionally, it introduced behavioural rules, including movement and site preference, reaction to population threshold or reaction to interaction. After running the model several times with different parameters, it correctly predicted a hostile relation between the two groups and inferred an initial location of the Rabbit-skinners in the northeastern quadrant and the Poppy-chewers in the southwestern quadrant, while also detecting the latters’ northward movement. The expansion rate was not captured by the initial assumption of logistic growth in the areas already densely inhabited by Poppy-chewers, which makes sense considering the different population trajectories of the two groups. 


### P2 

P2 (https://doi.org/10.5281/zenodo.12803445) used point-process modelling focused initially on the sampling process to build first-order models combining fitness and the available cells to predict potential occupation. It then focused on the question of whether there was conflict between the two groups. After building different archaeological phases, it computed the clustering patterns of the groups under the assumption that higher clustering could lead to higher conflict (Field, 2004). Following this, it computed the interaction distance between groups of settlements of hunter–gatherers and farmers through a multitype Strauss model. According to the results, hostilities increased over time. In the original model, the rules for hostility were not time dependent, but we have seen that, even if the rules for conflict did not change, as the population grew and the number of settlements increased, there was more probability of contact and thus more probability of conflict, which finally resulted in an increased mortality.

### P3 

P3 (https://doi.org/10.5281/zenodo.8260754) used species-distribution modelling in R to develop a four-stage research design. The first stage focused on determining which additional cells would result in a representative sample of the range variables (elevation and resource quality) present within the data. The second stage focused on data exploration to identify internal temporal, spatial and farmer and forager patterns. In the third stage, it generated summed probability distributions from the calibrated radiocarbon dates to generate relative estimates of population size through time for foragers and farmers. The fourth stage combined these data into a spatiotemporal species-distribution model, where both time and space were explicit predictors used to estimate the distribution of farmers and foragers in 100-year intervals. The spatiotemporal species distribution was successful in reproducing the directionality of the farming dispersal (from south to north) as well as the decline in hunter–gatherer populations.


### P4

P4 (https://github.com/AlexesMes/Archeaoriddle_RabbitWorld) first developed a strategy to aid in the selection of additional data using a friction calculation that considered the distance from a putative origin region, the elevation of the region and its environmental suitability. Using R, analysis focused on capturing local complexity in the dispersal of Poppy-chewers in the study area. To track this, it used a hierarchical Bayesian phase model that was informed by all the selected settlements, both the ones held in common with other proposals and the ones obtained after calculations for additional sampling. This method allowed uncertainty to be introduced. It divided Rabbithole into 25 subareas and calculated the times of arrival of the Poppy-chewers for each area, including their high-probability density intervals, resulting in a successful approach to expansion rates. 


### P5 

P5 used a qualitative approach (no code involved) that considered two cultures, A and B, initially established in regions R_A and R_B separated by a sea. Culture A eventually gained the ability to cross or bypass the barrier and to enter the region R_B and to make contact with B. Under the condition that all other parameters were equal (e.g. environmental fitness in both regions and no new pathogens or predators unknown to A), a significantly shorter settlement persistence of culture A in R_B, when compared with A’s settlement persistence assessed in other regions implies a hostile relation between A and B.
