--- 
title: "Methods Riddles, the ultimate revelations"
author: "CDAL"
date:  "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook: 
    config:
      toc:
        after: |
          <li><a href="https://github.com/acortell3/archaeoriddle/">
          source</a></li>

    nature:
      beforeInit: "https://platform.twitter.com/widgets.js"
documentclass: book
biblio-style: apalike
link-citations: yes
github-repo: acortell3/archaeoriddle
description: "First draft of Rbookdown to describe how data have been generated"
favicon: "favicon.ico"
---
# Preface{-}

```{r setup, include=FALSE}
library(magrittr)

knitr::opts_chunk$set(echo = TRUE,cache=FALSE,collapse=TRUE)
```

```{r write-bib, include=FALSE}
knitr::write_bib(c(.packages(), 'bookdown','tidyverse'), 'packages.bib')
```

This is a very first draft to setup a github hook that compile a bookdown in a private space

To compile the book:

```{r,eval=F}
bookdown::render_book(".")
```

For this to be successfull you will need a few aditional packages to install:

```{r packages, message=FALSE}
needed=c("DiagrammeR","viridis","terra","sf","igraph","rayshader","printr") 
## printr for bookdown compilation
success=sapply(needed,require,character.only=T)
sapply(needed[!success],install.packages)
success=sapply(needed,require,character.only=T)
```

These package will very likely need other package that will depend on your operating system, so check the error message from the previous commands!

If you want to use the function and methods described in the bookdown you will need to install the archaeoriddle package. With `devtools` installed, you can do:

```{r,eval=F}
devtools::install_github("acortell3/archaeoriddle")
```

Once this is done:

```{r}
library(archaeoriddle)
```


The book can also be found already compiled [here](https://cdal.arch.cam.ac.uk/archaeoriddle/).

